---
title: "DataVizForBTS"
format: html
editor: visual
---

```{r setup, include=FALSE}

if (!require("pacman")) install.packages("pacman")
pacman::p_load(brms, ggplot2, patchwork, plotly, tidybayes, modelr, viridis, tidyverse, dplyr, truncnorm, lme4, faux, scam, bayestestR)

# Set the current working directory to the one where this file is
current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path)
setwd(current_working_dir)

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Data & Wrangling

```{r}

# load in data 2025/09/11
xdata <- read_csv("../data/merged_250911.csv")

# derive some relevant measurements
xdata <- xdata |> 
  # filter erroneous
  mutate(response_direction_corrected = case_when(
                                                  final_rating == 11 ~ "m",
                                                  final_rating > 11 ~ "h",
                                                  final_rating < 11 ~ "l"),
         is.correct = case_when(
           response_direction_corrected == "l" & stimuli_direction == "n" ~ 1,
           response_direction_corrected == "h" & stimuli_direction == "p" ~ 1,
           .default = 0)) |> 
  # include trials when slider was not moved?
  #filter(stimuli_direction != 0) |>
  filter(delta_t != 0) |> 
  mutate(
         # make delta_f positive / negative
         delta_f_raw = ifelse(stimuli_direction == "p", delta_f, delta_f * -1),
         # make delta_f_raw into STs (reference 150Hz)
         delta_f_raw_st = 12 * log2((150-delta_f_raw)/150),
         # make response_f relative to baseline 150hz      
         response_f_rel = response_f - 150,
         # make response_f into STs (reference 150Hz)
         response_f_rel_st = 12 * log2((150-response_f_rel)/150),
         # make response_f absolute without sign differences
         response_f_rel_abs = abs(response_f_rel),
         # relative to target
         response_f_prop = response_f_rel / delta_f,
         # calculate actual f0 of presented tonal onset
         presented_f = 150 + delta_f_raw,
         presented_f_st = 12 * log2(presented_f/150),
         # make is.correct as.numeric
         correct = as.numeric(is.correct),
         # make part ID a char
         participant = as.character(participant)
         )

```

# Continuous dependent variable

## Exploration of landscape

We explore how pitch perception changes as a function of `delta_t` and `delta_f.`

```{r}

xdata_agg <- xdata |> 
  group_by(delta_f_raw, delta_t, participant) |> 
  summarise(response_f_rel = mean(response_f_rel, na.rm = TRUE))
  
xdata_agg_all <- xdata |> 
  group_by(delta_f_raw, delta_t) |> 
  summarise(response_f_rel = mean(response_f_rel, na.rm = TRUE))
  
descriptive_plot_03 <- 
ggplot(data = xdata_agg, 
       aes(x = delta_t,
           y = response_f_rel,
           color = delta_f_raw)) +
  # geom_line(data = xdata_agg, 
  #           aes(group = interaction(participant, delta_f_raw_st)),
  #         alpha = 0.2) +
  geom_point(data = xdata_agg_all) +
  geom_line(data = xdata_agg_all,
             aes(group = delta_f_raw),
             lwd = 2.5,
            color = "black") +
  geom_line(data = xdata_agg_all,
             aes(group = delta_f_raw),
             lwd = 2) +
  scale_y_continuous(limits = c(-20,20),
                     breaks = seq(-20,20,10)) +
  scale_x_continuous(breaks = c(unique(xdata$delta_t))) +
  scale_color_viridis_c(option= "viridis") +
  #scale_color_stepsn(colours = "viridis") +
  labs(x = "\nslope duration in ms",
       y = "average observed response in Hz\n",
       color = "pitch difference") +
  theme_minimal()
  
descriptive_plot_03

```

Slight non-linear relationship with delta_t. Let's look at it in 3-dimensional space:

```{r}

# set up matrix
z_matrix <- matrix(xdata_agg_all$response_f_rel, nrow = 8, ncol = 6, byrow = TRUE)

x_axis <- unique(xdata_agg_all$delta_t)
y_axis <- unique(xdata_agg_all$delta_f_raw)

plot_ly(x = ~x_axis, y = ~y_axis, z = ~z_matrix, 
        type = "surface",
        colors = 'RdPu') |> 
  layout( scene = list(
    xaxis = list(title = 'Onset Duration (ms)', tickvals = x_axis, ticktext = x_axis),
    yaxis = list(title = 'Onset Frequency (Hz)', tickvals = list(-5,-4,-3,-2,-1,0,1,2,3,4,5), ticktext = list(-5,-4,-3,-2,-1,0,1,2,3,4,5)),
    zaxis = list(title = 'Response (Hz)')
  ))

```

## Statistical model

Relevant parameters are estimated through the following model (following Marton's parametric suggestion) \# need justification for our model choice

`response_f_rel_st \~ a + b \* (( (delta_t - x0) / (x_max - x0) )\^inv_logit(c))`

a = assesses intercept, i.e. perceived pitch differences at `delta_t = 0` and `delta_f_raw_st = 0`

b = assesses linear relationship of `delta_t` and `response_f_rel_st` and how this relationship is modulated by `delta_f_raw_st`

c = delta_f_raw_st \# assess non-linear relationship of `delta_t` and `response_f_rel_st` and how this relationship is modulated by `delta_f_raw_st`

x0 \# ?

```{r}

# add ceiling
xdata <- xdata |> 
  mutate(
  #x_min=min(delta_t),
  x_max = max(delta_t)
)

# TO DO: Explore other functions?
## gumbel function?
## Naka-rushton?

# Model of the following basic structure:
# y ~ a + b*x^

# TO DO: Justify priors
# TO DO: random slopes for all these parameters? Computationally feasible?

# run model
parametric_model <- brm(
  bf(response_f_rel ~ a + b * (( (delta_t - x0) / (x_max - x0) )^inv_logit(c)),
     a ~ 1,
     b ~ delta_f_raw,
     c ~ delta_f_raw,
     x0 ~ 1,
     nl = T),
  data = xdata,
  file = "../models/model_nl_TR",
  prior = prior(normal(0,0.5), nlpar = "a")+
          prior(normal(0,0.5), nlpar = "b")+
          prior(normal(0,5), nlpar = "x0")+
          prior(normal(0,1), nlpar = "c", lb = 0, ub = 1),
  control = list(adapt_delta=0.99)
  )

# comment TR: takes 90 seconds on my machine, so computationally expensive
parametric_model
# model predicts mostly linear relationship

# posterior predictive check
pp_check(parametric_model)
# 1. not truly continuous measure, problem?
# 2. regardless, distribution looks spikier than gaussian

# check chains
plot(parametric_model, pars = "^b_") # okay

xdata$preds <- posterior_summary(posterior_epred(parametric_model))[,1]

# Visualize
model_plot_01 <- 
ggplot(data = xdata, 
       aes(x = delta_t,
           y = preds,
           color = delta_f_raw,
           group = delta_f_raw)) +
  geom_line(lwd = 2.5, color = "black") +
  geom_line(lwd = 2) +
   scale_y_continuous(limits = c(-20,20),
                     breaks = seq(-20,20,10)) +
  scale_x_continuous(breaks = c(unique(xdata$delta_t))) +
  scale_color_viridis_c(option= "viridis") +
  #scale_color_stepsn(colours = "viridis") +
  labs(x = "\nslope duration in ms",
       y = "predicted observed response in Hz\n",
       color = "pitch difference") +
  theme_minimal()
  
model_plot_01

data_v_model <- descriptive_plot_03 / model_plot_01
data_v_model

# pretty decent fit

```

In 3D:

```{r}

xdata_agg_pred <- xdata |> 
  group_by(delta_t, delta_f_raw) |> 
  summarise(preds = mean(preds))

# set up matrix
z_matrix <- matrix(xdata_agg_pred$preds, nrow = 6, ncol = 8, byrow = TRUE)

x_axis <- unique(xdata_agg_pred$delta_f_raw_st)
y_axis <- unique(xdata_agg_pred$delta_t)

plot_ly(x = ~x_axis, y = ~y_axis, z = ~z_matrix, 
        type = "surface",
        colors = 'RdPu') |> 
  layout( scene = list(
    yaxis = list(title = 'Onset Duration (ms)', tickvals = y_axis, ticktext = y_axis),
    xaxis = list(title = 'Onset Frequency (Hz)', tickvals = list(-5,-4,-3,-2,-1,0,1,2,3,4,5), ticktext = list(-5,-4,-3,-2,-1,0,1,2,3,4,5)),
    zaxis = list(title = 'Response (Hz)')
  ))

```
