---
title: "ManyTones analysis"
author: "Timo Roettger"
date: "2025-07-21"
output: html_document
---

```{r setup, include=FALSE}

if (!require("pacman")) install.packages("pacman")
pacman::p_load(brms, ggplot2, patchwork, plotly, tidybayes, modelr, viridis, tidyverse, dplyr, truncnorm, lme4, faux, scam, bayestestR)

# Set the current working directory to the one where this file is
current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path)
setwd(current_working_dir)

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Data & Wrangling

```{r}

# load in data 2025/09/11
xdata <- read_csv("../data/merged_250911.csv")

# derive some relevant measurements
xdata <- xdata |> 
  # filter erroneous
  mutate(response_direction_corrected = case_when(
                                                  final_rating == 11 ~ "m",
                                                  final_rating > 11 ~ "h",
                                                  final_rating < 11 ~ "l"),
         is.correct = case_when(
           response_direction_corrected == "l" & stimuli_direction == "n" ~ 1,
           response_direction_corrected == "h" & stimuli_direction == "p" ~ 1,
           .default = 0)) |> 
  # include trials when slider was not moved?
  #filter(stimuli_direction != 0) |>
  filter(delta_t != 0) |> 
  mutate(
         # make delta_f positive / negative
         delta_f_raw = ifelse(stimuli_direction == "p", delta_f, delta_f * -1),
         # make delta_f_raw into STs (reference 150Hz)
         delta_f_raw_st = 12 * log2((150-delta_f_raw)/150),
         # make response_f relative to baseline 150hz      
         response_f_rel = response_f - 150,
         # make response_f into STs (reference 150Hz)
         response_f_rel_st = 12 * log2((150-response_f_rel)/150),
         # make response_f absolute without sign differences
         response_f_rel_abs = abs(response_f_rel),
         # relative to target
         response_f_prop = response_f_rel / delta_f,
         # calculate actual f0 of presented tonal onset
         presented_f = 150 + delta_f_raw,
         presented_f_st = 12 * log2(presented_f/150),
         # make is.correct as.numeric
         correct = as.numeric(is.correct),
         # make part ID a char
         participant = as.character(participant)
         )

```

# Accuracy of pitch direction

The data will be complex and allow to quantitatively explore many different relationships. For the confirmatory part (and in turn sample size justification), I propose to stick to focusing on the central relationship between pitch perception and the two acoustic dimensions that are manipulated: Duration of pitch change between onset and level tone (`delta_t`) and the extend of pitch change between onset and level tone (`delta_f`). To simplify, we test the hypotheses that `delta_t` and `delta_f` affect the accuracy of perceiving the correct pitch direction (`correct`).

## Descriptively

A first descriptive look at those relationships:

```{r}

# aggregate over delta_f, delta_t and accuracy
xdata_agg_all <- xdata |> 
  group_by(delta_f, delta_t) |> 
  summarise(correct = mean(correct, na.rm = TRUE))
  
# Plot empirical accuracy
descriptive_plot_01 <- 
  ggplot(data = xdata_agg_all, 
       aes(x = delta_t,
           y = correct,
           color = as.factor(delta_f))) +
  geom_hline(yintercept = 0.5,
             lty = "dashed") +
  geom_path(lwd = 1.5, color = "black") +
  geom_path(lwd = 1.5) +
  geom_point(size = 3, color = "black") + 
  geom_point(size = 2.5) + 
  #scale_color_viridis(option="viridis") +
  scale_y_continuous(limits = c(0,1),
                     breaks = seq(0,1,.25)) +
  scale_x_continuous(breaks = seq(0,250,50)) +
  facet_wrap(delta_f ~ ., ncol = 2) +
  scale_color_viridis_d(option="viridis") +
  labs(x = "\nslope duration in ms",
       y = "observed accuracy of pitch direction\n",
       color = "delta f0\n") +
  theme_minimal()
  
descriptive_plot_01

```

As dependent on stimulus_type:

```{r}

# aggregate over delta_f, delta_t, stimuli_type and accuracy
xdata_agg_all <- xdata |> 
  group_by(delta_f, delta_t) |> 
  summarise(correct = mean(correct, na.rm = TRUE))
  
# Plot empirical accuracy
descriptive_plot_02 <- 
  ggplot(data = xdata_agg_all, 
       aes(x = delta_t,
           y = correct,
           color = as.factor(delta_f))) +
  geom_hline(yintercept = 0.5,
             lty = "dashed") +
  geom_path(lwd = 1.5, color = "black") +
  geom_path(lwd = 1.5) +
  geom_point(size = 3, color = "black") + 
  geom_point(size = 2.5) + 
  #scale_color_viridis(option="viridis") +
  scale_y_continuous(limits = c(0,1),
                     breaks = seq(0,1,.25)) +
  scale_x_continuous(breaks = seq(0,250,50)) +
  facet_grid(delta_f ~ stimuli_type) +
  scale_color_viridis_d(option="viridis") +
  labs(x = "\nslope duration in ms",
       y = "observed accuracy of pitch direction\n",
       color = "delta f0\n") +
  theme_minimal()
  
descriptive_plot_02

```

## Modelling relationship

Hypotheses are tested, and relevant parameters are estimated through the following model:

`correct ~ delta_t * delta_f + (delta_t_c * delta_f_c | participant))`

Predictors are contrast-coded.

### Statistical predictions:

(1) There is evidence that the coefficient `delta_t` is larger than 0.

(2) There is evidence that the coefficient `delta_f` is larger than 0.

```{r}

# scale and contrast code predictors
xdata <- xdata |> 
  mutate(delta_f_c = delta_f - mean(delta_f),
         delta_t_c = delta_t - mean(delta_t),
         # linguistic vs. non-linguistic
         stimuli_type_c = ifelse(stimuli_type == "tone", -0.5, 0.5))

# formula
formula = bf(correct ~ delta_t_c * delta_f_c + (delta_t_c * delta_f_c | participant))

# run model
logistic_model <- brm(
  formula,
  data = xdata,
  file = "../models/model_glm_TR",
  family = "bernoulli",
  # given that the predictors units are rather small, we pick relative narrow prior widths
  prior = c(prior(normal(0,0.5), class = "Intercept"),
            prior(normal(0,0.1), class = "b"),
            prior(lkj(1), class = "cor"),
            prior(normal(0,1), class = "sd")), 
  control = list(adapt_delta=0.99,
                 max_treedepth = 12)
  )


# comment TR: with interacting predictors in random slopes, computationally rather expensive (~15 mins)
print(logistic_model, digits = 5)

plot(logistic_model, variables = "^b_") # okay

# extract posteriors
posteriors <- logistic_model |> 
  spread_draws(c(b_Intercept, b_delta_t_c, b_delta_f_c, `b_delta_t_c:delta_f_c`)) |> 
  ## add sequences of delta_f and delta_t for prediction
  mutate(delta_t = list(seq(min(xdata$delta_t_c), max(xdata$delta_t_c), 2)),
         delta_f = list(seq(min(xdata$delta_f_c), max(xdata$delta_f_c), 10))) |> 
  unnest(delta_t) |> 
  unnest(delta_f) |> 
  # calculate combinations
  mutate(preds = plogis(b_Intercept + b_delta_t_c*delta_t + b_delta_f_c*delta_f + 
                          `b_delta_t_c:delta_f_c`*delta_t*delta_f)) |> 
    pivot_longer(cols = starts_with("preds"),
                 names_to = "factors") |> 
    select(-(1:6)) |> 
    separate(col = factors, into = c(NA, "stimuli_type"),  sep = "_") |> 
    group_by(delta_t, delta_f) |> 
    summarise(pred_m = mean(value, na.rm = TRUE),
              pred_0025 = quantile(value, prob = 0.025),
              pred_0975 = quantile(value, prob = 0.975)) |> 
  mutate(delta_t = delta_t + mean(xdata$delta_t),
         delta_f = delta_f + mean(xdata$delta_f),
         above_chance = ifelse(pred_0025 > 0.5, "above chance", "below chance"))
  
# plot
#model_plot_02 <- 
  ggplot(data = posteriors, 
       aes(x = as.numeric(delta_t),
           y = pred_m,
           #group = as.numeric(delta_t),
           fill = as.factor(delta_f))) +
  geom_hline(yintercept = 0.5,
             lty = "dashed") +
  geom_ribbon(aes(ymin = pred_0025,
                  ymax = pred_0975),
              color = NA,
              alpha = 0.5) +
  geom_line(#aes(colour = above_chance),
            colour = "black",
            lwd = 2) +
  geom_point(data = xdata_agg_all,
             aes(y = correct),
             size = 3) + 
  geom_point(data = xdata_agg_all,
             aes(y = correct, 
                 fill = as.factor(delta_f)),
             pch = 21,
             size = 3) + 
  scale_color_manual(values = c("black", "darkred")) +
  scale_fill_viridis_d(option="viridis") +
  scale_y_continuous(limits = c(0,1),
                     breaks = seq(0,1,.25)) +
  scale_x_continuous(limits = c(40,250),
                     breaks = c(40,80,120,250)) +
  facet_wrap( ~ delta_f, ncol = 2) +
  labs(x = "\nslope duration in ms",
       y = "posterior predictions for\nperceptual accuracy of pitch direction\n",
       fill = "delta f0\n") +
  guides(color = "none") +
  theme_minimal()

```

### Florian's Lapse model

```{r}

# Florians suggestion:
 # estimate guess
formula_FJ = bf(correct ~ guess + (1 -  guess - inv_logit(lambda)) * inv_logit(eta),
             # set guess rate to 1/3
             guess ~ 0 + offset(1/3),
             lambda ~ 1,
             eta ~ delta_t_c + delta_f_c + (delta_t_c + delta_f_c | participant),
             family = bernoulli(link="identity"),
             nl = TRUE)

logistic_model_lapse <- brm(
  formula_FJ,
  data = xdata,
  #family = bernoulli(link='identity'),
  # given that the predictors units are rather small, we pick relative narrow prior widths
  prior = c(
  prior(student_t(3, 0, 2.5), class = "b", nlpar = "eta"),
  prior(lkj(1), class = "cor"),
  prior(student_t(3, 0, 2.5), class = "sd", nlpar = "eta"),
  # Setting intercept prior to weakly regularizing around 5% lapses)
  prior(student_t(3, -2.944439, 2.5), class = "b", coef = "Intercept", nlpar = "lambda")),
  control = list(adapt_delta = 0.99,
                 max_treedepth = 10)
  )

newdata <- expand_grid(delta_t_c = seq(min(xdata$delta_t_c), max(xdata$delta_t_c), 2),
                       delta_f_c = seq(min(xdata$delta_f_c), max(xdata$delta_f_c), 10))

logistic_model_lapse |> 
  epred_draws(newdata = newdata,
              re_formula = NA) |> 
    group_by(delta_t_c, delta_f_c) |> 
    summarise(pred_m = mean(.epred, na.rm = TRUE),
              pred_0025 = quantile(.epred, prob = 0.025),
              pred_0975 = quantile(.epred, prob = 0.975)) |> 
  mutate(delta_t = delta_t_c + mean(xdata$delta_t),
         delta_f = delta_f_c + mean(xdata$delta_f)) |> 
  ggplot(aes(x = as.numeric(delta_t),
           y = pred_m,
           #group = as.numeric(delta_t),
           fill = as.factor(delta_f))) +
  geom_hline(yintercept = 0.5,
             lty = "dashed") +
  geom_ribbon(aes(ymin = pred_0025,
                  ymax = pred_0975),
              color = NA,
              alpha = 0.5) +
  geom_line(#aes(colour = above_chance),
            colour = "black",
            lwd = 2) +
  geom_point(data = xdata_agg_all,
             aes(y = correct),
             size = 3) + 
  geom_point(data = xdata_agg_all,
             aes(y = correct, 
                 fill = as.factor(delta_f)),
             pch = 21,
             size = 3) + 
  scale_fill_viridis_d(option="viridis") +
  scale_y_continuous(limits = c(0,1),
                     breaks = seq(0,1,.25)) +
  scale_x_continuous(limits = c(40,250),
                     breaks = c(40,80,120,250)) +
  facet_wrap( ~ delta_f, ncol = 2) +
  labs(x = "\nslope duration in ms",
       y = "posterior predictions for\nperceptual accuracy of pitch direction\n",
       fill = "delta f0\n") +
  guides(color = "none") +
  theme_minimal()

```

### Logistic of no direction response

```{r}

# Florian: Any chance that you could run a standard mixed-effects logistic regression but with response = “no direction”? I think that will shed light on the concern that I’m hinting at in my previous email. I predict that they are never below chance in terms of the direction. But they have a certain held-out prob mass for “no direction”, and that prob mass decreases for longer slope and bigger differences. So it’s a mixture model: respond “no direction” unless there is sufficient perceptual evidence of directionality; once there is respond based on perceptual evidence about that directionality.

xdata <- xdata |> 
  mutate(no_direction = ifelse(response_direction_corrected == "m", 1, 0))

xdata_agg_all <- xdata |> 
  group_by(delta_f, delta_t) |> 
  summarise(no_direction = mean(no_direction, na.rm = TRUE))

# formula
formula_no = bf(no_direction ~ delta_t_c * delta_f_c + (delta_t_c * delta_f_c | participant))

# run model
logistic_model_no_direction <- brm(
  formula_no,
  data = xdata,
  file = "../models/model_glm_no_direction_TR",
  family = "bernoulli",
  # given that the predictors units are rather small, we pick relative narrow prior widths
  prior = c(prior(normal(0,0.5), class = "Intercept"),
            prior(normal(0,0.1), class = "b"),
            prior(lkj(1), class = "cor"),
            prior(normal(0,1), class = "sd")), 
  control = list(adapt_delta=0.99,
                 max_treedepth = 12)
  )

newdata <- expand_grid(delta_t_c = seq(min(xdata$delta_t_c), max(xdata$delta_t_c), 2),
                       delta_f_c = seq(min(xdata$delta_f_c), max(xdata$delta_f_c), 10))

logistic_model_no_direction |> 
  epred_draws(newdata = newdata,
              re_formula = NA) |> 
    group_by(delta_t_c, delta_f_c) |> 
    summarise(pred_m = mean(.epred, na.rm = TRUE),
              pred_0025 = quantile(.epred, prob = 0.025),
              pred_0975 = quantile(.epred, prob = 0.975)) |> 
  mutate(delta_t = delta_t_c + mean(xdata$delta_t),
         delta_f = delta_f_c + mean(xdata$delta_f)) |> 
  ggplot(aes(x = as.numeric(delta_t),
           y = pred_m,
           #group = as.numeric(delta_t),
           fill = as.factor(delta_f))) +
  geom_hline(yintercept = 0.5,
             lty = "dashed") +
  geom_ribbon(aes(ymin = pred_0025,
                  ymax = pred_0975),
              color = NA,
              alpha = 0.5) +
  geom_line(#aes(colour = above_chance),
            colour = "black",
            lwd = 2) +
  geom_point(data = xdata_agg_all,
             aes(y = no_direction),
             size = 3) + 
  geom_point(data = xdata_agg_all,
             aes(y = no_direction, 
                 fill = as.factor(delta_f)),
             pch = 21,
             size = 3) + 
  scale_fill_viridis_d(option="viridis") +
  scale_y_continuous(limits = c(0,1),
                     breaks = seq(0,1,.25)) +
  scale_x_continuous(limits = c(40,250),
                     breaks = c(40,80,120,250)) +
  facet_wrap( ~ delta_f, ncol = 2) +
  labs(x = "\nslope duration in ms",
       y = "posterior predictions for\nprobability for no direction\n",
       fill = "delta f0\n") +
  guides(color = "none") +
  theme_minimal()

```

### Multinomial model

```{r}

# fit model to ternary response
xdata_agg_viz <- xdata |> 
  mutate(response_direction_corrected = as.factor(response_direction_corrected)) |> 
  group_by(delta_f, delta_t,response_direction_corrected) |> 
  count() |> 
  group_by(delta_f, delta_t) |> 
  mutate(pred_m = n/sum(n),
         ".category" = response_direction_corrected) 


# formula
formula_cat = bf(response_direction_corrected ~ delta_t_c + delta_f_c + (delta_t_c + delta_f_c | participant), family = "categorical")

# run model (takes 20mins)
multinomial_model <- brm(
  formula_cat,
  data = xdata,
  file = "../models/model_glm_cat_TR",
  # given that the predictors units are rather small, we pick relative narrow prior widths
  prior = c(prior(normal(0,0.3), class = "b", dpar = "mul"),
            prior(normal(0,0.3), class = "b", dpar = "mum"),
            prior(lkj(1), class = "cor")), 
  control = list(adapt_delta=0.9,
                 max_treedepth = 10)
  )

# model convergence?
plot(multinomial_model, variables = "^b_") # okay

newdata <- expand_grid(delta_t_c = seq(min(xdata$delta_t_c), max(xdata$delta_t_c), 2),
                       delta_f_c = seq(min(xdata$delta_f_c), max(xdata$delta_f_c), 10))

multinomial_model |> 
    epred_draws(newdata = newdata,
              re_formula = NA) |> 
    group_by(delta_t_c, delta_f_c, .category) |> 
    summarise(pred_m = mean(.epred, na.rm = TRUE),
              pred_0025 = quantile(.epred, prob = 0.025),
              pred_0975 = quantile(.epred, prob = 0.975)) |> 
  mutate(delta_t = delta_t_c + mean(xdata$delta_t),
         delta_f = delta_f_c + mean(xdata$delta_f)) |> 
  ggplot(aes(x = as.numeric(delta_t),
           y = pred_m,
           #group = as.numeric(delta_t),
           fill = as.factor(delta_f),
           group = .category)) +
  geom_hline(yintercept = 0.5,
             lty = "dashed") +
  geom_ribbon(aes(ymin = pred_0025,
                  ymax = pred_0975),
              color = NA,
              alpha = 0.5) +
  geom_line(aes(),
            colour = "black",
            lwd = 2) +
  geom_point(data = xdata_agg_viz,
             size = 3) + 
  geom_point(data = xdata_agg_viz,
             pch = 21,
             size = 3) + 
  scale_fill_viridis_d(option="viridis") +
  scale_y_continuous(limits = c(0,1),
                     breaks = seq(0,1,.25)) +
  scale_x_continuous(limits = c(40,250),
                     breaks = c(40,80,120,250)) +
  facet_grid(.category ~ delta_f) +
  labs(x = "\nslope duration in ms",
       y = "posterior predictions for\nprobability of different responses\n",
       fill = "delta f0\n") +
  guides(color = "none") +
  theme_minimal() +
  theme(strip.text.y = element_text(angle = 0))

```

### Multinomial model with correct/incorrect and middle

```{r}

xdata <- xdata |> 
  mutate(correct_m = ifelse(response_direction_corrected == "m", "no direction", 
         ifelse(correct == 1, "correct", "incorrect")))


# fit model to ternary response
xdata_agg_viz <- xdata |> 
  mutate(correct_m = as.factor(correct_m)) |> 
  group_by(delta_f, delta_t, correct_m) |> 
  count() |> 
  group_by(delta_f, delta_t) |> 
  rename("delta_f_c" = delta_f,
         "delta_t_c" = delta_t) |> 
  mutate(pred_m = n/sum(n),
         ".category" = correct_m) 


# formula
formula_cat_2 = bf(correct_m ~ delta_t_c * delta_f_c + (delta_t_c + delta_f_c | participant), family = "categorical")

formula_cat_2_mo = bf(correct_m ~ mo(delta_t) + mo(delta_f), family = "categorical")

# run model (takes 20mins)
multinomial_model_2 <- brm(
  formula_cat_2,
  data = xdata,
  #file = "../models/model_glm_cat2_TR",
  # given that the predictors units are rather small, we pick relative narrow prior widths
  prior = c(prior(normal(0,1), class = "b", dpar = "muincorrect"),
            prior(normal(0,1), class = "b", dpar = "munodirection"),
            prior(lkj(1), class = "cor")), 
  control = list(adapt_delta=0.99,
                 max_treedepth = 10)
  )


# run model with mo
multinomial_model_2_mo <- brm(
  formula_cat_2_mo,
  data = xdata,
  file = "../models/model_glm_cat2_mo_TR",
  # given that the predictors units are rather small, we pick relative narrow prior widths
  prior = c(prior(normal(0,1), class = "b", dpar = "muincorrect"),
            prior(normal(0,1), class = "b", dpar = "munodirection")
            #prior(lkj(1), class = "cor"),
            #prior(normal(0,1), class = "sd", dpar = "muincorrect"),
            #prior(normal(0,1), class = "sd", dpar = "munodirection")
            ), 
  control = list(adapt_delta=0.99,
                 max_treedepth = 10)
  )

# model convergence?
plot(multinomial_model_2, variables = "^b_") # okay

newdata <- expand_grid(delta_t_c = seq(min(xdata$delta_t_c), max(xdata$delta_t_c), 2),
                       delta_f_c = seq(min(xdata$delta_f_c), max(xdata$delta_f_c), 10))

multi_pred <- multinomial_model_2 |> 
    epred_draws(newdata = newdata,
              re_formula = NA) |> 
    group_by(delta_t_c, delta_f_c, .category) |> 
    summarise(pred_m = mean(.epred, na.rm = TRUE),
              pred_0025 = quantile(.epred, prob = 0.025),
              pred_0975 = quantile(.epred, prob = 0.975)) |> 
  mutate(delta_t_c = delta_t_c + mean(xdata$delta_t),
         delta_f_c = delta_f_c + mean(xdata$delta_f)) 


multi_pred |>  
  ggplot(aes(x = as.numeric(delta_t_c),
             y = pred_m,
           #group = as.numeric(delta_t),
           fill = as.factor(delta_f_c),
           group = .category)) +
  geom_hline(yintercept = 0.5,
             lty = "dashed") +
  geom_ribbon(aes(ymin = pred_0025,
                  ymax = pred_0975),
              color = NA,
              alpha = 0.5) +
  geom_line(aes(),
            colour = "black",
            lwd = 2) +
  geom_point(data = xdata_agg_viz,
             aes(x = delta_t_c, fill = as.factor(delta_f_c)),
             size = 3) +
  geom_point(data = xdata_agg_viz,
             aes(x = delta_t_c,  fill = as.factor(delta_f_c)),
             pch = 21,
             size = 3) +
  scale_fill_viridis_d(option="viridis") +
  scale_y_continuous(limits = c(0,1),
                     breaks = seq(0,1,.25)) +
  scale_x_continuous(limits = c(40,250),
                     breaks = c(40,80,120,250)) +
  facet_grid(.category ~ delta_f_c) +
  labs(x = "\nslope duration in ms",
       y = "posterior predictions for\nprobability of different responses\n",
       fill = "delta f0\n") +
  guides(color = "none") +
  theme_minimal() +
  theme(strip.text.y = element_text(angle = 0))

```

### Inference

We assess hypotheses (1) and (2) through Bayes Factor using the Savage-Dickey approximation (within the `bayesfactor` package):

```{r}

bayesfactor_pointnull(posterior = logistic_model,
                      null = 0)

# evidence for beta > 0 is overwhelming (BF = >10000)
# moderate evidence against an effect of the interaction btw.
# This inferential procedure would have to be sensitivity tested against different priors, but looks like an overwhelmingly large effect

```

### Questions:

-   What to do with those trials that are neither higher or lower but chosen to be level? Options: (1) categorical GLM with three response options (2) remove level responses, and run logistic regression or (3) consider level responses incorrect, and run logistic regression (currently implemented).
-   Can we test against anything meaningful that is not point-zero in this case?

# Continuous dependent variable

## Exploration of landscape

We explore how pitch perception changes as a function of `delta_t` and `delta_f.`

```{r}

xdata_agg <- xdata |> 
  group_by(delta_f_raw, delta_t, participant) |> 
  summarise(response_f_rel = mean(response_f_rel, na.rm = TRUE))
  
xdata_agg_all <- xdata |> 
  group_by(delta_f_raw, delta_t) |> 
  summarise(response_f_rel = mean(response_f_rel, na.rm = TRUE))
  
descriptive_plot_03 <- 
ggplot(data = xdata_agg, 
       aes(x = delta_t,
           y = response_f_rel,
           color = delta_f_raw)) +
  # geom_line(data = xdata_agg, 
  #           aes(group = interaction(participant, delta_f_raw_st)),
  #         alpha = 0.2) +
  geom_point(data = xdata_agg_all) +
  geom_line(data = xdata_agg_all,
             aes(group = delta_f_raw),
             lwd = 2.5,
            color = "black") +
  geom_line(data = xdata_agg_all,
             aes(group = delta_f_raw),
             lwd = 2) +
  # scale_y_continuous(limits = c(-20,20),
  #                    breaks = seq(-20,20,10)) +
  scale_x_continuous(breaks = c(unique(xdata$delta_t))) +
  scale_color_viridis_c(option= "viridis") +
  geom_hline(yintercept = 0.25,
             lty = "dashed") +
  #scale_color_stepsn(colours = "viridis") +
  labs(x = "\nslope duration in ms",
       y = "average observed pitch response in Hz\n",
       color = "pitch difference") +
  theme_minimal()
  

descriptive_plot_03

```

Slight non-linear relationship with delta_t. Let's look at it in 3-dimensional space:

```{r}

# set up matrix
z_matrix <- matrix(xdata_agg_all$response_f_rel_st, nrow = 8, ncol = 6, byrow = TRUE)

x_axis <- unique(xdata_agg_all$delta_t)
y_axis <- unique(xdata_agg_all$delta_f_raw_st)

plot_ly(x = ~x_axis, y = ~y_axis, z = ~z_matrix, 
        type = "surface",
        colors = 'RdPu') |> 
  layout( scene = list(
    xaxis = list(title = 'Onset Duration (ms)', tickvals = x_axis, ticktext = x_axis),
    yaxis = list(title = 'Onset Frequency in STs', tickvals = list(-5,-4,-3,-2,-1,0,1,2,3,4,5), ticktext = list(-5,-4,-3,-2,-1,0,1,2,3,4,5)),
    zaxis = list(title = 'Response STs')
  ))

```

## Statistical model

Relevant parameters are estimated through the following model (following Marton's parametric suggestion) \# need justification for our model choice

`response_f_rel_st \~ a + b \* (( (delta_t - x0) / (x_max - x0) )\^inv_logit(c))`

a = assesses intercept, i.e. perceived pitch differences at `delta_t = 0` and `delta_f_raw_st = 0`

b = assesses linear relationship of `delta_t` and `response_f_rel_st` and how this relationship is modulated by `delta_f_raw_st`

c = delta_f_raw_st \# assess non-linear relationship of `delta_t` and `response_f_rel_st` and how this relationship is modulated by `delta_f_raw_st`

x0 \# ?

```{r}

# add ceiling
xdata <- xdata |> 
  mutate(
  #x_min=min(delta_t),
  x_max = max(delta_t)
)

# TO DO: Explore other functions?
## gumbel function?
## Naka-rushton?

# Model of the following basic structure:
# y ~ a + b*x^

# TO DO: Justify priors
# TO DO: random slopes for all these parameters? Computationally feasible?

# run model
parametric_model <- brm(
  bf(response_f_rel ~ a + b * (( (delta_t - x0) / (x_max - x0) )^inv_logit(c)),
     a ~ 1,
     b ~ delta_f_raw,
     c ~ delta_f_raw,
     x0 ~ 1,
     nl = T),
  data = xdata,
  file = "../models/model_nl_TR",
  prior = prior(normal(0,0.5), nlpar = "a")+
          prior(normal(0,0.5), nlpar = "b")+
          prior(normal(0,5), nlpar = "x0")+
          prior(normal(0,1), nlpar = "c", lb = 0, ub = 1),
  control = list(adapt_delta=0.99)
  )

# comment TR: takes 90 seconds on my machine, so computationally expensive
parametric_model
# model predicts mostly linear relationship

# posterior predictive check
pp_check(parametric_model)
# 1. not truly continuous measure, problem?
# 2. regardless, distribution looks spikier than gaussian

# check chains
plot(parametric_model, pars = "^b_") # okay

xdata$preds <- posterior_summary(posterior_epred(parametric_model))[,1]

# Visualize
model_plot_01 <- 
ggplot(data = xdata, 
       aes(x = delta_t,
           y = preds,
           color = delta_f_raw,
           group = delta_f_raw)) +
  geom_line(lwd = 2.5, color = "black") +
  geom_line(lwd = 2) +
   scale_y_continuous(limits = c(-20,20),
                     breaks = seq(-20,20,10)) +
  scale_x_continuous(breaks = c(unique(xdata$delta_t))) +
  scale_color_viridis_c(option= "viridis") +
  #scale_color_stepsn(colours = "viridis") +
  labs(x = "\nslope duration in ms",
       y = "predicted response in Hz\n",
       color = "pitch difference") +
  theme_minimal()
  
model_plot_01

data_v_model <- descriptive_plot_03 / model_plot_01 +
  plot_layout(guides = 'collect',
              axes = 'collect')

data_v_model

# pretty decent fit

```

In 3D:

```{r}

xdata_agg_pred <- xdata |> 
  group_by(delta_t, delta_f_raw_st) |> 
  summarise(preds = mean(preds))

# set up matrix
z_matrix <- matrix(xdata_agg_pred$preds, nrow = 6, ncol = 8, byrow = TRUE)

x_axis <- unique(xdata_agg_pred$delta_f_raw_st)
y_axis <- unique(xdata_agg_pred$delta_t)

plot_ly(x = ~x_axis, y = ~y_axis, z = ~z_matrix, 
        type = "surface",
        colors = 'RdPu') |> 
  layout( scene = list(
    yaxis = list(title = 'Onset Duration (ms)', tickvals = y_axis, ticktext = y_axis),
    xaxis = list(title = 'Onset Frequency in STs', tickvals = list(-5,-4,-3,-2,-1,0,1,2,3,4,5), ticktext = list(-5,-4,-3,-2,-1,0,1,2,3,4,5)),
    zaxis = list(title = 'Response STs')
  ))

```


## Pitch proportions

We explore how pitch perception changes as a function of `delta_t` and `delta_f.`

```{r}

xdata_agg <- xdata |> 
  group_by(delta_f_raw, delta_t, participant) |> 
  summarise(response_f_rel = mean(response_f_prop, na.rm = TRUE))
  
xdata_agg_all <- xdata |> 
  group_by(delta_f_raw, delta_t) |> 
  summarise(response_f_rel = mean(response_f_prop, na.rm = TRUE))
  
descriptive_plot_03 <- 
ggplot(data = xdata_agg, 
       aes(x = delta_t,
           y = response_f_rel,
           color = delta_f_raw)) +
  # geom_line(data = xdata_agg, 
  #           aes(group = interaction(participant, delta_f_raw_st)),
  #         alpha = 0.2) +
  # geom_line(data = xdata_agg_all,
  #            aes(group = delta_f_raw),
  #            lwd = 2.5,
  #           color = "black", alpha = 0.5) +
  geom_line(data = xdata_agg_all,
             aes(group = delta_f_raw),
             lwd = 2, 
            alpha = 0.5) +
  geom_point(data = xdata_agg_all, size = 3, color = "black") +
  geom_point(data = xdata_agg_all, size = 2) +
  # scale_y_continuous(limits = c(-20,20),
  #                    breaks = seq(-20,20,10)) +
  scale_x_continuous(breaks = c(unique(xdata$delta_t))) +
  scale_color_viridis_c(option= "viridis") +
  geom_hline(yintercept = 0.25,
             lty = "dashed") +
  #scale_color_stepsn(colours = "viridis") +
  labs(x = "\nslope duration in ms",
       y = "average observed ST perceived pitch proportion\n",
       color = "pitch difference") +
  theme_minimal()
  

descriptive_plot_03

```

```{r}


# add ceiling
xdata <- xdata |> 
  mutate(
  #x_min=min(delta_t),
  x_max = max(delta_t),
  # glissando threshold according to the 2/3 rule
  glissando = 2/3 * 0.75
)

# TO DO: Explore other functions?
## exponential
formula <- bf(response_f_prop ~ reference + (ceiling - reference) * exp(-k),
     reference ~ 0,
     ceiling ~ 0.4,
     k ~ 1 + delta_t,
     nl = T)

## gumbel function?

## Naka-rushton?

# Model of the following basic structure:
# y ~ a + b*x^
formula <- bf(response_f_prop ~ a + b * (( (delta_t - x0) / (x_max - x0) )^inv_logit(c)),
     a ~ 1,
     b ~ delta_f_raw,
     c ~ delta_f_raw,
     x0 ~ 1 ,
     nl = T)

# TO DO: Justify priors
# TO DO: random slopes for all these parameters? Computationally feasible?

# run model
parametric_model <- brm(
  formula,
  data = xdata,
  #file = "../models/model_prop_nl",
  prior = prior(normal(0,2), nlpar = "a")+
          prior(normal(0,2), nlpar = "b")+
          prior(normal(0,2), nlpar = "x0")+
          prior(normal(0,1), nlpar = "c", lb = 0, ub = 1),
  control = list(adapt_delta=0.99)
  )

# comment TR: takes 90 seconds on my machine, so computationally expensive
parametric_model
# model predicts mostly linear relationship

# posterior predictive check
pp_check(parametric_model)
# 1. not truly continuous measure, problem?
# 2. regardless, distribution looks spikier than gaussian

# check chains
plot(parametric_model, pars = "^b_") # okay

newdata <- expand_grid(delta_t = seq(40,250,10),
                       delta_f_raw = c(-40,-30,-20,-10,10,20,30,40),
                       x_max = 250)


# Visualize
model_plot_01 <- parametric_model |> 
  epred_draws(newdata = newdata,
              re_formula = NA) |> 
    group_by(delta_t, delta_f_raw) |> 
    summarise(pred_m = mean(.epred, na.rm = TRUE),
              pred_0025 = quantile(.epred, prob = 0.025),
              pred_0975 = quantile(.epred, prob = 0.975),
              threshold = ifelse(pred_m > 0.25, "yes", "no")) |> 
ggplot(aes(x = delta_t,
           y = pred_m,
           group = delta_f_raw,
           color = delta_f_raw
           )
       ) +
  geom_line(lwd = 2.5) +
  geom_hline(yintercept = 0.25, 
             lty = "dashed",
             colour = "black") +
  scale_x_continuous(breaks = c(unique(xdata$delta_t))) +
  scale_color_viridis_c(option= "viridis") +
  labs(x = "\nslope duration in ms",
       y = "predicted proportion of perceived pitch to hear pitch\n",
       color = "pitch difference") +
  theme_minimal()
  
model_plot_01

data_v_model <- descriptive_plot_03 + model_plot_01 +
  plot_layout(guides = 'collect',
              axes = 'collect')

data_v_model

# pretty decent fit

```
