---
title: "Data Cleaning"
author: Timo Roettger
date: 2026-06-06
format: html
editor: visual
---

```{r}
#| include: FALSE

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

```{r}
#| label: setup
#| include: FALSE


# read in libraries
if (!require("pacman")) install.packages("pacman")
pacman::p_load(viridis, tidyverse, rstudioapi, scales, knitr, kableExtra)

# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# set the current working directory to the one where this file lives
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

```

```{r}
#| label: load-data
#| include: FALSE

# latest pilot from 2026/06/01
xdata <- read.csv("../data/raw/merged_260601.csv", colClasses = c("response_list" = "character"))

# before exclusion
all_data <- nrow(xdata) # 5040

# how many subjects
length(unique(xdata$participantID)) # 40

# data points per subject
xtabs(~xdata$participantID)

```

```{r}
#| label: data-wrangling

xdata <- xdata |> 
  mutate(
         # make delta_f positive / negative
         delta_f_raw = ifelse(stimuli_direction == "p", delta_f, delta_f * -1),
         # make delta_f_raw into STs (reference 150Hz)
         delta_f_raw_st = 12 * log2((150-delta_f_raw)/150),
         # make response_f relative to baseline 150hz      
         response_f_rel = response_f - 150,
         # make response_f into STs (reference 150Hz)
         response_f_rel_st = 12 * log2((150-response_f_rel)/150),
         # make response_f absolute without sign differences
         response_f_rel_abs = abs(response_f_rel),
         # relative to target
         response_f_prop = response_f_rel / delta_f,
         # calculate actual f0 of presented tonal onset
         presented_f = 150 + delta_f_raw,
         presented_f_st = 12 * log2(presented_f/150),
         # recovered f0
         recovered_f0 = response_f_rel / delta_f_raw,
         # make is.correct as.numeric
         correct = as.numeric(is_correct),
         # make part ID a char
         participantID = as.character(participantID)
         ) 

# rename participant column to identify in plots
unique_IDs <- unique(xdata$participantID)
n_unique   <- length(unique_IDs)

# generate a pool of unique, short random strings
# This creates 4-character alphanumeric strings (e.g., "4x9b")
set.seed(42) 
random_strings <- replicate(n_unique, {
  paste(sample(c(0:9, letters), 4, replace = TRUE), collapse = "")
})

# create a named lookup dictionary (Maps: "apple" -> "4x9b")
lookup_table <- setNames(random_strings, unique_IDs)

# swap the original strings for the random ones
xdata$participantID <- lookup_table[xdata$participantID]

```

# Data cleaning

In the following, we will explore catch trials, control trials, reaction times and dynamics related to how often and how people moved the slider to clean the data. We will suggest and implement exclusion criteria.

## Attention checks

There were 4 attention checks in which participants were asked to indicate a particular slide position. We flag all responses that are more than one slider position of the target slider position.

```{r}
#| label: catch-trials

# catch trials (four catch trials)
xdata_catch <- xdata |> 
  # we implement a lenient criterion that considers imprecise responses valid
  mutate(catch1 = ifelse(catch_tone_loop_res %in% c(0,1), 1, 0),
         catch2 = ifelse(catch_tone_end_res %in% c(21,20), 1, 0),
         catch3 = ifelse(catch_cv_loop_res %in% c(0,1), 1, 0),
         catch4 = ifelse(catch_cv_end_res %in% c(0,1), 1, 0)) |> 
  group_by(participantID) |> 
  summarise(catch1 = mean(catch1),
            catch2 = mean(catch2),
            catch3 = mean(catch3),
            catch4 = mean(catch4),
            all_catch = sum(catch1, catch2, catch3, catch4))

# how many need to be removed?
## we remove all participants who have not correctly responded to at least 3 out of 4 catch trials
catch_trial_exclusions <- round(((nrow(xdata_catch[xdata_catch$all_catch < 3,]) / nrow(xdata_catch)) * 100), 1) # 15% data loss 

# merge for later
xdata <- full_join(xdata, xdata_catch)

# print table
xdata_catch |> 
  group_by(all_catch) |> 
  count()

```

There were `r nrow(xdata_catch[xdata_catch$all_catch < 3,])` participants (`r catch_trial_exclusions` %) who failed to correctly respond to 3 out of 4 catch trials. These participants are excluded.

```{r}
#| label: exclude-catch-trials
#| include: FALSE

# before exclusion
all_data # 5040

# remove all participants that did not get the catch trials correctly, i.e. both in pilot 1 and 3 out 4 in pilot 2
xdata_filtered <- xdata |>
  filter(all_catch > 2)

# data loss?
filter1 <- nrow(xdata_filtered) # 4284
(all_data - filter1) / all_data # ~15% participants excluded

```

## Control trials

There were 14 trials per participant that presented a flat tone (`delta_t & delta_f = 0`) for which the correct response was 0 Hz.

```{r}
#| label: "distribution-control"
#| fig-cap: "Distribution of control trials on flat tones."

xdata_filtered |> 
  filter(delta_t == 0) |> 
  ggplot(aes(x = response_f_rel)) + 
  geom_histogram(aes(y = after_stat(count) / sum(after_stat(count))), 
                 binwidth = 5, 
                 fill = "grey", 
                 color = "black") +
  stat_bin(binwidth = 5, 
           geom = "text", 
           aes(y = after_stat(count) / sum(after_stat(count)), 
               label = percent(after_stat(count) / sum(after_stat(count)), 
                               accuracy = 1)), 
           vjust = -0.5, 
           size = 3.5) +
  scale_x_continuous(breaks = seq(-50,50,5)) +
  labs(x = "Responses to control trials",
       y = "Proportion") +
  theme_minimal()

```

Most responses to the control trials are correct followed by the adjacent pitch responses (-5 Hz and 5 Hz) with a slight bias towards higher values. We expect that some people might have slight response biases (hearing a flat tone as slightly rising or falling) but consistently judge the flat pitch with some level of symmetric-ish noise. Ideally we identify participants that do not show such a pattern, indicating either inattentiveness or other non-desirable response behavior.

```{r}
#| label: "distribution-control-participants"
#| fig-cap: "Distribution of control trials on flat tones per participant."

xdata_filtered |> 
  filter(delta_t == 0) |> 
  ggplot(aes(x = response_f_rel)) + 
  geom_histogram(aes(y = after_stat(count) / sum(after_stat(count))), 
                 binwidth = 5, 
                 fill = "white", 
                 color = "black") +
  geom_histogram(aes(y = after_stat(count) / sum(after_stat(count))), 
                 binwidth = 5, 
                 fill = "white", 
                 color = "black") +
  scale_x_continuous(breaks = seq(-50,50,50)) +
  facet_wrap(~participantID) +
  labs(x = "Responses to control trials",
       y = "Proportion") +
  theme_minimal()

```

We can see that most distributions look unimodal-ish. Some participants have wider distributions than others, so we need to take this into account. As a first step we calculate the mean and SD of control responses per participant and store the information for later.

```{r}
#| label: "calc-control-participants"

control_trials_distributions <- 
  xdata_filtered |> 
  filter(delta_t == 0) |> 
  group_by(participantID) |> 
  summarise(control_response_bias = mean(response_f_rel),
            control_response_noise = sd(response_f_rel))

print(control_trials_distributions |>  arrange(desc(abs(control_response_noise))))
      
```

Response biases vary quite a bit between participants but are generally very small. Consistency (measured as standard deviation) of their control responses also varies a lot. We suggest to simply flag all control trials with extreme values (below -10 Hz and above 10 Hz).

```{r}
#| label: "flag-control-trials"

extreme_controls <- xdata_filtered |> 
  # merge and filter for control trials
  full_join(control_trials_distributions) |> 
  filter(delta_t == 0) |> 
  # group by participant and calculate total extreme trials
  group_by(participantID) |> 
  summarize(
    # counts how many times the absolute value exceeds 10
    n_extreme = sum(abs(response_f_rel) > 10, na.rm = TRUE), 
    .groups = "drop"
  ) |> 
  # apply exclusion rule based on the total count
  mutate(
    exclude_due_extreme = if_else(n_extreme > 4, "exclude", "keep")
  )

xdata_filtered <- xdata_filtered |> full_join(extreme_controls)

```

We will remove participants who give extreme control responses in more than 30% of control trials corresponding to 4 out of 14 control trials.

```{r}
#| label: exclude-control-fails
#| include: FALSE

# before exclusion
filter1 # 4284

# remove all participants that did exhibit over 30% extreme control responses
xdata_filtered <- xdata_filtered |>
  filter(exclude_due_extreme != "exclude")

# data loss?
filter2 <- nrow(xdata_filtered) # 4284
(filter1 - filter2) / filter1 # ~0% participants excluded

```

There were `r nrow(extreme_controls[extreme_controls$exclude_due_extreme == "exclude",])` participants who responded more than 4 times to the control task with extreme values (`abs(10)` Hz).

## Move dynamics

Participants were able to move the scale position as often as they wished. Each move was recorded as a string in `response_list`.

```{r}
#| label: wrangle-trial-sequence
#| include: FALSE

# Define max amount of moves in data set
max_moves <- max(lengths(strsplit(xdata_filtered$response_list, ","))) # 47

# double-check by counting commas in string
max_moves == max(str_count(xdata_filtered$response_list, ",")) + 1 # TRUE

# Define the maximum column names
num_cols <- paste0("move_sequence_", 1:max_moves)

# Step 1: Split the comma-separated string into 30 distinct columns
xdata_filtered_temp <- xdata_filtered |> 
  separate_wider_delim(
    cols = response_list,
    delim = ",",
    names = num_cols,
    too_few = "align_start",  # Fills missing spaces with NA up to 30 cols
    too_many = "drop"         # Drops elements if there happen to be more than 30
  ) |> 
  # Convert all the newly created columns from text to numeric
  mutate(across(all_of(num_cols), as.numeric))


# Step 2: Calculate numerical distance between adjacent pairs
# Running a vector-based loop 
for (i in 1:29) {
  col_left  <- paste0("move_sequence_", i)
  col_right <- paste0("move_sequence_", i + 1)
  dist_name <- paste0("dist_", i, "_", i + 1)
  
  # Calculate absolute difference; returns NA if either column is NA
  xdata_filtered_temp[[dist_name]] <- (xdata_filtered_temp[[col_right]] - xdata_filtered_temp[[col_left]])
}


# Step 3: Calculate the mean and median distance for each row
xdata_filtered_temp_final <- xdata_filtered_temp %>%
  rowwise() %>%
  mutate(
    mean_distance   = mean(c_across(starts_with("dist_")), na.rm = TRUE),
    median_distance = median(c_across(starts_with("dist_")), na.rm = TRUE)
  ) %>%
  ungroup() 

```

```{r}
#| label: hist-trial-sequence
#| fig-cap: "Distribution of slider distance between subsequent moves."

# long format to plot
xdata_filtered_temp_final_long <- xdata_filtered_temp_final |> 
  pivot_longer(
    cols = starts_with("dist_"),    # Target all the newly created distance columns
    names_to = "distance_pair",     # Column to store the old column names (e.g., "dist_1_2")
    values_to = "distance_value",   # Column to store the actual calculated distances
    values_drop_na = TRUE           # Automatically drops the padded NA values
  )

xdata_filtered_temp_final_long |> 
  ggplot(aes(x = distance_value)) + 
  geom_histogram(fill = "grey",
                 color = "black", 
                 binwidth = 1) +
  labs(x = "slider distance between subsequent moves") +
  theme_minimal()

```

As expected, moves were mostly one step up or down to adjacent slider positions, resulting in a symmetric distribution around zero. High values indicate a strong jump and can be expected to occur at the beginning of a trial but not at the end. Let's look at the final move which ideally should not be a big jump.

```{r}
#| label: final-move
#| fig-cap: "Distribution of slider distanct only for the last move before submitting."

dist_matrix <- xdata_filtered_temp_final |>  
  select(starts_with("dist_"))

# use apply to find the last non-NA value in each row
xdata_filtered_temp_final$final_distance <- 
  apply(dist_matrix, 1, function(row) {
  valid_indices <- which(!is.na(row))
  if(length(valid_indices) == 0) return(NA) # In case a row is completely empty
  
  last_index <- max(valid_indices)
  return(row[last_index])
})

# histogram
xdata_filtered_temp_final |> 
  ggplot(aes(x = final_distance)) + 
  geom_histogram(fill = "grey",
                 color = "black", 
                 binwidth = 1) +
  labs(x = "slider distance between the prefinal and final move (before submitting)") +
  theme_minimal()

```

There are indeed some larger values, up to jumps of 10 and higher. I think we should exclude moves that are beyond a certain threshold. A jump of 5 slider positions corresponds to 25 Hz.

```{r}
#| label: exclude-too-large-final

# before exclusion
# filter2 # 4284

# remove all trials with too large final moves
xdata_filtered <- xdata_filtered_temp_final |>
  filter(final_moves == 1 |
           final_distance < 5 & 
           final_distance > -5)

# data loss?
filter3 <- nrow(xdata_filtered) # 4284
excluded_final <- round((((filter2 - filter3) / filter2) * 100), 1) # ~1.8% data excluded

```

There were `r excluded_final`% of trials in which the final slider change was above 25 Hz. We exclude these trials.

## Minimal reaction time

```{r}
#| label: exclude-minimal-rt

# remove all trials that are below minimal reaction time which corresponds to the stimulus length: 
xdata_filtered <- xdata_filtered |> 
  filter(response_time > 0.15)

# data loss?
filter4 <- nrow(xdata_filtered) # 4207
minimal_rt <- round((((filter3 - filter4) / filter3) * 100), 1) # 0% exclusion

```

All trials with a response time lower than the length of the auditory stimuli (\< 150 ms) are excluded on a trial-by-trial bases. This led to an exclusion of additional `r minimal_rt`%.

## Reaction time and amount of moves

Participants were able to make multiple moves before arriving at their final choice. How much did they move?

```{r}
#| label: move-distribtuion
#| fig-cap: "Distribution of amount of moves overall"

xdata_filtered |> 
  ggplot(aes(x = final_moves)) + 
  geom_histogram(color = "black",
                 fill = "white") +
  labs(x = "amount of moves before submitting the answer",
       y = "") +
  theme_minimal()

```

Mostly 1-2 moves with a heavy right tale. Presumably this changed throughout the experiment.

```{r}
#| label: move-distribution-trial
#| fig-cap: "Distribution of amount of moves across trials"

xdata_filtered <- 
  xdata_filtered |> 
  # make block information
  mutate(block = ifelse(!is.na(tone_trials.thisN), "first", "second"),
         trials = ifelse(block == "first", tone_trials.thisN, cv_trials.thisN + 63)) 

xdata_filtered |> 
  group_by(trials) |> 
  summarise(mean_move = mean(final_moves, na.rm = TRUE))  |> 
  ggplot(aes(x = trials, y = mean_move)) + 
  geom_point(pch = 21,
             size = 2,
              color = "white",
              fill = alpha("black", 0.5)) +
  geom_smooth(method = "lm", 
              se = FALSE) +
  labs(x = "trial number",
       y = "average amount of moves") +
  theme_minimal()

```

The amount of moves decreases over the experiment as expected. Given that, we expect reaction times to substantially decrease over the course of the experiment. When excluding trials based on reaction times, we will take this into consideration.

```{r}
#| label: rt-distribution-trial
#| fig-cap: "Distribution of RT across trials"

xdata_filtered |> 
  group_by(trials) |> 
  summarise(mean_rt = mean(response_time, na.rm = TRUE))  |> 
  ggplot(aes(x = trials, y = mean_rt)) + 
  geom_point(pch = 21,
             size = 2,
              color = "white",
              fill = alpha("black", 0.5)) +
  geom_smooth(method = "lm", 
              se = FALSE) +
  labs(x = "trial number",
       y = "average reaction time in seconds") +
  theme_minimal()

```

```{r}
#| label: exclude-outliers

# now removing outliers for each participant in terms of RT
xdata_filtered <- xdata_filtered |> 
  mutate(trial_bracket = cut(trials, 
                             breaks = 4, 
                             labels = seq(1,4,1))
  ) |> 
  group_by(participantID, trial_bracket) |> 
  # 4 SDs trim
  filter(response_time < (median(response_time) + 4 * sd(response_time)) & 
         response_time > (median(response_time) - 4 * sd(response_time))) |>
  ungroup()

# data loss?
filter5 <- nrow(xdata_filtered) # 4152
outlier_rt <- round((((filter4 - filter5) / filter4) * 100), 1) # 1.3% exclusion

```

To account for outliers due to attentional lapses and distractions, we further removed trials that elicited reaction times that were more than 4 SDs from the median (`r outlier_rt`%) for each participant for each quarter of the experiment (trials binned into 4 equal bins), respectively.

```{r}
#| label: cleaned_rt-distribution-trial
#| fig-cap: "Distribution of RT across trials after data trimming"

xdata_filtered |> 
  group_by(trials) |> 
  summarise(mean_rt = mean(response_time, na.rm = TRUE))  |> 
  ggplot(aes(x = trials, y = mean_rt)) + 
  geom_point(pch = 21,
             size = 2,
              color = "white",
              fill = alpha("black", 0.5)) +
  geom_smooth(method = "lm", 
              se = FALSE) +
  labs(x = "trial number",
       y = "average reaction time in seconds") +
  theme_minimal()

```

After these exclusions, we assess how many valid data points remain for each participant. If a participant has less than 80% of trials left, we exclude that participant from the sample as well.

```{r}
#| label: final-subject-removal

data_loss_threshold = 0.8

# How many data points left per participant?
xdata_final_exclusion <- 
  xdata_filtered |> 
  group_by(participantID) |> 
  count() |> 
  mutate(too_much_data_loss = ifelse((n / 126) > data_loss_threshold, "keep", "exclude"))

# How many need to be excluded?
excluded_at_end <- 
nrow(xdata_final_exclusion |> filter(too_much_data_loss == "exclude"))

# merge for later
xdata_filtered <- full_join(xdata_filtered, xdata_final_exclusion) |> 
  filter(too_much_data_loss == "keep")

# data loss?
filter6 <- nrow(xdata_filtered) # 4065
final_exclusions <- round((((filter5 - filter6) / filter6) * 100), 1) # 2.1% exclusion


```

This resulted in the exclusion of `r excluded_at_end` more participants.

```{r}
#| label: overall-exclusion

exclusions <- round((((all_data - filter6) / all_data) * 100), 1)  # 19.3% exclusion

```

All in all, we excluded `r exclusions`% of data.

```{r}

## Store clean dataframe
xdata_wrangled <- xdata_filtered |> 
  select(labID, participantID, 
         # demographics
         birthyear, sex, 
         # origin
         residence_country, country_of_origin,
         # language
         language, 
         # music
         instrument, instrument_training, musical_level, 
         musical_training_year, music_hour, group_singing,
         group_singing_year, solo_vocal, solo_vocal_year,
         music_hour, met_score,
         # response
         final_rating, response_f_rel,
         stimuli_direction, response_dir,
         response_f, correct,
         recovered_f0,
         # predictors
         delta_t, delta_f, 
         delta_f_raw
         ) 
```

```{r}
#| label: write-data

# write clean data after exclusion
write_csv(xdata_wrangled, "../data/derived/data_after_exclusion.csv")

```
