---
title: "Data Cleaning"
author: Timo Roettger
date: 2026-06-06
format:
  html:
    embed-resources: true
    toc: true
editor: visual
---

```{r}
#| include: FALSE

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

```{r}
#| label: setup
#| include: FALSE

# --- Libraries -------------------------------------------------------------
# pacman checks whether each package is installed and installs it if missing,
# then loads everything in one call. Convenient, but note that it means the
# script can silently install new packages on someone else's machine the
# first time they knit it - worth being aware of for a shared project.
if (!require("pacman")) install.packages("pacman")
pacman::p_load(viridis, tidyverse, rstudioapi, scales, jsonlite, knitr, kableExtra, here)

# global seed for random draws
set.seed(42)

```

```{r}
#| label: load-data
#| include: FALSE

# latest pilot from 2026/06/01
xdata <- read.csv(here("data", "raw", "merged_260601.csv"), 
                  colClasses = c("response_list" = "character"))

# before exclusion
all_data <- nrow(xdata) # 5040

# how many subjects
length(unique(xdata$participantID)) # 40

# data points per subject
xtabs(~xdata$participantID)

# sanity check if all participants have the same amount of trials
stopifnot(all(table(xdata$participantID) == nrow(xdata) / length(unique(xdata$participantID))))

```

# Data cleaning

In the following, we explore catch trials, control trials, reaction times, and slider-movement dynamics in order to define and apply data-quality exclusion criteria before any statistical modeling. The pipeline proceeds in four stages: (1) recoding raw variables into the units used in later analyses, (2) flagging and excluding participants who failed attention checks, (3) flagging and excluding participants/trials with implausible control-trial or slider-movement behavior, and (4) trimming reaction-time outliers. At the end, we report the cumulative data loss and write out a single cleaned dataset for downstream modeling.

## Recoding raw variables

Before any exclusion, we first recode the raw experiment variables into the units used throughout the rest of the analysis: semitones (relative to a 150 Hz reference), signed pitch shifts, and a few convenience columns.

```{r}
#| label: data-wrangling

xdata <- xdata |>
  mutate(
         # `stimuli_direction` codes whether the pitch shift was upward ("p")
         # or downward; this recovers the *signed* magnitude of the shift
         # from the (presumably unsigned) raw `delta_f` column.
         delta_f_raw = ifelse(stimuli_direction == "p", delta_f, delta_f * -1),
         # convert the signed pitch shift (delta_f_raw, in Hz) into
         # semitones (ST), using 150 Hz as the reference pitch
         delta_f_raw_st = 12 * log2((150+delta_f_raw)/150),

         # center the observed response pitch on the same 150 Hz reference
         response_f_rel = response_f - 150,

         # convert the centered response into semitones, same reference
         response_f_rel_st = 12 * log2((150+response_f_rel)/150),

         # unsigned/absolute deviation from baseline, ignoring direction
         response_f_rel_abs = abs(response_f_rel),

         # response magnitude expressed as a proportion of the presented
         # pitch shift (delta_f) - i.e., how much of the intended pitch
         # change the participant actually reproduced
         response_f_prop = ifelse(delta_f == 0, NA_real_, response_f_rel / delta_f),

         # an attempt to "recover" the underlying f0 shift from the
         # response, relative to the *raw* (signed) pitch shift
         recovered_f0 = ifelse(delta_f_raw == 0, NA_real_, response_f_rel / delta_f_raw),

         # convert is_correct (presumably logical TRUE/FALSE) to numeric
         # 0/1, useful for modeling accuracy as a proportion/rate
         correct = as.numeric(is_correct),

         # store participant ID as a character rather than whatever type
         # it was read in as (guards against accidental numeric coercion
         # or factor-level issues later)
         participantID = as.character(participantID)
         )

# --- Randomly distribute control trials across delta_t levels --------------
# Each participant completed 14 control trials (delta_t = delta_f = 0). For
# later modeling, we want these control trials spread evenly across the 7
# `delta_t` levels (20, 40, ..., 140 ms), 2 control trials per level per
# participant, assigned in random order.

# sanity check if not exactly 14 controls
stopifnot(all(table(xdata$participantID[xdata$delta_t == 0]) == 14))

xdata <- xdata  |>
  group_by(participantID) |>
  mutate(
    # 1. Build a shuffled 14-item pool per participant: each of the 7
    #    delta_t levels appears exactly twice, in random order. Because
    #    the seed was fixed once at the top of the script (see setup
    #    chunk), this shuffle is now reproducible across re-knits.
    shuffled_pool = list(sample(rep(seq(20, 140, 20), each = 2))),

    # 2. Number this participant's control trials 1 through 14, in the
    #    order they appear in the data; non-control rows get NA here.
    control_index = NA_integer_,
    control_index = ifelse(delta_t == 0, cumsum(delta_t == 0), NA)
  ) |>
  rowwise() |>
  mutate(
    # 3. For each control row, pull the corresponding item out of this
    #    participant's shuffled pool and use it as the (fake) delta_t.
    #    Non-control rows are left with their original delta_t unchanged.
    delta_t = ifelse(
      !is.na(control_index),
      shuffled_pool[[control_index]],
      delta_t
    )
  ) |>
  ungroup() |>
  select(-shuffled_pool, -control_index) # clean up temporary helper columns

# --- Pseudonymize participant IDs ------------------------------------------
# Replace the original participantID (a long UUID/Prolific-style string)
# with a short random 4-character code, purely so that later plots (facets,
# axis labels, etc.) are readable.
unique_IDs <- unique(xdata$participantID)
n_unique   <- length(unique_IDs)

# generate a pool of unique, short random strings
# This creates 4-character alphanumeric strings (e.g., "4x9b")
random_strings <- replicate(n_unique, {
  paste(sample(c(0:9, letters), 4, replace = TRUE), collapse = "")
})

# sanity check if two participant received the same name
stopifnot(length(unique(random_strings)) == n_unique)                     

# create a named lookup dictionary (maps original ID -> short code)
lookup_table <- setNames(random_strings, unique_IDs)

# store look up table if ever needed
write.csv(lookup_table, here::here("data", "derived", "participant_mapping.csv"))

# swap the original identifiers for the random ones throughout the dataset
xdata$participantID <- lookup_table[xdata$participantID]

```

## Attention checks

There were 4 attention checks in which participants were asked to indicate a particular slide position. We flag all responses that are more than one slider position of the target slider position.

```{r}
#| label: catch-trials

# catch trials (four catch trials)
xdata_catch <- xdata |>
  # we implement a lenient criterion that considers imprecise responses valid
  # (i.e., a response within 1 slider position of the correct answer counts
  # as "correct" for the purposes of this check)
  mutate(catch1 = ifelse(catch_tone_loop_res %in% c(0,1), 1, 0),
         catch2 = ifelse(catch_tone_end_res %in% c(21,20), 1, 0),
         catch3 = ifelse(catch_cv_loop_res %in% c(0,1), 1, 0),
         catch4 = ifelse(catch_cv_end_res %in% c(0,1), 1, 0)) |>
  group_by(participantID) |>
  summarise(catch1 = mean(catch1),
            catch2 = mean(catch2),
            catch3 = mean(catch3),
            catch4 = mean(catch4),
            # total number of the 4 catch checks passed, per participant
            catch_amount = sum(catch1, catch2, catch3, catch4)) |>
  ungroup()

# how many participants need to be removed?
## we remove all participants who have not correctly responded to at least 3 out of 4 catch trials
catch_trial_exclusions <- round(((nrow(xdata_catch[xdata_catch$catch_amount < 3,]) / nrow(xdata_catch)) * 100), 1) # 15% data loss

# merge the per-participant catch-trial summary back onto the full trial-level data
xdata <- full_join(xdata, xdata_catch, by = "participantID")

# print table: how many participants passed 0, 1, 2, 3, or 4 catch trials?
xdata_catch |>
  group_by(catch_amount) |>
  count()

```

There were `r nrow(xdata_catch[xdata_catch$catch_amount < 3,])` participants (`r catch_trial_exclusions` %) who failed to correctly respond to 3 out of 4 catch trials.

These participants are excluded.

```{r}
#| label: exclude-catch-trials
#| include: FALSE

# before exclusion
all_data # 5040

xdata_after_attention_check <- xdata |> filter(catch_amount > 2)

# data loss?
passed_attention_checks <- nrow(xdata_after_attention_check) # 4284
(all_data - passed_attention_checks) / all_data # ~15% participants excluded
```

## Control trials

There were 14 trials per participant that presented a flat tone (`delta_t & delta_f = 0`) for which the correct response was 0 Hz. We use these control trials both to understand overall response bias/noise and to flag participants who show implausible response patterns on the easiest possible trials.

```{r}
#| label: "distribution-control"
#| fig-cap: "Distribution of control trials on flat tones."

# distribution of responses to control (flat-tone) trials, pooled across all
# participants who passed the attention checks
xdata_after_attention_check |>
  filter(delta_f_raw == 0) |>
  ggplot(aes(x = response_f_rel)) +
  geom_histogram(aes(y = after_stat(count) / sum(after_stat(count))),
                 binwidth = 5,
                 fill = "grey",
                 color = "black") +
  stat_bin(binwidth = 5,
           geom = "text",
           aes(y = after_stat(count) / sum(after_stat(count)),
               label = percent(after_stat(count) / sum(after_stat(count)),
                               accuracy = 1)),
           vjust = -0.5,
           size = 3.5) +
  scale_x_continuous(breaks = seq(-50,50,5)) +
  labs(x = "Responses to control trials",
       y = "Proportion") +
  theme_minimal()

```

Most responses to the control trials are correct followed by the adjacent pitch responses (-5 Hz and 5 Hz) with a slight bias towards higher values. We expect that some people might have slight response biases (hearing a flat tone as slightly rising or falling) but consistently judge the flat pitch with some level of symmetric-ish noise. Ideally we identify participants that do not show such a pattern, indicating either inattentiveness or other non-desirable response behavior.

```{r}
#| label: "distribution-control-participants"
#| fig-cap: "Distribution of control trials on flat tones per participant."

# same plot as above, but faceted by participant, to check whether any
# single participant's control-trial responses look qualitatively different
# from the group pattern (e.g. bimodal, wildly dispersed, or off-center)
xdata_after_attention_check |>
  filter(delta_f_raw == 0) |>
  ggplot(aes(x = response_f_rel)) +
  geom_histogram(aes(y = after_stat(count) / sum(after_stat(count))),
                 binwidth = 5,
                 fill = "white",
                 color = "black") +
  # NOTE: this second geom_histogram() layer is an exact duplicate of the
  # one above it (identical binwidth/fill/color/aes) - it has no visible
  # effect (drawing the same bars twice on top of each other) but is
  # redundant. SUGGESTION: safe to delete one of the two identical layers.
  geom_histogram(aes(y = after_stat(count) / sum(after_stat(count))),
                 binwidth = 5,
                 fill = "white",
                 color = "black") +
  scale_x_continuous(breaks = seq(-50,50,50)) +
  facet_wrap(~participantID) +
  labs(x = "Responses to control trials",
       y = "Proportion") +
  theme_minimal()
```

We can see that most distributions look unimodal-ish. Some participants have wider distributions than others, so we need to take this into account. Let's calculate the mean and SD of control responses per participant and store the information for later.

```{r}
#| label: "calc-control-participants"

# per-participant summary of control-trial behavior: average bias
# (systematic over/under-estimation of the flat tone) and noise
# (variability/consistency of responses to the same, unchanging stimulus)
control_trials_distributions <-
  xdata_after_attention_check |>
  filter(delta_f_raw == 0) |>
  group_by(participantID) |>
  summarise(control_response_bias = mean(response_f_rel),
            control_response_noise = sd(response_f_rel))

print(control_trials_distributions |>  arrange(desc(abs(control_response_noise))))
```

Response biases vary quite a bit between participants but are generally very small. Consistency (measured as standard deviation) of their control responses also varies a lot. Having a strict exclusion criterion on a by-trial bias, could be too strict. So we define extreme responses to the control as a abs(20) response, which corresponds to four slider positions above/below the correct response.

```{r}
#| label: "flag-control-trials"

# what we consider an extreme control response
extreme_control_response <- 20
exclusion_threshold <- 4

extreme_controls <- xdata_after_attention_check |>
  # merge in the per-participant bias/noise summary and keep only control trials
  full_join(control_trials_distributions, by = "participantID") |>
  filter(delta_f_raw == 0) |>
  # group by participant and count how many of their control trials were "extreme"
  group_by(participantID) |>
  summarize(
    # counts how many times the absolute value exceeds extreme_control_response
    n_extreme = sum(abs(response_f_rel) > extreme_control_response, na.rm = TRUE),
    .groups = "drop"
  ) |>
  # apply exclusion rule based on the total count
  mutate(
    exclude_too_many_extreme_controls = if_else(n_extreme > exclusion_threshold, "exclude", "keep")
  )

xdata_after_attention_check <- full_join(xdata_after_attention_check, extreme_controls, by = "participantID")
```

We will remove participants who give extreme control responses in more than 30% of control trials corresponding to 4 out of 14 control trials.

```{r}
#| label: exclude-control-fails
#| include: FALSE

# before exclusion
passed_attention_checks # 4284

# remove all participants that did exhibit over 30% extreme control responses
xdata_after_controls_check <- xdata_after_attention_check |>
  filter(exclude_too_many_extreme_controls != "exclude")

# data loss?
control_trials_approved <- nrow(xdata_after_controls_check) # 4284
(passed_attention_checks - control_trials_approved) / passed_attention_checks # ~0% participants excluded
```

There were `r nrow(extreme_controls[extreme_controls$exclude_too_many_extreme_controls == "exclude",])` participants who responded more than 4 times to the control task with extreme values (`abs(20)` Hz).

## Move dynamics

Participants were able to move the scale position as often as they wished. Each move was recorded as a string in `response_list`. Here we reconstruct the sequence of moves for each trial to check whether people were making large, erratic jumps (especially right before submitting their final answer).

```{r}
#| label: wrangle-trial-sequence
#| include: FALSE

# work out the maximum number of moves anyone made in a single trial, so we
# know how many "move_sequence_" columns we need to create
max_moves <- max(lengths(strsplit(xdata_after_controls_check$response_list, ","))) # 47

# double-check by counting commas in string (should always be one fewer
# comma than the number of moves, since commas only separate items)
max_moves == max(str_count(xdata_after_controls_check$response_list, ",")) + 1 # TRUE

# build column names move_sequence_1, move_sequence_2, ..., move_sequence_max_moves
num_cols <- paste0("move_sequence_", 1:max_moves)

# Step 1: split each participant's comma-separated move string into one
# column per move (padding with NA for trials with fewer moves than the max)
xdata_added_move_infos <- xdata_after_controls_check |>
  separate_wider_delim(
    cols = response_list,
    delim = ",",
    names = num_cols,
    too_few = "align_start",  # pads missing positions with NA
    too_many = "drop"         # (safety net; shouldn't trigger given max_moves above)
  ) |>
  # convert all newly created move columns from text to numeric
  mutate(across(all_of(num_cols), as.numeric))

# Step 2: for each pair of consecutive moves, calculate the (signed)
# distance the slider traveled between them
for (i in 1:(max_moves - 1)) {
  col_left  <- paste0("move_sequence_", i)
  col_right <- paste0("move_sequence_", i + 1)
  dist_name <- paste0("dist_", i, "_", i + 1)

  # signed difference; NA if either move column is NA (i.e. the trial ended
  # before reaching this pair of moves)
  xdata_added_move_infos[[dist_name]] <- (xdata_added_move_infos[[col_right]] - xdata_added_move_infos[[col_left]])
}

# Step 3: for each trial, summarize the move-to-move distances with a mean
# and median (across however many move-pairs that trial actually had)
xdata_added_move_infos <- xdata_added_move_infos %>%
  rowwise() %>%
  mutate(
    mean_distance   = mean(c_across(starts_with("dist_")), na.rm = TRUE),
    median_distance = median(c_across(starts_with("dist_")), na.rm = TRUE)
  ) %>%
  ungroup()
```

```{r}
#| label: hist-trial-sequence
#| fig-cap: "Distribution of slider distance between subsequent moves."

# reshape the many "dist_*" columns into long format for plotting
xdata_added_move_infos_long <- xdata_added_move_infos |>
  pivot_longer(
    cols = starts_with("dist_"),    # target all the distance columns created above
    names_to = "distance_pair",     # store the original column name (e.g., "dist_1_2")
    values_to = "distance_value",   # store the actual calculated distance
    values_drop_na = TRUE           # drop the padded NA rows automatically
  )

xdata_added_move_infos_long |>
  ggplot(aes(x = distance_value)) +
  geom_histogram(fill = "grey",
                 color = "black",
                 binwidth = 1) +
  labs(x = "slider distance between subsequent moves") +
  theme_minimal()
```

As expected, moves were mostly one step up or down to adjacent slider positions, resulting in a symmetric distribution around zero. High values indicate a strong jump and can be expected to occur at the beginning of a trial but not at the end. Let's look at the final move which ideally should not contains big jumps.

```{r}
#| label: final-move
#| fig-cap: "Distribution of slider distanct only for the last move before submitting."

# isolate just the distance columns so we can scan each row for its last
# non-missing value (i.e., the distance of the *final* move before submitting)
dist_matrix <- xdata_added_move_infos |>
  select(starts_with("dist_"))

# use apply() to find the last non-NA value in each row
xdata_added_move_infos$final_distance <-
  apply(dist_matrix, 1, function(row) {
  valid_indices <- which(!is.na(row))
  if(length(valid_indices) == 0) return(NA) # in case a row is completely empty (single-move trial)

  last_index <- max(valid_indices)
  return(row[last_index])
})

# histogram of the final move's distance, restricted to trials with more
# than one move (single-move trials have no "final distance" to speak of)
xdata_added_move_infos |>
  filter(final_moves != 1) |>
  ggplot(aes(x = final_distance)) +
  geom_histogram(fill = "grey",
                 color = "black",
                 binwidth = 1
                 ) +
  labs(x = "slider distance between the prefinal and final move (before submitting)") +
  theme_minimal()
```

There are indeed some larger values, up to jumps of 10 and higher. I think we should exclude moves that are beyond a certain threshold. Having said that, it could be that someone explores negative and positive scale positions to get a better feel for them and then jumps to an earlier explored part of the scale, even if that means a larger jump? However, a jump of 8 slider positions corresponds to 40 Hz, corresponding to a jump from a flat tone to our most extreme f0 perturbation. We exclude all trials with a final jump as large, or larger than 40 Hz.

```{r}
#| label: exclude-too-large-final

# before exclusion
control_trials_approved # 4284

# remove all trials with too large final moves
xdata_after_final_moves <- xdata_added_move_infos |>
  filter(final_moves == 1 |
         (final_distance < 8 &
          final_distance > -8))

# data loss?
xdata_with_legit_final_moves <- nrow(xdata_after_final_moves) # 4268
excluded_large_final_moves <- round((((control_trials_approved - xdata_with_legit_final_moves) / control_trials_approved) * 100), 1) # ~0.4% data excluded
```

There were `r excluded_large_final_moves`% of trials in which the final slider change was above the threshold. We exclude these trials.

## Reaction time and amount of moves

Participants were able to make multiple moves before arriving at their final choice. How much did they move?

```{r}
#| label: move-distribtuion
#| fig-cap: "Distribution of amount of moves overall"

xdata_after_final_moves |>
  ggplot(aes(x = final_moves)) +
  geom_histogram(color = "black",
                 fill = "white") +
  labs(x = "amount of moves before submitting the answer",
       y = "") +
  theme_minimal()
```

Mostly 1-2 moves with a heavy right tale. Presumably this changed throughout the experiment.

```{r}
#| label: move-distribution-trial
#| fig-cap: "Distribution of amount of moves across trials"

# reconstruct a single continuous trial index across both experimental
# blocks (the "tone" block and the "cv" block), so trial-level trends can
# be plotted across the whole experiment rather than separately per block
xdata_after_final_moves <-
  xdata_after_final_moves |>
  mutate(block = ifelse(!is.na(tone_trials.thisN), "first", "second"),
         trials = ifelse(block == "first", tone_trials.thisN, cv_trials.thisN + (max(tone_trials.thisN,
                                                                                     na.rm = TRUE) + 1)))
xdata_after_final_moves |>
  group_by(trials) |>
  summarise(mean_move = mean(final_moves, na.rm = TRUE))  |>
  ggplot(aes(x = trials, y = mean_move)) +
  geom_point(pch = 21,
             size = 2,
              color = "white",
              fill = alpha("black", 0.5)) +
  geom_smooth(method = "lm",
              se = FALSE) +
  labs(x = "trial number",
       y = "average amount of moves") +
  theme_minimal()
```

The amount of moves decreases over the experiment as expected. Given that, we expect reaction times to substantially decrease over the course of the experiment. When excluding trials based on reaction times, we will take this into consideration.

```{r}
#| label: rt-distribution-trial
#| fig-cap: "Distribution of RT across trials"

xdata_after_final_moves |>
  group_by(trials) |>
  summarise(mean_rt = mean(response_time, na.rm = TRUE))  |>
  ggplot(aes(x = trials, y = mean_rt)) +
  geom_point(pch = 21,
             size = 2,
              color = "white",
              fill = alpha("black", 0.5)) +
  geom_smooth(method = "lm",
              se = FALSE) +
  labs(x = "trial number",
       y = "average reaction time in seconds") +
  theme_minimal()
```

```{r}
#| label: exclude-outliers

# Log rt
xdata_before_rt_cleaning <- xdata_after_final_moves |> 
  mutate(log_rt = log(response_time))

# Fit a linear mixed-effects model
m <- lmer(log_rt ~ 1 + trials + (1 + trials | participantID), data = xdata_before_rt_cleaning)

# Add model-implied fitted values and raw residuals
xdata_before_rt_cleaning <- xdata_before_rt_cleaning |> 
  mutate(fitted_log_rt = fitted(m),
         resid_log_rt  = log_rt - fitted_log_rt)

# Scale residuals per participant (z-score within each participant)
xdata_before_rt_cleaning <- xdata_before_rt_cleaning |> 
  group_by(participantID) |> 
  mutate(scaled_resid = as.numeric(scale(resid_log_rt))) |>
  ungroup()

# Flag and exclude outliers: |scaled residual| > 4
xdata_before_rt_cleaning <- xdata_before_rt_cleaning |> 
  mutate(outlier = abs(scaled_resid) > 4)

xdata_after_rt_cleaning <- xdata_before_rt_cleaning |>
  filter(!outlier)

# Quick summary of how many trials were excluded, overall and per participant
cat("Total excluded:", sum(xdata_before_rt_cleaning$outlier), "out of", nrow(xdata_before_rt_cleaning), "\n")

xdata_before_rt_cleaning %>%
  group_by(participantID) %>%
  summarise(n_total = n(),
            n_excluded = sum(outlier),
            pct_excluded = 100 * n_excluded / n_total) %>%
  print(n = Inf)

# data loss?
after_rt_cleaning <- nrow(xdata_after_rt_cleaning) # 4256
outlier_rt <- round((((xdata_with_legit_final_moves - after_rt_cleaning) / xdata_with_legit_final_moves) * 100), 1) # 0.3% exclusion
```

To account for outliers due to attentional lapses and distractions, we further removed trials that elicited reaction times that were more than 4 median absolute deviation from the median (`r outlier_rt`%) for each participant for each quarter of the experiment (trials binned into 4 equal bins), respectively.

```{r}
#| label: cleaned_rt-distribution-trial
#| fig-cap: "Distribution of RT across trials after data trimming"

xdata_after_rt_cleaning |>
  group_by(trials) |>
  summarise(mean_rt = mean(response_time, na.rm = TRUE))  |>
  ggplot(aes(x = trials, y = mean_rt)) +
  geom_point(pch = 21,
             size = 2,
              color = "white",
              fill = alpha("black", 0.5)) +
  geom_smooth(method = "lm",
              se = FALSE) +
  labs(x = "trial number",
       y = "average reaction time in seconds") +
  theme_minimal()
```

How much data did we overall lose?

```{r}
#| label: overall-exclusion

exclusions <- round((((all_data - after_rt_cleaning) / all_data) * 100), 1)  # 15.6% exclusion
```

All in all, we excluded `r exclusions`% of data across all exclusion steps combined (attention checks, control-trial performance, oversized final slider moves, and reaction-time trimming).

# Wrangle, store and save for later use

Finally, we select just the columns needed for downstream modeling (demographics, response variables, and predictors), dropping the intermediate helper columns created along the way (e.g., per-trial catch flags, move-sequence columns), and write the cleaned dataset to disk.

```{r}
#| label: select-final-columns

## Store clean dataframe
xdata_wrangled <- xdata_after_rt_cleaning |>
  select(labID, participantID,
         # demographics
         birthyear,sex,
         # origin
         residence_country, country_of_origin,
         # language
         language,
         # music background
         instrument, instrument_training, musical_level,
         musical_training_year, music_hour, group_singing,
         group_singing_year, solo_vocal, solo_vocal_year,
         met_score,
         # response variables
         final_rating, response_f_rel,
         response_f_rel_st,
         stimuli_direction, response_dir,
         response_f, correct,
         recovered_f0,
         # predictors
         delta_t, delta_f,
         delta_f_raw, delta_f_raw_st
         )

```

```{r}
#| label: parse-language_column

# ============================================================
# Step 1: reduce to one row per participant before parsing
# ============================================================
# `language` is repeated identically across every trial row for a given
# participant, so there's no need to parse the same JSON string dozens of
# times - we parse it once per participant instead.
language_raw <- xdata_wrangled |>
  distinct(participantID, language)

# ============================================================
# Step 2: parse each participant's JSON array of languages
# ============================================================
# Each row's `language` field looks like:
#   [{"Column 1":"English","Column 2":1,"Column 3":0,"Column 5":1, ...},
#    {"Column 1":"Polish", ...}]
# - one object per self-reported language. Not every object has every key
# (some rating fields are occasionally missing, and a few entries are
# completely empty {}), so the parser needs to tolerate that gracefully.

parse_language_json <- function(json_string, pid) {
  # handle missing / empty / malformed entries without crashing the pipeline
  if (is.na(json_string) || trimws(json_string) %in% c("", "[]")) {
    return(NULL)
  }
  
  parsed <- tryCatch(
    jsonlite::fromJSON(json_string, flatten = TRUE),
    error = function(e) {
      warning(paste("Could not parse language JSON for participant", pid, ":", e$message))
      NULL
    }
  )
  
  if (is.null(parsed) || length(parsed) == 0) return(NULL)
  
  # fromJSON() can return a data.frame (typical case) or, for a single
  # object with no array wrapper, a named list - normalize to a data.frame
  if (!is.data.frame(parsed)) parsed <- as.data.frame(as.list(parsed), stringsAsFactors = FALSE)
  if (nrow(parsed) == 0) return(NULL)
  
  parsed$participantID <- pid
  parsed
}

language_long <- map2_dfr(
  language_raw$language,
  language_raw$participantID,
  parse_language_json
)

# ============================================================
# Step 3: rename generic "Column N" fields to meaningful names
# ============================================================

language_long <- language_long |>
  rename(any_of(c(
    language_name             = "Column 1",
    overall_fluency           = "Column 2",  
    age_of_acquisition_onset  = "Column 3",
    speaking                  = "Column 5",
    reading                   = "Column 6",
    writing                   = "Column 7",
    listening                 = "Column 8"
  )))

# drop fully-empty rows that came from `{}` entries (no language reported)
language_long <- language_long |>
  filter(!is.na(language_name))

# ============================================================
# Step 4: reshape to one row per participant, wide format
# ============================================================

language_wide <- language_long |>
  group_by(participantID) |>
  mutate(language_number = row_number()) |>
  ungroup() |>
  pivot_wider(
    id_cols     = participantID,
    names_from  = language_number,
    values_from = -c(participantID, language_number),
    names_sep   = "_"
  )

# ============================================================
# Step 5: rename levels
# ============================================================



# ============================================================
# Step 6: join back onto the main dataset as needed
# ============================================================

# Wide format (one row per participant) - good if you just want, e.g.,
# "first/primary reported language" as a single demographic predictor:
xdata_with_languages_wide <- xdata_wrangled |>
  left_join(language_wide, by = "participantID") |> 
  select(-language)


```

```{r}
#| label: write-data

# write clean data after exclusion
write_csv(xdata_with_languages_wide, here::here("data", "derived", "data_after_exclusion.csv"))

```
