---
title: "Descriptive Exploration"
author: Timo Roettger
date: 2026-06-06
format:
  html:
    embed-resources: true
editor: visual
---

```{r}
#| include: FALSE

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

```{r}
#| label: setup
#| include: FALSE

# read in libraries
if (!require("pacman")) install.packages("pacman")
pacman::p_load(viridis, tidyverse, rstudioapi, knitr, kableExtra, broom, patchwork, plotly, htmlwidgets, webshot2, ggrepel, gghighlight)

# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# set the current working directory to the one where this file lives
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

```

```{r}
#| label: data_wrangling

# load in data 2026/06/01
xdata <- read_csv("../data/derived/data_after_exclusion.csv")
hombert <- read_csv("../data/derived/hombert.csv")

```

# Descriptive exploration

We explore how pitch perception changes as a function of the f0 slope (`delta_f`) and the slope duration (`delta_t`). All analyses will work with the semitone transformed pitch responses.

## Overall means

```{r}
#| label: descriptive-exploration
#| fig-cap: "Descriptive means for all combinations of f0 slope in Hz and slope duration in ms"

xdata_agg <- xdata |> 
  filter(delta_t != 0) |> 
  group_by(delta_f_raw, delta_t, participantID) |> 
  summarise(response_f_rel_st = mean(response_f_rel_st, na.rm = TRUE)) |> 
  ungroup()
  
xdata_agg_all <- xdata |> 
  filter(delta_t != 0) |> 
  group_by(delta_f_raw, delta_t) |> 
  summarise(response_f_rel_st = mean(response_f_rel_st, na.rm = TRUE)) |> 
  ungroup() |> 
  mutate(delta_f_raw_ord = as.ordered(delta_f_raw)) |> 
  arrange(delta_t, delta_f_raw_ord)

descriptive_plot <- 
ggplot(data = xdata_agg_all, 
       aes(x = delta_t,
           y = response_f_rel_st,
           color = delta_f_raw_ord,       
           group = delta_f_raw_ord)) +    
  
  geom_line(lwd = 2.5, color = "black") + # Background outline
  geom_line(lwd = 2) +                    # Colored line
  
  scale_y_continuous(limits = c(-3, 3), breaks = seq(-3, 3, 1)) +
  scale_x_continuous(breaks = unique(xdata$delta_t)) +
  labs(x = "\nslope duration in ms",
       y = "average observed pitch response in semitones\n",
       color = "f0 slope in Hz") +
  # Note: You might need to toggle reverse = TRUE/FALSE here 
  # now that the underlying data order is flipped!
  guides(color = guide_legend(reverse = FALSE)) + 
  theme_minimal()
  
descriptive_plot

```

Now we compare these pilot results to the projected results based on Hombert (1975).

```{r}
#| label: hombert-comparison
#| fig-cap: "Comparison between descriptive means of Hombert (1975) and our pilot for all combinations of f0 slope and slope duration."

# rename to match datasets
hombert_renamed <- hombert |> 
  rename(delta_f_raw = target,
         delta_t = slope_dur) |> 
  mutate(dataset = "Hombert (1975)",
         response_f_rel_st = val_st
         ) |> 
  select(delta_f_raw, delta_t, response_f_rel_st, dataset)

xdata_agg_all_2 <- xdata_agg_all |> 
  select(-delta_f_raw_ord) |> 
  mutate(dataset = "Pilot") |> 
  full_join(hombert_renamed) |> 
  mutate(delta_f_raw_ord = as.ordered(delta_f_raw))
  

hombert_comparison <- 
ggplot(data = xdata_agg_all_2, 
       aes(x = delta_t,
           y = response_f_rel_st,
           color = delta_f_raw_ord)) +
  # geom_vline(xintercept = unique(xdata_agg_all$delta_t),
  #            lty = "dashed",
  #            color = "grey") +
  geom_line(aes(group = interaction(delta_f_raw, dataset)), lwd = 2, color = "black") +
  geom_line(aes(group = interaction(delta_f_raw, dataset)), lwd = 1.5) +
  # --- UPDATED: AUTOMATICALLY SPREAD LABELS WITH CONNECTOR LINES ---
 geom_text_repel(
    data = xdata_agg_all_2 |> 
      group_by(dataset, delta_f_raw) |> 
      filter(delta_t == max(delta_t)), # Finds the end of *each individual line*
    aes(label = delta_f_raw_ord),
    
    direction = "y",             # Move labels vertically to avoid overlap
    nudge_x = 25,                # Push labels out to the right
    min.segment.length = 0,      # Always draw connector lines
    segment.color = alpha("grey40",0.2),    
    segment.size = 0.4,          
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_y_continuous(limits = c(-7,7),
                     breaks = seq(-6,6,2)) +
  # Expanded x-axis limits to 300 to give the repelled text and lines room to exist
  scale_x_continuous(breaks = seq(0, 250, 50), limits = c(0, 300)) + 
  labs(title = "Comparison between original study and pilot",
       subtitle = "Hombert (1975, n = 10) vs. Pilot (n = 34)",
       x = "\nslope duration in ms",
       y = "average observed pitch response in semitones\n",
       color = "f0 slope") +
  facet_grid(~dataset) +
  theme_minimal() +
  theme(legend.position = "none") # Hide sidebar legend since lines are directly labeled
  
hombert_comparison

ggsave(filename = "../plots/hombert_comparison.png",
       plot = hombert_comparison,
       device = "png",
       width = 200, 
       height = 120,
       units = "mm",
       dpi = 300)


```

It becomes clear that whatever Hombert (1975) has observed based on 10 American speakers is hardly representative of our pilot results. Looking at only those slope combinations that were used in both Hombert (1975) and our pilot, we can see that the effect magnitudes are substantially smaller:

```{r}
#| label: hombert-comp-table
#| tbl-cap: "Project data from Hombert (1975), projected onto reference value of 150Hz, compared to pilot data"

pilot_tbl <- xdata_agg_all_2 |> 
  filter(delta_f_raw %in% c(-20,-10,10,20),
         delta_t %in% c(40,60,100),
         dataset == "Pilot") |> 
  select(delta_f_raw, delta_t, response_f_rel_st) |> 
  mutate(response_f_rel_st = round(response_f_rel_st,2)) |> 
  pivot_wider(id_cols = delta_t,
              names_from = delta_f_raw,
              values_from = response_f_rel_st)

hombert_tbl <- xdata_agg_all_2 |> 
  filter(delta_f_raw %in% c(-20,-10,10,20),
         delta_t %in% c(40,60,100),
         dataset == "Hombert (1975)") |> 
  select(delta_f_raw, delta_t, response_f_rel_st) |> 
  mutate(response_f_rel_st = round(response_f_rel_st,2)) |> 
  pivot_wider(id_cols = delta_t,
              names_from = delta_f_raw,
              values_from = response_f_rel_st)

knitr::kable(pilot_tbl) |> 
  add_header_above(c(
    "Pilot" = 1,                   # Empty space above the row labels (width = 1 column)
    "f0 perturbation" = 4      # Your variable name spanning across the 2 split columns
  ))

knitr::kable(hombert_tbl) |> 
  add_header_above(c(
    "Hombert" = 1,                   # Empty space above the row labels (width = 1 column)
    "f0 perturbation" = 4      # Your variable name spanning across the 2 split columns
  ))

```

## 3-dimensional

Plotting linear prediction of pilot results in three dimensions:

```{r}
#| label: 3d-plot

xdata_3d <- xdata |> 
  filter(delta_t != 0) |> 
  group_by(delta_f_raw, delta_t) |> 
  summarise(response_f_rel_st = mean(response_f_rel_st, na.rm = TRUE)) |> 
  ungroup()

# estimate a linear model
xmdl <- lm(response_f_rel_st~ delta_t*delta_f_raw, xdata_3d)

# extract predictions
newdata_grid <- expand_grid(
  delta_t = unique(xdata_3d$delta_t),
  delta_f_raw = unique(xdata_3d$delta_f_raw)
)

# extract predictions along with standard errors / fit intervals
raw_predictions <- as_tibble(predict(xmdl))

# bind the predictions back onto your grid mapping for a complete table
final_table <- newdata_grid  |> 
  bind_cols(raw_predictions)

# set up matrix
z_matrix <- matrix(final_table$value, nrow = 8, ncol = 7, byrow = TRUE)

x_axis <- unique(xdata_3d$delta_t)
y_axis <- unique(xdata_3d$delta_f_raw)

x_step <- (max(x_axis) - min(x_axis)) / 20
y_step <- (max(y_axis) - min(y_axis)) / 20

threeD_plot <- 
  plot_ly(x = ~x_axis, y = ~y_axis, z = ~z_matrix, 
        type = "surface",
        #opacity = 0.75,
        colors = 'viridis',
        showscale = FALSE,
        
        contours = list(
          x = list(
            show = TRUE, 
            usecolormap = FALSE, 
            project = list(x = TRUE), 
            color = "rgba(255, 255, 255, 0.4)", # Subtle transparency prevents aliasing lines
            width = 1,                          # Thinner lines look crisper at angles
            start = min(x_axis),
            end = max(x_axis),
            size = x_step                       # Explicitly sets high granularity
          ),
          y = list(
            show = TRUE, 
            usecolormap = FALSE, 
            project = list(y = TRUE), 
            color = "rgba(255, 255, 255, 0.4)", 
            width = 1,                          
            start = min(y_axis),
            end = max(y_axis),
            size = y_step                       # Explicitly sets high granularity
          )
        )) |> 
  layout(
    scene = list(
      xaxis = list(title = 'slope duration (ms)', 
                   tickvals = x_axis, 
                   ticktext = x_axis,
                   showgrid = TRUE),
      yaxis = list(title = 'f0 slope (Hz)', 
                   tickvals = y_axis, 
                   ticktext = y_axis,
                   showgrid = TRUE),
      zaxis = list(title = 'f0 response (ST)')
    ))

threeD_plot

saveWidget(threeD_plot, "../plots/3d_plot.html", selfcontained = TRUE)
#webshot("../plots/3d_plot.html", file = "../plots/3d_plot.png", delay = 2)


```

## Individual differences

Let's have a look at individual differences across the pilot data:

```{r}
#| label: descriptive-exploration-individuals
#| fig-cap: "Descriptive means for all combinations of f0 slope in Hz and slope duration in ms across all listeners. Color indicates f0 slope."

xdata_agg <- xdata |> 
  group_by(delta_f_raw, delta_t, participantID) |> 
  summarise(response_f_rel_st = mean(response_f_rel_st, na.rm = TRUE))

xdata_reference <- xdata |> 
  filter(delta_t == 0) |> 
  group_by(participantID) |> 
  summarise(response_f_rel_st = mean(response_f_rel_st, na.rm = TRUE))

# plot delta_t against response
#subject_plot_1 <- 
ggplot(data = xdata_agg, 
       aes(x = delta_t,
           y = response_f_rel_st,
           color = as.ordered(delta_f_raw))) +
  geom_line(aes(group = delta_f_raw),
            lwd = 1) +
  # scale_y_continuous(limits = c(-20,20),
  #                    breaks = seq(-20,20,10)) +
  scale_x_continuous(breaks = c(20,60,100,140)) +
  #scale_color_viridis_c(option= "viridis") +
  geom_hline(data = xdata_reference,
             aes(yintercept = response_f_rel_st),
             lty = "dashed",
             color = "black") +
  facet_wrap(~participantID, 
             ncol = 6, 
             scales = "fixed") + 
  #scale_color_stepsn(colours = "viridis") +
  labs(x = "\nslope duration in ms",
       y = "average observed pitch response in Hz\n",
       color = "f0 slope") +
    guides(color = guide_legend(reverse = TRUE)) +
  theme_minimal()

```

The majority of listeners do not seem to show much sensitivity to the manipulation. Two notable exceptions are `f2cb` and `r1u7` who show sensitivity to the directional manipulation of f0 slope.

An alternative perspective can be gained from switching the aesthetic mappings:

```{r}
#| label: descriptive-exploration-individuals-2
#| fig-cap: "Descriptive means for all combinations of f0 slope in Hz and slope duration in ms across all listeners. Color now indicates slope duration."

# plot delta_f_raw against response
ggplot(data = xdata_agg, 
       aes(x = delta_f_raw,
           y = response_f_rel_st,
           color = as.ordered(delta_t))) +
  geom_line(aes(group = delta_t),
            lwd = 1) +
  # scale_y_continuous(limits = c(-20,20),
  #                    breaks = seq(-20,20,10)) +
  #scale_color_viridis_c(option= "viridis") +
  geom_hline(yintercept = 0.25,
             lty = "dashed") +
  facet_wrap(~participantID, 
             ncol = 6, 
             scales = "fixed") + 
  #scale_color_stepsn(colours = "viridis") +
  labs(x = "\npitch difference in Hz",
       y = "average observed pitch response in Hz\n",
       color = "slope duration") +
  guides(color = guide_legend(reverse = TRUE)) +
  theme_minimal()

```

To quantify these individual differences, we calculate one metric across all conditions that captures the recovered f0 from the target stimulus. First we explore the proportion of recovered f0 across all f0 perturbations:

```{r}
#| label: recovered-f0
#| fig-cap: "Proportion of recovered f0 across f0 slopes. Grey lines are individual participants; think pink line is population mean. Dashed line at zero indicates null effect; dashed line at 33% indicates predicted ceiling of 1/3 of slope based on Rossi's work."

xdata <- 
  xdata |> 
    mutate(recovered_f0_st = response_f_rel_st / delta_f_raw_st) 

xagg_r <- xdata |> 
  group_by(delta_f_raw, participantID) |> 
  summarise(recovered_f0_st = mean(recovered_f0_st)) |> 
  filter(delta_f_raw != 0)

xagg_r_all <- xdata |> 
  group_by(delta_f_raw) |> 
  summarise(recovered_f0_st = mean(recovered_f0_st)) |>   
  filter(delta_f_raw != 0)

  ggplot(xagg_r,
         aes(x = delta_f_raw,
             y = recovered_f0_st)) +
  geom_line(aes(group = participantID),
            color = "grey",
            alpha = 0.5) +
    geom_line(data = xagg_r_all,
            color = "#dd1c77",
            size = 3) +
    geom_point(data = xagg_r_all,
            size = 3,
            color = "black") +
    geom_point(data = xagg_r_all,
            size = 2,
            color = "#dd1c77") +
    geom_hline(yintercept = c(0, 0.33), 
               lty = "dashed") +
    scale_x_continuous(breaks = c(-40,-30,-20,-10, 10, 20, 30, 40)) +
    scale_y_continuous(breaks = c(-0.8, -0.3, -0.2, -0.1, 0, 0.1, 0.2, 0.3, 0.6),
                       labels = paste0((c(-0.8, -0.3, -0.2, -0.1, 0, 0.1, 0.2, 0.3, 0.6) * 100), "%")) +
    labs(x = "f0 perturbation",
         y = "percentage of recovered f0") +
    theme_minimal()

```

```{r}

ggplot(xagg_r,
         aes(x = delta_f_raw,
             y = recovered_f0_st)) +
  geom_line(aes(group = participantID),
            color = "#dd1c77",
            linewidth = 1) +
  gghighlight(use_direct_label = FALSE,
              unhighlighted_params = list(colour = alpha("grey85", 1), linewidth = 0.3)) +
  geom_hline(yintercept = c(0, 0.33), 
               lty = "dashed") +
  facet_wrap(~participantID, ncol = 6) +
  scale_x_continuous(breaks = c(-40, 0, 40)) +
  scale_y_continuous(breaks = c(-0.8, 0, 0.6),
                       labels = paste0((c(-0.8, 0, 0.6) * 100), "%")) +
    labs(x = "f0 perturbation",
         y = "percentage of recovered f0") +
    theme_minimal()

```

It does not look like as if the amount of f0 perturbation matters for the recovered f0. As opposed to the literature, the average recovered pitch lies between 0 and 10% and not at 1/3 as predicted by Rossi's work.

```{r}

xagg_r_participants <- xdata |> 
  filter(delta_f_raw != 0) |> 
  group_by(participantID, delta_t, delta_f_raw) |> 
  summarise(recovered_f0_st = mean(recovered_f0_st)) |> 
  group_by(participantID) |> 
  summarise(recovered_f0_st = mean(recovered_f0_st))

ggplot(xagg_r_participants,
         aes(x = recovered_f0_st)) +
  geom_point(y = 0,
             size = 3,
             pch = 21,
             color = "white",
             fill = "#dd1c77") +
  geom_density(fill = "#dd1c77",
               alpha = 0.5) +
  scale_x_continuous(limits = c(-0.1,0.4),
                     breaks = seq(-0.1,0.4,0.1),
                     labels = paste0(seq(-0.1,0.4,0.1)*100, "%")) +
  geom_vline(lty = c("dashed", "dotted"),
             xintercept = c(0,1/3)) +
  labs(title = "Individual differences for recovered f0 across all conditions",
       subtitle = "",
       y = "",
       x = "percentage of recovered f0") + 
  theme_minimal() +
  theme(legend.position = "none",
        strip.text = element_blank())

```

Looking at the distribution of participants' average recovered f0, we can see a strong mode between 0 and 10% with three outliers that recover f0 above 20%.
