---
title: "Inferential estimation"
author: Timo Roettger
date: 2026-06-06
format: html
editor: visual
---

```{r}
#| include: FALSE

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

```{r}
#| label: setup

# Read in libraries
if (!require("pacman")) install.packages("pacman")
pacman::p_load(brms, tidybayes, tidyverse, dplyr, truncnorm, bayestestR)

# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# set the current working directory to the one where this file lives
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))


# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

```{r}
#| label: data_wrangling

# load in data 2026/06/01
xdata <- read_csv("../data/derived/data_after_exclusion.csv")
```

# Estimates for simplified model

In Bayesian models, we approximate a posterior distribution, that is, a distribution of plausible parameter values, given the data, the model and any prior assumptions about the estimated parameter values. Because the approximation of a posterior distribution is computationally costly, we use algorithms to sample values from the posterior distribution. The practical consequence of this choice is that we do not calculate a single point estimate for a parameter, but we draw a sample of, in our case, 8,000 plausible values for that parameter and quantify our uncertainty about it by summarizing the distribution of values (see Vasishth et al. 2018, and Franke & Roettger 2019 for tutorials using speech science-specific examples). 

We estimate a single contrast to make estimate a target sample size that allows us to reliably detect the difference we have found in the pilot.

\(1\) $$y_{ij} = \beta_0 + \beta_{1xij} + u_{0j} + u_{1xij} + \epsilon_{ij}$$

```{r}
#| label: model_specs

# target condition for delta_f
target_cond = 10

# prepare reduced data set for model
xdata_red <- xdata |> 
  # only for delta_t = 140 ms
  filter(delta_t  == 140,
         # only for delta_f = -10,10 Hz
         delta_f_raw %in% c(-target_cond,target_cond)) |>
  # make factor
  mutate(delta_f_raw = as.factor(delta_f_raw))

# contrast-code with -10 -> -.5
contrasts(xdata_red$delta_f_raw) <- c(-.5,.5)

# brms formula
# formula_d_f0 = bf(response_f_rel ~ delta_f_raw + (delta_f_raw | participant))

# brms formula for student
formula_d_f0 = bf(response_f_rel ~ delta_f_raw + (delta_f_raw | participantID), nu = 4)

## We used regularizing, weakly informative priors for relevant parameters (Gelman, Simpson, & Betancourt 2017). The width of those regularizing priors was chosen based on existing models on the perception of dynamic pitch changes (Rossi 1971, 1978) suggesting that the perceived pitch corresponds to a time point at ⅔ of the pitch change, corresponding to ⅓ of the Δf0. Thus, for the chosen Δf0s (-10 Hz  vs 10 Hz), we would expect reasonable ceiling performances for them to be -3.33 Hz and -3.33 Hz, respectively, rendering the expected difference between the two conditions at 6.7 Hz.

# my priors 
# NOTE: could even specify 2/3 as upper boundary
mypriors <- c(
              prior(cauchy(0, 3.33), class = "b", ub = 100, lb = -100),
              prior(cauchy(0, 3.33), class = "Intercept", ub = 50, lb = -50),
              prior(cauchy(0, 3.33), class = "sd", ub = 100),
              prior(lkj(1), class = "cor")
              )

# run prior-only model for inspection
simple_model_d_f0_priors <- brm(
  formula_d_f0,
  data = xdata_red,
  file = "../models/model_simple_df0_priors",
  family = student,
  sample_prior = "only",
  seed = 123,
  iter = 4000,
  warmup = 2000,
  prior = mypriors, 
  control = list(adapt_delta = 0.99,
                 max_treedepth = 10)
  )


```

## Prior Predictive Check

The priors look good, all possible values are represented, but priors down weight values much beyond the actual target values.

```{r}
#| label: plot_priors
#| fig-cap: "Prior predictive check"

# plot priors
  simple_model_d_f0_priors |> 
  spread_draws(b_delta_f_raw1) |> 
  ggplot(aes(x = b_delta_f_raw1)) + 
  stat_slab(slab_color = "darkred",
                    slab_fill = alpha("darkred", 0.5),
                    fill = NA,
                    color = NA,
                    outline_bars = FALSE) +
  geom_vline(xintercept = c(-target_cond, target_cond),
             lty = "dashed") +
  labs(x = "\n prior pitch difference between conditions",
       y = "") +
  scale_x_continuous(limits = c(-50,50),
                     breaks = seq(-50,50,25)) +
  theme_minimal() +
  theme(axis.text.y = element_blank())


```

## Run model

```{r}
#| label: run_model

# Run model with generous amount of iterations
simple_model_d_f0 <- brm(
  formula_d_f0,
  data = xdata_red,
  file = "../models/model_simple_df0",
  family = student,
  seed = 123,
  iter = 4000,
  warmup = 2000,
  prior = mypriors, 
  control = list(adapt_delta = 0.9,
                 max_treedepth = 10)
  )

simple_model_d_f0

```

The model estimates the critical effect to be `r round(fixef(simple_model_d_f0)[2,1],1)` \[95% CrI `r round(fixef(simple_model_d_f0)[2,3],1)` - `r round(fixef(simple_model_d_f0)[2,4],1)`\]. Effective sample size looks sufficient, and Rhat values indicate good convergence.

```{r}
#| label: check-diagnostics
#| fig-cap: "Sampling chains mixed well."

# check caterpillar
plot(simple_model_d_f0).

```

Chains also seem to have mixed well.

```{r}
#| label: posterior-predictive-check
#| fig-cap: "Posterior Predictive Check seems to indicate that the likelihood link captures the data well."

# check samples against data
pp_check(simple_model_d_f0, ndraws = 100) # looks great

```

## Plot estimates

```{r}
#| label: plot_model
#| fig-cap: "Estimated effect magnitudes for slope duration of 140 ms, comparing f0 slopes of -10 Hz vs. 10 Hz. Error bars indicate 95% Credible Intervals."

# create new data to predict
newdata_delta_f0 <- expand_grid(delta_f_raw = c(-target_cond,target_cond))

  simple_model_d_f0 |> 
  epred_draws(newdata = newdata_delta_f0,
              re_formula = NA) |> 
    group_by(delta_f_raw) |> 
    summarise(pred_m = mean(.epred, na.rm = TRUE),
              pred_0025 = quantile(.epred, prob = 0.025),
              pred_0975 = quantile(.epred, prob = 0.975)) |> 
  ggplot(aes(x = as.factor(delta_f_raw),
           y = pred_m,
           #group = as.numeric(delta_t),
           color = as.factor(delta_f_raw))) +
  geom_errorbar(aes(ymin = pred_0025,
                    ymax = pred_0975),
                width = 0.1,
                color = "black") +
  geom_point(size = 4,
             color = "black") +
  geom_point(size = 3) +
  scale_color_viridis_d(option="viridis") +
  scale_x_discrete(labels = paste0(c(-10,10), " Hz")) +
  scale_y_continuous(limits = c(-10,10),
                     breaks = seq(-10,10,5),
                     labels = paste0(seq(-10,10, 5), " Hz")) +
  labs(x = "\npresented pitch difference",
       y = "predicted pitch rating\n",
       fill = "delta f0\n") +
  guides(colour = "none") +
  theme_minimal()

```

# Inference using Bayes Factor (through Savage-Dickey)

```{r}
#| label: inference

# bf directional 
hyp1 <- hypothesis(simple_model_d_f0, "delta_f_raw1 > 0")
hyp1$hypothesis[[6]]
# BF = 108.6

```
