---
title: "Inferential estimation"
author: Timo Roettger
date: 2026-06-06
format:
  html:
    embed-resources: true
    toc: true       
editor: visual
---

```{r}
#| include: FALSE

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)
```

```{r}
#| label: setup

# Read in libraries
if (!require("pacman")) install.packages("pacman")
pacman::p_load(brms, here, tidybayes, tidyverse, dplyr)

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

```{r}
#| label: load-data

# load in data 2026/06/01
xdata <- read_csv(here::here("data","derived","data_after_exclusion.csv"))
```

# Estimates for simplified model

In Bayesian models, we approximate a posterior distribution, that is, a distribution of plausible parameter values, given the data, the model and any prior assumptions about the estimated parameter values. Because the approximation of a posterior distribution is computationally costly, we use algorithms to sample values from the posterior distribution. The practical consequence of this choice is that we do not calculate a single point estimate for a parameter, but we draw a sample of, in our case, 8,000 plausible values for that parameter and quantify our uncertainty about it by summarizing the distribution of values (see Vasishth et al. 2018, and Franke & Roettger 2019 for tutorials using speech science-specific examples).

We estimate a single contrast from the pilot data - the difference in perceived pitch between a -10 Hz and a +10 Hz f0 shift at the longest slope duration (140 ms) - which we will later use to inform a target sample size for detecting an effect of this magnitude reliably.

\(1\) $$y_{ij} = \beta_0 + \beta_{1xij} + u_{0j} + u_{1xij} + \epsilon_{ij}$$

We restrict this analysis to a single condition (the longest slope duration, 140 ms) and its two most extreme, symmetric pitch shifts (-10 Hz and +10 Hz), so that the estimated effect reflects the clearest possible test of directional pitch perception before we generalize to the full design in later models.

```{r}
#| label: model_specs

# target condition for delta_f: we restrict the analysis to the +/-10 Hz
# pitch-shift conditions (the two most extreme, symmetric conditions
# available), and to the longest slope duration (140 ms, filtered below)
target_cond = 10

# prepare reduced data set for model
xdata_red <- xdata |>
  # only for delta_t = 140 ms
  filter(delta_t  == 140,
         # only for delta_f = -10,10 Hz
         delta_f_raw %in% c(-target_cond,target_cond)) |>
  # turn the numeric two-level predictor into a factor so we can apply
  # custom (deviation/sum) contrast coding below
  mutate(delta_f_raw = as.factor(delta_f_raw))

# sanity check to confirm the factor's level order is what the next line assumes before applying contrasts.
stopifnot(identical(levels(xdata_red$delta_f_raw), c("-10", "10")))

# contrast-code with -10 -> -.5, 10 -> .5 (deviation/sum coding), so that
# the resulting model coefficient represents the FULL difference between
# the two conditions (rather than, e.g., a difference from a reference
# level under default treatment coding)
contrasts(xdata_red$delta_f_raw) <- c(-.5,.5)


# brms formula for a Student-t distributed outcome (robust to occasional
# outlying responses, at the cost of assuming a fixed tail-heaviness
# rather than estimating it - see `nu = 4` below), with random
# intercepts and random slopes for delta_f_raw by participant.
# We fix dfs of the student distribution to 4, instead of estimating the parameter which would affect the computational feasibility of the sample size simulation
formula_d_f0 = brms::bf(response_f_rel_st ~ delta_f_raw + (delta_f_raw | participantID), nu = 4)


## We used regularizing, weakly informative priors for relevant parameters (Gelman, Simpson, & Betancourt 2017). The width of those regularizing priors was chosen based on existing models on the perception of dynamic pitch changes (Rossi 1971, 1978) suggesting that the perceived pitch corresponds to a time point at ⅔ of the pitch change, corresponding to ⅓ of the Δf0. Thus, for the chosen Δf0s (-10 Hz  vs 10 Hz), we would expect reasonable ceiling performances for them to be 3.33 Hz and -3.33 Hz, respectively, rendering the expected difference between the two conditions at 6.7 Hz. At this reference level, the predicted ceiling difference is 0.76 semitones, thus we pick half of that (= 0.38) as our scaling factor for the priors.
# The slider spans 100 Hz to 200 Hz, that means the maximally possibly difference is abs(12) semitones.

# expected "ceiling" perceived pitch shift under Rossi's 1/3-of-delta_f0
# account, converted to semitones relative to the 150 Hz reference -
# used below purely to annotate the prior-predictive plot, not in the
# priors themselves
ceiling = 12 * log2(((1/3 * 10) + 150)/150)

# the slider's hard physical limits (100-200 Hz), converted to semitones
# relative to the 150 Hz reference - these define the widest range of
# values the outcome variable can possibly take
min_response = 12 * log2(100/150)
max_response = 12 * log2(200/150)

# the largest possible DIFFERENCE between two responses, given the
# slider's range - i.e., one response pinned at the minimum and the other
# at the maximum
max_delta = max_response - min_response

# my priors
mypriors <- c(
              # prior on the fixed effect of delta_f_raw (the contrast of
              # interest): a Cauchy prior centered at 0 (no effect a
              # priori), scaled so most prior mass sits well within a
              # plausible effect size, truncated at the largest
              # difference physically expressible on the slider
              brms::prior(cauchy(0, 0.38),
                    class = "b",
                    lb = -12,
                    ub = 12),
              # prior on the model intercept (i.e., the grand mean
              # response, since -.5/+.5 contrast coding centers the
              # intercept between conditions): bounded by the slider's
              # physical min/max response range
              brms::prior(cauchy(0, 0.38),
                    class = "Intercept",
                    lb = -7.01955,
                    ub = 4.98045),
              # prior on the by-participant random-effect SDs (intercept
              # and slope): same weakly-informative scale, truncated to
              # be non-negative and bounded by the same maximum plausible
              # difference as the fixed effect
              brms::prior(cauchy(0, 0.38),
                    class = "sd",
                    ub = 12),
              # weakly informative prior on the correlation between
              # by-participant random intercepts and slopes
              brms::prior(lkj(1), class = "cor")
              )

# run prior-only model for inspection (samples exclusively from the
# priors, ignoring the data/likelihood entirely - this lets us visualize
# what the priors imply BEFORE looking at how they interact with real data)
simple_model_d_f0_priors <- brm(
  formula_d_f0,
  data = xdata_red,
  file = here::here("models", "model_simple_df0_priors"),
  file_refit = "on_change",
  family = student,
  sample_prior = "only",
  seed = 123,
  iter = 4000,
  warmup = 2000,
  prior = mypriors,
  control = list(adapt_delta = 0.99,
                 max_treedepth = 10)
  )

```

## Prior Predictive Check

The priors look good, all possible values are represented, but priors down weight values much beyond the actual target values.

```{r}
#| label: plot_priors
#| fig-cap: "Prior predictive check"

# visualize the prior distribution implied for the critical contrast
# (the difference between the -10 Hz and +10 Hz conditions), before any
# data has informed it - the dashed lines mark the theoretically expected
# "ceiling" difference under Rossi's account, computed above
  simple_model_d_f0_priors |>
  spread_draws(b_delta_f_raw1) |>
  ggplot(aes(x = b_delta_f_raw1)) +
  stat_slab(slab_color = "darkred",
                    slab_fill = alpha("darkred", 0.5),
                    fill = NA,
                    color = NA,
                    outline_bars = FALSE) +
  geom_vline(xintercept = c(-ceiling, ceiling),
             lty = "dashed") +
  labs(x = "\n prior pitch difference between conditions",
       y = "") +
  scale_x_continuous(limits = c(-12,12),
                     breaks = seq(-12,12,6)) +
  theme_minimal() +
  theme(axis.text.y = element_blank())
```

## Run model

Having confirmed the priors are reasonable (weakly informative, covering the full plausible range without concentrating unrealistic mass near the boundaries), we now fit the full model to the actual pilot data.

```{r}
#| label: run_model
#| include: FALSE

# Run model with generous amount of iterations
simple_model_d_f0 <- brm(
  formula_d_f0,
  data = xdata_red,
  file = here::here("models", "model_simple_df0"),
  file_refit = "on_change",
  family = student,
  seed = 123,
  iter = 8000,
  warmup = 2000,
  prior = mypriors,
  control = list(adapt_delta = 0.99,
                 max_treedepth = 10)
  )

summary(simple_model_d_f0)$fixed
```

The model estimates the critical effect to be `r round(fixef(simple_model_d_f0)[2,1],1)` \[95% CrI `r round(fixef(simple_model_d_f0)[2,3],1)` - `r round(fixef(simple_model_d_f0)[2,4],1)`\]. Effective sample size looks sufficient, and Rhat values indicate good convergence.

```{r}
#| label: check-diagnostics
#| fig-cap: "Sampling chains mixed well."

# visual check: trace plots (are the chains exploring the same region,
# with no drift/trend?) and posterior density plots for each parameter
plot(simple_model_d_f0)
```

Chains also seem to have mixed well.

```{r}
#| label: posterior-predictive-check
#| fig-cap: "Posterior Predictive Check seems to indicate that the likelihood link captures the data well."

# overlay 100 datasets simulated from the fitted model's posterior against
# the real, observed data - if the model is a good fit, the simulated
# (light) distributions should closely track the observed (dark) one
pp_check(simple_model_d_f0, ndraws = 100) # looks great
```

## Plot estimates

The model estimates a `r round(fixef(simple_model_d_f0)[2], 2)` semitone difference between -10 and 10 Hz with a 95% Credible Interval of `r round(fixef(simple_model_d_f0)[2,3], 2)` to `r round(fixef(simple_model_d_f0)[2,4], 2)` semitones.

```{r}
#| label: plot_model
#| fig-cap: "Estimated effect magnitudes for slope duration of 140 ms, comparing f0 slopes of -10 Hz vs. 10 Hz. Error bars indicate 95% Credible Intervals."

# build a small prediction grid: just the two conditions we're comparing
newdata_delta_f0 <- expand_grid(delta_f_raw = c(-target_cond,target_cond))
# NOTE: `delta_f_raw` is supplied here as plain numeric (-10, 10), even
# though the model was fit with `delta_f_raw` as a factor with custom
# contrasts. brms's internal new-data validation coerces new columns to
# match the factor levels used at fit time (via character matching, so
# -10 -> "-10" -> the existing "-10" level), so this works correctly -
# just flagging it since it may look inconsistent to a reader who
# expects new data types to exactly mirror the fitting data's types.

  simple_model_d_f0 |>
  # draw posterior-predicted (population-level) expected values for each
  # of the two conditions; re_formula = NA averages over/ignores
  # participant-level random effects, giving the population-average
  # prediction rather than any specific participant's predicted curve
  epred_draws(newdata = newdata_delta_f0,
              re_formula = NA) |>
    group_by(delta_f_raw) |>
    summarise(pred_m = mean(.epred, na.rm = TRUE),
              pred_0025 = quantile(.epred, prob = 0.025),
              pred_0975 = quantile(.epred, prob = 0.975)) |>
  ggplot(aes(x = as.factor(delta_f_raw),
           y = pred_m,
           color = as.factor(delta_f_raw))) +
  geom_errorbar(aes(ymin = pred_0025,
                    ymax = pred_0975),
                width = 0.1,
                color = "black") +
  geom_point(size = 4,
             color = "black") +
  geom_point(size = 3) +
  scale_color_viridis_d(option="viridis") +
  scale_x_discrete(labels = paste0(c(-10,10), " Hz")) +
  scale_y_continuous(limits = c(-1,1),
                     breaks = seq(-1,1,0.25),
                     labels = paste0(seq(-1.00,1.00,0.25), " STs")) +
  labs(x = "\npresented pitch difference",
       y = "predicted pitch rating in semitones\n",
       fill = "delta f0\n") +
  guides(colour = "none") +
  theme_minimal()
# NOTE: the y-axis is hard-limited to +/-1 semitone. Given the current
# estimated effect this comfortably contains both conditions' credible
# intervals, but if this script is re-run on updated/different data with
# a larger effect or wider uncertainty, points or error bars could be
# silently clipped without any warning. Consider deriving the limits from
# the data (e.g. `range(c(pred_0025, pred_0975)) * 1.1`) if this plot will
# be regenerated across multiple pilots/datasets.
```

## Inference using Bayes Factor (i.e. through evidence ratio)

To complement the credible interval above, we also quantify directional evidence for the effect using `brms::hypothesis()`, which for a one-sided hypothesis like this one reports the ratio of posterior probability mass favoring the hypothesis over its complement. <!-- SUGGESTION (methodological framing, not a code change): technically,
what brms::hypothesis() reports for a *directional* (one-sided, ">"/"<")
hypothesis is a posterior odds ratio derived from the posterior
probability mass on either side of zero - this differs from the classical
Savage-Dickey density-ratio Bayes Factor, which applies to *point-null*
("=") hypotheses. Calling this an "evidence ratio" (as the section header
already does) is precise; calling it a "Bayes Factor" outright (as the
prose below does) is common shorthand in the literature and not wrong
per se, but if this report will be read by reviewers familiar with the
distinction, a one-sentence clarification (e.g. a footnote citing the
brms::hypothesis() documentation) can preempt that question. -->

```{r}
#| label: inference

# test the directional hypothesis that the effect of delta_f_raw is
# positive (i.e., that ratings are higher in the +10 Hz condition than
# the -10 Hz condition, in the direction predicted by accurate pitch
# perception)
hyp1 <- hypothesis(simple_model_d_f0, "delta_f_raw1 > 0")

# extract the evidence ratio by NAME rather than position
BF <- hyp1$hypothesis$Evid.Ratio
# FIX APPLIED: previously `hyp1$hypothesis[[6]]`, which relies on
# "Evid.Ratio" always being exactly the 6th column of the hypothesis()
# output - true in current brms versions, but a positional index like
# this would silently pull the WRONG column (with no error) if a future
# brms update ever adds, removes, or reorders columns in this output.
# Referencing the column by name is equivalent today and safe against
# that kind of silent breakage going forward.
# BF = 157
```

The Bayes Factor, assessing evidence for a positive directional pitch perception is `r round(BF,1)`, suggesting extreme evidence for the alternative hypothesis.
