---
title: "01_sample_size_estimation"
format: html
editor: visual
---

```{r}
#| label: setup
#| include: FALSE

if (!require("pacman")) install.packages("pacman")
pacman::p_load(brms, ggplot2, patchwork, plotly, tidybayes, modelr, viridis, tidyverse, dplyr, truncnorm, lme4, faux, scam, bayestestR, faux)

# Set the current working directory to the one where this file is
current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path)
setwd(current_working_dir)

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Data & Wrangling

Load in data.

```{r}
#| label: data_wrangling

# load in data 2026/06/01
xdata <- read_csv("../data/derived/data_after_exclusion.csv")

```

# Simulating data based on model parameters

Based on the estimated values the difference between Δf0 = -20 ms and -20 ms (#.# Hz \[95% CrI: #.#-#.# Hz\]), alongside the estimated variance components (random intercepts, random slopes, and correlation between them), we generated 100 datasets with these properties for a range of sample sizes \[20-80\]. For each simulation, we fitted a model as specified in (1) to the data and estimated the Bayes Factor for the critical coefficient being larger than point-zero, using the Savage-Dickey density ratio implemented in the bayestestR packages (Makowski et al. 2019 The Savage-Dickey density ratio (e.g. Dickey & Lientz 1970, Wagenmakers et al. 2010) relates the amount of evidence (the density of posterior samples) at point-zero for the model based on the priors only (i.e. before seeing the data) to the amount of evidence for the model based on both priors and likelihood (i.e. after seeing the data) (Liao et al. 2021). The Savage-Dickey method lets us assess evidence for and against a null hypothesis using Bayes Factor (BF) and is particularly suited for nested models, especially when models only differ with respect to one parameter.

```{r}
#| label: parameter_extraction

# NOTE: currently still based on the gaussian models. needs to be updated if we go with the student error terms

# load in model
xmdl <- readRDS("../models/model_simple_df0.rds")

# intercept
intercept = fixef(xmdl)[[1]]

# effect magnitude between conditions
beta = fixef(xmdl)[[2]]

# get random effects
vc <- as.data.frame(VarCorr(xmdl))

# get random intercept sd
ran_int = vc$participantID.sd.Estimate[1]

# get random slope sd
ran_slope = vc$participantID.sd.Estimate[2]

# get cor between intercept and slope
ran_cor = vc$participantID.cor.Estimate.Intercept[2]

# get resiudal sd
sigma = vc$residual__.sd.Estimate[1]

```

```{r}
#| label: wrapper_around_data_sim

my_sim_data <- function(
# participants
subj_n = 50,    # number of subjects per language
rep_n = 2,      # number of repetitions

# predictors
# lang_n = 4   # number of languages
b0 = intercept,        # intercept
b1 = beta,             # fixed effect of delta_f (difference between delta_f = -20 vs. 20) 

# random effects
u0s_sd = ran_int,      # random intercept SD for subjects
u1s_sd = ran_slope,    # random b1 slope SD for subjects (delta_t)
r01s = ran_cor,        # correlation between random effects b0 and b1 for subjects

sigma_sd = sigma) {    # residuals SD

# create data frame
df <-   add_random(subject = subj_n) |> 
        # set up within subject predictors
        add_within("subject", 
                   delta_f = c(-0.5,0.5)) |> 
        add_within("subject", 
                   repetition = seq(1, rep_n, 1)) |> 
        # set up by_subject random intercept and slope for delta_f
        add_ranef("subject", u0s = u0s_sd, u1s = u1s_sd, .cors = r01s) |> 
        # add error term (by observation)
        add_ranef(sigma = sigma_sd) |> 
        # generate responses 
        mutate(response = b0 + u0s + (b1 + u1s) * delta_f + sigma,
               delta_f = as.factor(delta_f))
        # contrast code
        contrasts(df$delta_f) = c(-0.5,0.5)
        
df
}


## sanity check
my_sim_data() |> 
  ggplot() +
  geom_density(aes(x = response, 
                   fill = delta_f),
               alpha = 0.5) 

```

```{r}
#| label: wrapper_around_model

# model formula
formula_sim = bf(response ~ delta_f + (delta_f | subject), nu = 4)

# my priors (default for now plus weakly regularizing for beta)
mypriors <- c(
              prior(cauchy(0, 3.33), class = "b", ub = 100, lb = -100),
              prior(cauchy(0, 3.33), class = "Intercept", ub = 50, lb = -50),
              prior(cauchy(0, 3.33), class = "sd", ub = 100),
              prior(lkj(1), class = "cor")
              )

# number of sims per specification (target = 500?)
n_sims <- 200

# Container for model fits
bfs_list <- list()
subject_list <- list()

# Define subject number for different models (25-75)
subjects <- as.list(seq(25,75,5))

# Loop through priors and fit models
for (i in seq_along(subjects)) {
    # Inner loop over simulations
    for (j in 1:n_sims) {

    dat_sim <- my_sim_data(subj_n = subjects[[i]])
   
    fit <- brm(
    formula = formula_sim,
    data = dat_sim,
    iter = 4000,
    warmup = 2000,
    control = list(adapt_delta = 0.90),
    family = student(),
    prior = mypriors
  )

  # test BF
  #bf <- bayesfactor_parameters(fit, direction = ">")
  #bf <- bayesfactor_pointnull(posterior = fit, null = 0)
  #extract critical bf
  #bf_beta <- exp(bf[2,2])  
    
  hyp1 <- hypothesis(fit, "delta_f1 > 0") 
  bf_beta <- hyp1$hypothesis[[6]]
  
  bfs_list[[paste0("bf_beta", i, "_sim", j)]] <- bf_beta
  subject_list[[paste0("nsubj =", i, "_sim", j)]] <- length(fit$basis$group_levels$subject)
    } 
}


```

# Plot simulations

```{r}
#| label: extract_results

# wrangle loop results
loop_results <- as.data.frame(bfs_list) |> 
  pivot_longer(cols = everything(), names_to = "condition", values_to = "BF") |> 
  separate_wider_delim(condition, "_", names = c(NA, "subject_condition","sim_n"))

subject_list_df <- as.data.frame(subject_list) |> 
  pivot_longer(cols = everything(), names_to = "sim", values_to = "subj_n") 

loop_results$subj_n = subject_list_df$subj_n

# save results
write.csv(loop_results, "../data/BF_sample_size_calculation.csv")
```

```{r plot_results}

loop_results <- read_csv("../data/BF_sample_size_calculation.csv")
  
# plot results
loop_agg <- loop_results |> 
  group_by(subj_n) |> 
  summarise(BF_l05 = quantile(BF, prob = 0.05),
            BF = median(BF)) 
  

# colors for sequential

myPalette <- colorRampPalette(colors = c("#f1eef6","#d4b9da",
                                         "#c994c7","#df65b0",
                                         "#dd1c77","#980043"))

BF_sample_plot <- 
ggplot(data = loop_agg,
       aes(x = subj_n, y = BF,
           fill = BF)) +
  geom_point(data = loop_results,
             alpha = 0.1,
             color = "white",
             pch = 21) +
  #geom_path() +
  geom_segment(aes(x = subj_n-1,
                   xend = subj_n+1,
                   y = BF_l05,
                   yend = BF_l05),
              size = 1,
              color = "black") +
  geom_point(pch = 21,
             size = 5) +
  geom_hline(yintercept = 10,
             lty = "dashed") +
  scale_fill_gradientn(colours = myPalette(100),
                       trans = "log10",
                       breaks = c(0.1,1,10,100,1000,10000),
                       labels = c(0.1,1,10,100,1000,10000)) +
  # scale_fill_viridis(option= "plasma", 
  #                    trans = "log10",
  #                    limits = c(0.01,10000),
  #                    breaks = c(0.01,0.1,1,10,100,1000,10000),
  #                    labels = c(0.01,0.1,1,10,100,1000,10000)) +
  scale_y_continuous(trans = "log10",
                     limits = c(0.1,10000),
                     breaks = c(0.1,1,10,100,1000,10000),
                     labels = c(0.1,1,10,100,1000,10000)
                     ) +
  scale_x_continuous(limits = c(20,80),
                     breaks = seq(25,75,5)
                     ) +
  labs(x = "\nnumber of subjects",
       y = "Bayes Factor for beta > 0 over beta < 0\n",
       fill = "Bayes Factor for beta > 0\n") +
  theme_minimal()

BF_sample_plot

# save
ggsave(filename = "../plots/BF_sample_sim.png", 
       plot = BF_sample_plot,
       width = 200, 
       height = 100,
       units = "mm", 
       bg = "transparent",
       dpi = 300)

```
