---
title: "01_sample_size_estimation"
format:
  html:
    embed-resources: true
    toc: true       
editor: visual
---

```{r}
#| label: setup
#| include: FALSE

if (!require("pacman")) install.packages("pacman")
pacman::p_load(brms, ggplot2, quantreg,here, patchwork, plotly, tidybayes, modelr, viridis, tidyverse, dplyr, truncnorm, lme4, faux, scam, bayestestR, faux, quantreg)

# global seed 
set.seed(211)

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Data & Wrangling

Load in data.

```{r}
#| label: data_wrangling

# load in data 2026/06/01
xdata <- read_csv(here::here("data", "derived", "data_after_exclusion.csv"))

```

# Simulating data based on model parameters

Based on the critical contrast implied by the full pilot model (Q1_mono_model, fitted to all N = 3972 pilot observations; Δf0 = -10 vs +10 Hz at Δt = 140 ms), alongside the implied variance components (random intercepts, random slopes, and correlation between them), we generated 200 datasets with these properties for a range of sample sizes \[25-75\]. To propagate the uncertainty of the pilot estimates into the design analysis, each simulated dataset uses one joint draw from the posterior of these parameters rather than their point estimates. The residual SD is drawn from the posterior of the two-condition model (model_simple_df0) fitted to the two cells of the critical contrast only, so that no homogeneity-of-variance assumption across cells is introduced. For each simulation, we fitted a model as specified in (1) to the data and estimated the Bayes Factor for the critical coefficient being larger than point-zero, using the Savage-Dickey density ratio implemented in the bayestestR packages (Makowski et al. 2019 The Savage-Dickey density ratio (e.g. Dickey & Lientz 1970, Wagenmakers et al. 2010) relates the amount of evidence (the density of posterior samples) at point-zero for the model based on the priors only (i.e. before seeing the data) to the amount of evidence for the model based on both priors and likelihood (i.e. after seeing the data) (Liao et al. 2021). The Savage-Dickey method lets us assess evidence for and against a null hypothesis using Bayes Factor (BF) and is particularly suited for nested models, especially when models only differ with respect to one parameter.

```{r}
#| label: parameter_extraction

# NOTE: currently still based on the gaussian models. needs to be updated if we go with the student error terms

# load in models
# model_simple_df0 (N = 131): source of the intercept and of the residual SD
# (fitted to the two cells of the critical contrast only, so cell-specific)
xmdl <- readRDS(here::here("models", "model_simple_df0.rds"))

# Q1_mono_model (N = 3972): source of the fixed effect and the by-participant
# variance components. In model_simple_df0 these are essentially unidentified
# (random slope SD 95% CrI [0.008, 0.479]; intercept-slope correlation
# [-0.93, +0.96]); the full model pins them down tightly, and its implied
# point values are close to the ones used before (beta 0.235 vs 0.229).
xmdl_full <- readRDS(here::here("models", "Q1_mono_model.rds"))

# intercept
intercept = fixef(xmdl)[[1]]

# residual sd: posterior draws instead of the point estimate
sigma_draws <- as_draws_df(xmdl, variable = "sigma")$sigma

# Derive the implied parameters of the critical contrast (delta_f = -10 vs
# +10 Hz at delta_t = 140 ms) from Q1_mono_model: for each posterior draw,
# simulate hypothetical new participants from the fitted random-effect
# distribution and predict both cells of the contrast. Keeping one row per
# posterior draw (rather than collapsing to point estimates) lets each
# simulated dataset below use one joint draw, so the uncertainty of the pilot
# estimates is propagated into the design analysis.

n_new   <- 1000   # hypothetical new participants per draw
n_draws <- 2000   # posterior draws

f_levels <- sort(unique(xmdl_full$data$delta_f_raw_st_o))
f_lo <- f_levels[4]   # -10 Hz (-1.194 st)
f_hi <- f_levels[6]   # +10 Hz (+1.117 st)

newdat <- expand_grid(
  participantID = paste0("new_", seq_len(n_new)),
  delta_f_raw_st_o = c(f_lo, f_hi),
  delta_t_o = 140
)

draw_ids <- sample(ndraws(xmdl_full), n_draws)

ep <- posterior_epred(
  xmdl_full, newdata = newdat, draw_ids = draw_ids,
  allow_new_levels = TRUE, sample_new_levels = "gaussian"
)

lo <- ep[, newdat$delta_f_raw_st_o == f_lo]
hi <- ep[, newdat$delta_f_raw_st_o == f_hi]

contrast <- hi - lo        # per draw x participant
cellmean <- (hi + lo) / 2

# joint posterior of the simulation parameters (one row per draw)
par_draws <- tibble(
  beta      = rowMeans(contrast),
  ran_int   = apply(cellmean, 1, sd),
  ran_slope = apply(contrast, 1, sd),
  ran_cor   = sapply(seq_len(n_draws), function(i) cor(cellmean[i, ], contrast[i, ])),
  sigma     = sample(sigma_draws, n_draws, replace = TRUE)
)

# posterior medians (used only for the sanity check below)
beta      = median(par_draws$beta)
ran_int   = median(par_draws$ran_int)
ran_slope = median(par_draws$ran_slope)
ran_cor   = median(par_draws$ran_cor)
sigma     = median(par_draws$sigma)

```

```{r}
#| label: wrapper_around_data_sim

my_sim_data <- function(
# participants
subj_n = 50,    # number of subjects per language
rep_n = 2,      # number of repetitions

# predictors
# lang_n = 4   # number of languages
b0 = intercept,        # intercept
b1 = beta,             # fixed effect of delta_f (difference between delta_f = -20 vs. 20) 

# random effects
u0s_sd = ran_int,      # random intercept SD for subjects
u1s_sd = ran_slope,    # random b1 slope SD for subjects (delta_t)
r01s = ran_cor,        # correlation between random effects b0 and b1 for subjects

sigma_sd = sigma) {    # residuals SD

# create data frame
df <-   add_random(subject = subj_n) |> 
        # set up within subject predictors
        add_within("subject", 
                   delta_f = c(-0.5,0.5)) |> 
        add_within("subject", 
                   repetition = seq(1, rep_n, 1)) |> 
        # set up by_subject random intercept and slope for delta_f
        add_ranef("subject", u0s = u0s_sd, u1s = u1s_sd, .cors = r01s) |> 
        # add error term (by observation)
        add_ranef(sigma = sigma_sd) |> 
        # generate responses 
        mutate(response = b0 + u0s + (b1 + u1s) * delta_f + sigma,
               delta_f = as.factor(delta_f))
        # contrast code
        contrasts(df$delta_f) = c(-0.5,0.5)
        
df
}


## sanity check
my_sim_data() |> 
  ggplot() +
  geom_density(aes(x = response, 
                   fill = delta_f),
               alpha = 0.5) 

```

```{r}
#| label: wrapper_around_model

# model formula
formula_sim = brms::bf(response ~ delta_f + (delta_f | subject), nu = 4)

# my priors (default for now plus weakly regularizing for beta)
mypriors <- c(
              brms::prior(cauchy(0, 0.38), 
                    class = "b", 
                    lb = -12,
                    ub = 12),
              brms::prior(cauchy(0, 0.38), 
                    class = "Intercept", 
                    lb = -7.01955, 
                    ub = 4.98045),
              brms::prior(cauchy(0, 0.38), 
                    class = "sd", 
                    ub = 12),
              brms::prior(lkj(1), class = "cor")
              )

# number of sims per specification (target = 500?)
n_sims <- 200

# Container for model fits
bfs_list <- list()
subject_list <- list()

# Define subject number for different models (25-75)
subjects <- as.list(seq(25,75,5))

# Loop through priors and fit models
for (i in seq_along(subjects)) {
    # Inner loop over simulations
    for (j in 1:n_sims) {

    # one joint posterior draw per simulated dataset
    pd <- par_draws[sample(nrow(par_draws), 1), ]

    dat_sim <- my_sim_data(subj_n = subjects[[i]],
                           b1 = pd$beta,
                           u0s_sd = pd$ran_int,
                           u1s_sd = pd$ran_slope,
                           r01s = pd$ran_cor,
                           sigma_sd = pd$sigma)

    fit <- brm(
    formula = formula_sim,
    data = dat_sim,
    iter = 4000,
    warmup = 2000,
    control = list(adapt_delta = 0.90),
    family = student(),
    prior = mypriors
  )

  # test BF
  #bf <- bayesfactor_parameters(fit, direction = ">")
  #bf <- bayesfactor_pointnull(posterior = fit, null = 0)
  #extract critical bf
  #bf_beta <- exp(bf[2,2])  
    
  hyp1 <- hypothesis(fit, "delta_f1 > 0") 
  bf_beta <- hyp1$hypothesis[[6]]
  
  bfs_list[[paste0("bf_beta", i, "_sim", j)]] <- bf_beta
  subject_list[[paste0("nsubj =", i, "_sim", j)]] <- length(fit$basis$group_levels$subject)
    } 
}


```

# Plot simulations

```{r}
#| label: extract_results

# wrangle loop results
loop_results <- as.data.frame(bfs_list) |> 
  pivot_longer(cols = everything(), names_to = "condition", values_to = "BF") |> 
  separate_wider_delim(condition, "_", names = c(NA, "subject_condition","sim_n"))

subject_list_df <- as.data.frame(subject_list) |> 
  pivot_longer(cols = everything(), names_to = "sim", values_to = "subj_n") 

loop_results$subj_n = subject_list_df$subj_n

# save results
write.csv(loop_results, here::here("data", "derived", "BF_sample_size_calculation.csv"))

```

```{r plot_results}

loop_results <- read_csv(here::here("data", "derived", "BF_sample_size_calculation.csv")) |> 
  mutate(BF = ifelse(is.infinite(BF), 8000, BF))

# fit models for the 5th percentiles on the original scale
# (Or on log(BF) if you prefer, but rq can handle the raw data directly!)
# the lower 5% percentile implements the report's criterion: the sample size
# for which 95% of the simulations reach a Bayes Factor > 10
model_q05 <- rq(BF ~ subj_n, data = loop_results, tau = 0.05)
model_med <- rq(BF ~ subj_n, data = loop_results, tau = 0.50) # Empirical Median

# generate predictions across your 25, 75 subject range
new_data <- data.frame(subj_n = seq(25, 75, by = 1))

empirical_bounds <- data.frame(
  subj_n    = new_data$subj_n,
  BF        = predict(model_med, newdata = new_data),
  BF_l05    = predict(model_q05, newdata = new_data)
)

# ~60 participants: in our runs, n = 50 reached BF > 10 in 92% of simulations,
# n = 60 in ~95% (i.e. recruit ~72 per language under the 20% exclusion buffer)

# colors for sequential
myPalette <- colorRampPalette(colors = c("#f1eef6","#d4b9da",
                                         "#c994c7","#df65b0",
                                         "#dd1c77","#980043"))

BF_sample_plot <-
ggplot(data = empirical_bounds,
       aes(x = subj_n, y = BF)) +
  geom_point(data = loop_results,
             alpha = 0.1,
             fill = "#dd1c77",
             color = "white",
             pch = 21) +
  geom_line(size = 2,
            color = "black") +
  geom_line(size = 1,
            color = "#dd1c77") +
  geom_line(aes(y = BF_l05),
            size = .7,
            color = "black") +
  geom_hline(yintercept = 10,
             lty = "dashed") +
  scale_y_continuous(trans = "log10",
                     breaks = c(0.1, 1, 10, 100, 1000, 8000),
                     labels = c(0.1, 1, 10, 100, 1000, 8000)) +
  scale_x_continuous(limits = c(20, 90),
                     breaks = seq(25, 75, 5)) +
  labs(x = "\nnumber of subjects",
       y = "Bayes Factor for beta > 0 over beta < 0\n") +
  theme_minimal() +
  theme(
    legend.position = "none",
    # Crucial step: add right margin space so the labels fit inside the saved file
    plot.margin = margin(t = 5.5, r = 40, b = 5.5, l = 5.5, unit = "pt") 
  ) +
  # Prevents R from clipping/hiding text placed outside the x-axis limits
  coord_cartesian(clip = "off") +
  
  # --- DIRECT LABELS ---
  # Label for the main BF line (placed at the far right end of the line)
  geom_text(data = tail(empirical_bounds, 1), 
            aes(x = subj_n + 1.5, y = BF, label = "est. BF"), 
            hjust = 0, 
            color = "#dd1c77", 
            fontface = "bold") +
           
  # Label for the BF_l05 line (using the last row of your data)
  geom_text(data = tail(empirical_bounds, 1), 
            aes(x = subj_n + 1.5, y = BF_l05, label = "lower 5% perc."), 
            hjust = 0, 
            color = "black", 
            fontface = "bold")

BF_sample_plot

# save
ggsave(filename = here::here("plots","BF_sample_sim.png"), 
       plot = BF_sample_plot,
       width = 150, 
       height = 100,
       units = "mm", 
       bg = "transparent",
       dpi = 300)

```
