---
title: "06_language_variance_design_analysis"
format:
  html:
    embed-resources: true
    toc: true
editor: visual
---

```{r}
#| label: setup
#| include: FALSE

if (!require("pacman")) install.packages("pacman")
pacman::p_load(brms, here, tidyverse)

# global seed
set.seed(211)

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# A smallest effect size of interest (SESOI) for RQ2.1

When it comes to the critical question of whether there is meaningful language variation, the LOOIC-based model comparison provides no criterion for how large a between-language difference would need to be considered practically meaningful. We anchor the a smallest effect size of interest we deem theoretically interesting (SESOI; Lakens et al. 2018) in the measurement precision of the task itself: the median within-participant standard deviation (SD) of responses to the flat control trials (Δ𝑓 = 0), i.e. the noise with which a single listener reproduces a target that requires no perceptual repair. Matching-response SD is the standard precision measure for matching tasks (Hartmann, Rakerd & Packard, 1985). The SESOI is defined directly on this SD scale, comparing like with like: language populations should systematically differ more than a single listener differs within themselves. We used simulation to evaluate how well the study's planned design (minimally 20 languages, 60 participants each, see participant sample size justification below) is capable of detecting such a difference if it exists.

Note that the exclusion pipeline in `00_data_cleaning.qmd` screens participants partly on their control-trial behavior, so computing the floor on post-exclusion data trims it a slightly stricter (smaller) SESOI, i.e. the conservative direction — and describes the analyzable population that the Stage 2 models will actually see.

```{r}
#| label: sesoi

# load in data
xdata <- read_csv(here::here("data", "derived", "data_after_exclusion.csv"))

# median within-participant matching SD on flat control trials (Hz)
flat_sd <- xdata |>
  filter(delta_f_raw == 0) |>
  group_by(participantID) |>
  summarise(sd_hz = sd(response_f_rel))

delta_star_hz <- median(flat_sd$sd_hz)

# convert to semitones at the 150 Hz reference
delta_star <- 12 * log2((150 + delta_star_hz) / 150)

# the SESOI applies directly on the SD scale: the between-language SD tau is
# compared against the within-participant matching SD
tau_star <- delta_star

delta_star_hz   # ~2.4 Hz
tau_star        # ~0.28 st

```

# Design analysis

With a SESOI in place we can ask: does the planned design distinguish practically relevant from practically negligible language variation? We run this at the planned design — 20 core languages, with 60 participants — rather than solving for a required sample size, because the binding constraint on estimating the between-language SD τ is the number of languages: 20 language means are the 20 observations from which τ is estimated, and this number is fixed by the minimal sample plan of the project. The per-language sample size only shrinks the sampling error of each language mean, with rapidly diminishing returns beyond \~60. This analysis targets the aggregate between-language variation τ: it is not designed to identify which specific languages differ, nor to test grouped contrasts such as tonal vs. non-tonal language populations, which remain part of the exploratory analyses.

For the critical contrast (Δf0 = -10 vs +10 Hz at Δt = 140 ms, 2 repetitions per condition), each language's mean contrast has sampling variance (τ²_participant + σ²) / 50, so the language level reduces to a 20-point meta-analytic problem: a hierarchical model of 20 language means with known sampling error. This keeps each simulated study cheap (a 20-row brms fit).

Participant-level parameters are drawn per simulation from the pilot posteriors — the fixed effect and by-participant contrast SD implied by `models/Q1_mono_model` (N = 3972), the residual SD from `models/model_simple_df0` — so pilot uncertainty is propagated throughout.

```{r}
#| label: implied_parameters

# load in models
xmdl_full <- readRDS(here::here("models", "Q1_mono_model.rds"))
xmdl      <- readRDS(here::here("models", "model_simple_df0.rds"))

# residual sd: posterior draws
sigma_draws <- as_draws_df(xmdl, variable = "sigma")$sigma

# implied fixed effect and by-participant SD of the critical contrast:
# per posterior draw, simulate hypothetical new participants from the fitted
# random-effect distribution and predict both cells of the contrast
n_new   <- 1000
n_draws <- 2000

f_levels <- sort(unique(xmdl_full$data$delta_f_raw_st_o))
f_lo <- f_levels[4]   # -10 Hz
f_hi <- f_levels[6]   # +10 Hz

newdat <- expand_grid(
  participantID = paste0("new_", seq_len(n_new)),
  delta_f_raw_st_o = c(f_lo, f_hi),
  delta_t_o = 140
)

draw_ids <- sample(ndraws(xmdl_full), n_draws)

ep <- posterior_epred(
  xmdl_full, newdata = newdat, draw_ids = draw_ids,
  allow_new_levels = TRUE, sample_new_levels = "gaussian"
)

contrast <- ep[, newdat$delta_f_raw_st_o == f_hi] -
            ep[, newdat$delta_f_raw_st_o == f_lo]

par_draws <- tibble(
  beta   = rowMeans(contrast),
  tau_b1 = apply(contrast, 1, sd)
)

```

Per simulated study we draw 20 true language means (SD = the true between-language τ), add their sampling error, fit the meta-analytic model, and record the posterior of between-language variation. Verdicts follow the posterior probability relative to the SESOI τ\*, using the same 95% standard as the sample size estimation:

- *practically relevant*: P(τ \> τ\*) ≥ 0.95
- *practically negligible*: P(τ \< τ\*) ≥ 0.95
- otherwise inconclusive.

True τ is varied over 0, τ\*/2, τ\*, 1.5 τ\*, and 2 τ\* (boundary cases at τ\* should come out inconclusive).

```{r}
#| label: design_simulation

# true between-language variation
tau_grid <- c(0, tau_star / 2, tau_star, 1.5 * tau_star, 2 * tau_star)

# planned design
n_lang <- 20

# analyzable participants per language: 60 as recommended 
n_subj_grid <- 60

# number of sims per specification
n_sims <- 200

# meta-analytic model: language means with known sampling error
formula_lang = brms::bf(contrast | se(se_j) ~ 1 + (1 | language))

# my priors (same widths as in 03_sample_size_estimation)
mypriors <- c(
              brms::prior(cauchy(0, 0.38),
                    class = "Intercept"),
              brms::prior(cauchy(0, 0.38),
                    class = "sd")
              )

# Container for results
tau_list <- list()

# Loop through sample sizes, true language variation, and fit models
for (h in seq_along(n_subj_grid)) {
  for (i in seq_along(tau_grid)) {
    # Inner loop over simulations
    for (j in 1:n_sims) {

    # one joint posterior draw per simulated dataset
    k <- sample(nrow(par_draws), 1)
    s <- sample(sigma_draws, 1)

    # sampling error of a language mean with n_subj participants
    se_j <- sqrt((par_draws$tau_b1[k]^2 + s^2) / n_subj_grid[h])

    # simulate language means
    dat_sim <- data.frame(
      language = factor(seq_len(n_lang)),
      contrast = par_draws$beta[k] + rnorm(n_lang, 0, tau_grid[i]) + rnorm(n_lang, 0, se_j),
      se_j     = se_j
    )

    if (h == 1 & i == 1 & j == 1) {
      fit0 <- brm(
        formula = formula_lang,
        data = dat_sim,
        iter = 4000,
        warmup = 2000,
        control = list(adapt_delta = 0.90),
        prior = mypriors
      )
      fit <- fit0
    } else {
      # same model, new data; recompile = FALSE reuses the compiled program
      fit <- update(fit0, newdata = dat_sim, recompile = FALSE)
    }

    # posterior of the between-language SD
    tau_post <- as_draws_df(fit, variable = "sd_language__Intercept")$sd_language__Intercept

    tau_list[[paste0("nsubj", h, "_tau", i, "_sim", j)]] <-
      data.frame(n_subj    = n_subj_grid[h],
                 tau_true  = tau_grid[i],
                 sim_n     = j,
                 tau_med   = median(tau_post),
                 p_gt_star = mean(tau_post > tau_star),
                 p_lt_star = mean(tau_post < tau_star))
    }
  }
}

```

# Results

```{r}
#| label: extract_results

# wrangle loop results
loop_results <- bind_rows(tau_list)

# save results
write.csv(loop_results, here::here("data", "derived", "language_variance_analysis.csv"))

```

```{r}
#| label: design_results
#| tbl-cap: "Verdict rates per true between-language tau (proportion of simulated studies)."

loop_results <- read_csv(here::here("data", "derived", "language_variance_analysis.csv"))

loop_results |>
  group_by(n_subj, tau_true) |>
  summarise(
    relevant     = mean(p_gt_star >= 0.95),
    negligible   = mean(p_lt_star >= 0.95),
    inconclusive = mean(p_gt_star < 0.95 & p_lt_star < 0.95),
    tau_med      = median(tau_med)
  )

```

In our run: when the true language variation is zero, 100% of simulated studies affirm *practically negligible* effects, and 99% still do at half the SESOI; at exactly τ\*, \~90% are inconclusive, as a boundary case should be; at 1.5 τ\*, 75% detect *practically relevant* variation, rising to 100% at 2 τ\*. Median estimated τ tracks the true value closely at all levels, i.e. estimation is essentially unbiased. With the SESOI in place, the already-planned design therefore supports an interpretable answer to RQ2.1 in both directions: it reliably certifies independence for true variation at or below half the SESOI, reliably detects dependence at or above 2 τ\*, and is explicitly agnostic in between.
