---
title: "Random Forests"
author: Timo Roettger
date: 2026-06-06
format: html
editor: visual
---

```{r}
#| include: FALSE

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

```{r}
#| label: setup
#| include: FALSE

# install dev version of SAEforest
# devtools::install_github("krennpa/SAEforest")

# install.packages("devtools")
#devtools::install_github("ddebeer/permimp", force = TRUE)

# read in libraries
if (!require("pacman")) install.packages("pacman")
pacman::p_load(viridis, tidyverse, rstudioapi, knitr, kableExtra, broom, patchwork, plotly, ranger, party, SAEforest, perimp, pkgbuild)

# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# set the current working directory to the one where this file lives
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

```

```{r}
#| label: prepare-data

# load in data 2026/06/01
data <- read_csv("../data/derived/data_after_exclusion.csv") 
data$delta_f_raw_st_o <- round(data$delta_f_raw_st, 3) 

predictions <- read_csv("../data/derived/participant_estimated_forRF.csv")

# left_join with other infos
xdata <- 
  data |> 
  filter(delta_f_raw_st != 0) |> 
  distinct(participantID, .keep_all = TRUE) |> 
  left_join(predictions, by = c("participantID")) 


```

```{r}
#| label: prepare-for-rf

# predictors
demographic_vars <- c("birthyear", "sex", "instrument", "instrument_training", "musical_level",
                      "musical_training_year", "music_hour", "group_singing", "group_singing_year", 
                      "solo_vocal", "solo_vocal_year")

# model formula
model_formula <- as.formula(
  paste("P_prop ~", paste(demographic_vars, collapse = " + "))
)

model_data <- xdata[, c("P_prop", demographic_vars)]   # keep only relevant columns
model_data <- model_data |> 
  # cforest doesn't handle NAs internally
  mutate(instrument_training = ifelse(is.na(instrument_training), FALSE, instrument_training),
         musical_level = ifelse(is.na(musical_level), "no information", musical_level),
         musical_training_year = ifelse(is.na(musical_training_year), 0, musical_training_year),
         group_singing_year = ifelse(is.na(group_singing_year), 0, group_singing_year),
         solo_vocal_year = ifelse(is.na(solo_vocal_year), 0, solo_vocal_year)
         )


model_data <- model_data  |> 
  mutate(across(where(is.character), as.factor))  |> 
  mutate(across(where(is.logical), as.factor)) # cforest cannot handle character or logical vector

```

## Tuning the RF

First, map out the parameter combinations you want to test.

```{r}
#| label: define-grid-for-tuning

tuning_grid <- expand.grid(
  mtry      = 1:length(demographic_vars),
  minbucket = c(5, 10, 20),
  maxdepth  = c(0, 5, 10)   # 0 = unlimited depth in party::cforest_control
)

```

Loop through your grid, fit a ranger model for each combination, and extract the Out-of-Bag Mean Squared Error ($OOB\_MSE$) from the forest component to evaluate performance.

```{r}
#| label: run-grid-search

ntree_tuning <- 500

tune_results <- vector("list", nrow(tuning_grid))

set.seed(123)

for (i in seq_len(nrow(tuning_grid))) {
  
  params <- tuning_grid[i, ]
  
  cf_tune <- cforest(
    model_formula,
    data = model_data,
    controls = cforest_unbiased(
      ntree     = ntree_tuning,
      mtry      = params$mtry,
      minbucket = params$minbucket,
      maxdepth  = params$maxdepth
    )
  )
  
  oob_preds_tune <- predict(cf_tune, OOB = TRUE)
  mse_tune <- mean((model_data$P_prop - oob_preds_tune)^2)
  
  tune_results[[i]] <- data.frame(
    mtry      = params$mtry,
    minbucket = params$minbucket,
    maxdepth  = params$maxdepth,
    oob_mse   = mse_tune
  )
  
  cat("Combination", i, "of", nrow(tuning_grid),
      "| mtry =", params$mtry,
      "| minbucket =", params$minbucket,
      "| maxdepth =", params$maxdepth,
      "| OOB MSE =", round(mse_tune, 5), "\n")
}

tuning_df <- bind_rows(tune_results) |> 
  arrange(oob_mse)
print(head(tuning_df, 10))

# Select best combination
best_params <- tuning_df[1, ]
cat("\nBest hyperparameters selected:\n")
print(best_params)


```

## Optimal Model

Once the loop finishes, look at your grid to find the row with the lowest error score and use those exact parameters to train your final model.

```{r}
#| label: identify-optimal-model

set.seed(123)

cf_model <- cforest(
  model_formula,
  data = model_data,
  controls = cforest_unbiased(
    ntree     = 1000,   # bump up ntree for the final model for stability
    mtry      = best_params$mtry,
    minbucket = best_params$minbucket,
    maxdepth  = best_params$maxdepth
  )
)

oob_preds <- predict(cf_model, OOB = TRUE)

mse_oob <- mean((model_data$P_prop - oob_preds)^2)
r2_oob  <- 1 - (sum((model_data$P_prop - oob_preds)^2) /
                 sum((model_data$P_prop - mean(model_data$P_prop))^2))

cat("Final model - Out-of-Bag MSE:", round(mse_oob, 4), "\n")
cat("Final model - Out-of-Bag R2:", round(r2_oob, 4), "\n")

```

## Predictor importance

```{r}
set.seed(123)

cond_importance <- permimp(
  cf_model,
  conditional = TRUE,
  threshold = 0.2,
  asParty = TRUE,      # use the original party-style algorithm instead
  progressBar = TRUE
)

importance_df <- data.frame(
  predictor  = names(cond_importance$values),
  importance = as.numeric(cond_importance$values)
) %>%
  arrange(desc(importance))

print(importance_df)

# Plot
ggplot(importance_df, 
       aes(x = reorder(predictor, importance), y = importance,
       xend = reorder(predictor, importance))) +
  geom_hline(yintercept = 0, 
            lty = "dashed") +
  geom_segment(
    aes(xend = reorder(predictor, importance), y = 0, yend = importance),
    color = "grey70",
    linewidth = 0.8
  ) +
  geom_point(color = "#dd1c77", size = 2) +

  coord_flip() +  # Flips to horizontal for readable variable names
  labs(
    title = "Variable Importance (Fixed Effects Component)",
    subtitle = "Extracted from MERFranger Object",
    x = "",
    y = "Importance Score (Permutation)"
  ) +
  theme_minimal()

```

## Partial Dependence

```{r}

top_n <- 5
top_predictors <- importance_df$predictor[1:top_n]

pred_fun <- function(object, newdata) {
  predict(object, newdata = newdata, OOB = FALSE)
}

pdp_list <- lapply(top_predictors, function(var) {
  pd <- pdp::partial(
    cf_model,
    pred.var  = var,
    pred.fun  = pred_fun,
    train     = model_data,
    grid.resolution = 10,
    ice = FALSE
  )
  names(pd)[1] <- "x_value"
  pd$predictor_name <- var
  pd
})

combined_tbl <- map_dfr(pdp_list, function(pd) {
  pd |> 
    as_tibble() |> 
    mutate(x_value = as.character(x_value))})  |> 
  group_by(predictor_name, x_value)  |> 
  summarise(yhat = mean(yhat, na.rm = TRUE), .groups = "drop")  |> 
  mutate(x_numeric = suppressWarnings(round(as.numeric(x_value), 2)))

# separate factors from numeric to better plot
combined_tbl_numeric  <- 
  combined_tbl |> 
  filter(predictor_name %in% c("group_singing_year", 
                              "music_hour", 
                              "musical_training_year"))
combined_tbl_factors <- 
  combined_tbl |> 
  filter(predictor_name %in% c("instrument", 
                              "instrument_training"))

# plot
combined_tbl_numeric |> 
  ggplot(aes(x =  x_numeric, 
             y = yhat*100)) + 
  geom_point() + 
  geom_line() +
  facet_wrap(~predictor_name,
             scales = "free_x") +
  labs(x = "",
      y = "recovered pitch proportion") +
  theme_minimal()

combined_tbl_factors |> 
  ggplot(aes(x =  x_value, 
             y = yhat*100,
             group = 1)) + 
  geom_point() + 
  geom_line() +
  facet_wrap(~predictor_name,
             scales = "free_x") +
    labs(x = "",
      y = "recovered pitch proportion") +
  theme_minimal()




```


