---
title: "Random Forests"
author: Timo Roettger
date: 2026-06-06
format: html
editor: visual
---

```{r}
#| include: FALSE

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

```{r}
#| label: setup
#| include: FALSE

# install dev version of SAEforest
# devtools::install_github("krennpa/SAEforest")

# install.packages("devtools")
#devtools::install_github("ddebeer/permimp", force = TRUE)

# read in libraries
if (!require("pacman")) install.packages("pacman")
pacman::p_load(viridis, tidyverse, here, knitr, kableExtra, broom, patchwork, plotly, ranger, party, SAEforest, permimp, pkgbuild)

```

# Data Preparation

## Read Data

```{r}
#| label: prepare-data

# latest pilot from 2026/06/01
data <- read.csv(here("data", "derived", "data_after_exclusion.csv"))
predictions <- read_csv(here("data", "derived", "participant_estimated_forRF.csv"))
tone <- read.csv(here("data", "derived", "language_tone.csv"), header = TRUE) |> 
  rename(language_name = name) |> 
  mutate(language_name = tolower(language_name))
ting <- read_csv(here("data", "derived", "language_cf0_effects.csv"))


# left_join with other infos
xdata <- 
  data |> 
  filter(delta_f_raw_st != 0) |> 
  distinct(participantID, .keep_all = TRUE) |> 
  left_join(predictions, by = c("participantID")) 

```

## Wrangle language information

```{r}
#| label: prep-wrangle-demo

# We have to do some wrangling
# --- 0. Fix language strings
unique(xdata$language_name_1)
unique(xdata$language_name_2)
unique(xdata$language_name_3)

# Remove special characters and make everything lower case
# Identify columns starting with 'language_name_'
lang_cols <- grep("^language_name_", colnames(xdata), value = TRUE)

# Loop through each column to convert to lower and remove special characters
for (col in lang_cols) {
  # Convert to lowercase
  xdata[[col]] <- tolower(xdata[[col]])
  
  # Trim spaces from start and end only
  xdata[[col]] <- trimws(xdata[[col]])
  
  # Remove special characters (keeps letters, numbers, and spaces)
  xdata[[col]] <- gsub("[^[:alnum:] ]", "", xdata[[col]])
  
  # Clean up double spaces left behind by removed symbols
  xdata[[col]] <- gsub("\\s+", " ", xdata[[col]])
}

unique(xdata$language_name_1)
unique(xdata$language_name_2)
unique(xdata$language_name_3)

# bin same languages
xdata <- xdata|> 
  mutate(across(starts_with("language_name_"), 
                ~ str_replace(., "american english", "english")))

```

```{r}
#| label: wrangle-demo

# Now creating categories
# Add a row id if you don't already have a unique participant identifier
xdata <- xdata |> mutate(.row_id = row_number())

# --- 1. Reshape to long format: one row per participant per language slot ----
long_data <- xdata |>
  select(.row_id, matches("^(language_name|overall_fluency|age_of_acquisition_onset)_\\d+$")) |>
  pivot_longer(
    cols = -.row_id,
    names_to = c(".value", "slot"),
    names_pattern = "^(language_name|overall_fluency|age_of_acquisition_onset)_(\\d+)$"
  ) |>
  mutate(slot = as.integer(slot)) |>
  filter(!is.na(language_name) & language_name != "") |>  # drop empty/unfilled slots
  mutate(age_of_acquisition_onset = ifelse(slot == 1 & is.na(age_of_acquisition_onset), 0, age_of_acquisition_onset))  # missing AoA -> 0

# --- 2. Identify L1 per participant ------------------------------------------
# Candidate L1s: age_of_acquisition_onset <= 5
# Among candidates, pick lowest overall_fluency (most native);
# tie-break by lowest slot number (i.e. named first)
l1_data <- long_data |>
  filter(age_of_acquisition_onset <= 5) |>
  group_by(.row_id) |>
  arrange(overall_fluency, slot, .by_group = TRUE) |>
  slice(1) |>
  ungroup() |>
  select(.row_id, L1 = language_name)

# Fail-safe: for participants with NO language meeting the age <= 5 criterion,
# fall back to the language in the first slot (lowest slot number)
l1_fallback <- long_data |>
  filter(!(.row_id %in% l1_data$.row_id)) |>   # only participants missing an L1 so far
  group_by(.row_id) |>
  arrange(slot, .by_group = TRUE) |>
  slice(1) |>
  ungroup() |>
  select(.row_id, L1 = language_name)

# Combine primary L1 assignments with fallback assignments
l1_data <- bind_rows(l1_data, l1_fallback)

# Merge L1 back into main dataset
xdata <- xdata |>
  left_join(l1_data, by = ".row_id")

# --- 3. Map L1 onto tonal_system from the `tone` lookup table ----------------
xdata <- xdata |>
  left_join(tone |> select(language_name, description) |> rename(L1_tonal_system = description),
            by = c("L1" = "language_name"))

# --- 4. Classify non-L1 languages: only non-tonal / at least simple / at least complex ----
# Get tonal_system for every language slot (not just L1), then exclude L1 itself
non_l1_data <- long_data |>
  left_join(xdata |> select(.row_id, L1), by = ".row_id") |>
  filter(language_name != L1) |>                      # drop the L1 slot
  left_join(tone |> select(language_name, description), by = "language_name")

# Summarize per participant: what's the "highest" tonal complexity among non-L1 languages?
non_l1_summary <- non_l1_data |>
  group_by(.row_id) |>
  summarise(
    has_complex    = any(description == "Comnplex tone system", na.rm = TRUE),
    has_simple     = any(description == "Simple tone system", na.rm = TRUE),
    n_non_l1_langs = n(),
    .groups = "drop"
  ) |>
  mutate(
    is_monolingual = n_non_l1_langs == 0,
    non_L1_tonal_system = case_when(
      is_monolingual ~ "monolingual",
      has_complex    ~ "at least one complex",
      has_simple     ~ "at least one simple",
      TRUE           ~ "only non-tonal"
    )
  ) |>
  select(.row_id, is_monolingual, non_L1_tonal_system)

# --- 5. Merge everything back into the main dataset --------------------------
xdata <- xdata |>
  left_join(non_l1_summary, by = ".row_id") |>
  mutate(
    is_monolingual = ifelse(is.na(is_monolingual) & !is.na(L1), TRUE, is_monolingual),
    non_L1_tonal_system = ifelse(is_monolingual & is.na(non_L1_tonal_system), "monolingual", non_L1_tonal_system)
  ) |>
  select(-.row_id)

```

## Combine with Ting et al.s productions averages

```{r}
#| label: add-ting-et-al

ting <- ting |> 
  rename(L1 = language)

# left_join with other infos
xdata <- 
  xdata |> 
  left_join(ting, by = c("L1")) 

```

## Wrangle Musical Experiences

```{r}
#| label: wrangle-music

# if group_singing / solo_vocal false, store as 0 else store years
xdata <- xdata |>
  mutate(group_singing_merged = ifelse(group_singing == FALSE, 0,
                                       group_singing_year),
         solo_vocal_merged = ifelse(solo_vocal == FALSE, 0,
                                    solo_vocal_year),
         instrument_training = ifelse(is.na(instrument_training), FALSE, instrument_training),
         musical_level = ifelse(is.na(musical_level), "beginner", musical_level),
         musical_training_year = ifelse(is.na(musical_training_year), 0, musical_training_year)
         )

```

# Random Forests

## Prepare for RF

```{r}
#| label: prepare-for-rf

# predictors
demographic_vars <- c(# general demographics
                      "birthyear", "sex", 
                      # linguistics
                      "L1", "L1_tonal_system",
                      "non_L1_tonal_system",
                      # production
                      "effect_size_est", "effect_length_est", 
                      # music experience
                      "instrument",
                      "instrument_training", "musical_level",
                      "musical_training_year", "music_hour",
                      "group_singing_merged", "solo_vocal_merged")

# model formula
model_formula <- as.formula(
  paste("P_prop ~", paste(demographic_vars, collapse = " + "))
)

model_data <- xdata[, c("P_prop", demographic_vars)]   # keep only relevant columns

model_data <- model_data  |> 
  mutate(across(where(is.character), as.factor))  |> 
  mutate(across(where(is.logical), as.factor)) # cforest cannot handle character or logical vector

```

## Tuning the RF

First, map out the parameter combinations you want to test.

```{r}
#| label: define-grid-for-tuning

tuning_grid <- expand.grid(
  mtry      = 1:length(demographic_vars),
  minbucket = c(5, 10, 20),
  maxdepth  = c(0, 5, 10)   # 0 = unlimited depth in party::cforest_control
)

```

Loop through the grid, fit a ranger model for each combination, and extract the Out-of-Bag Mean Squared Error ($OOB\_MSE$) from the forest component to evaluate performance.

```{r}
#| label: run-grid-search

ntree_tuning <- 500

tune_results <- vector("list", nrow(tuning_grid))

set.seed(123)

for (i in seq_len(nrow(tuning_grid))) {
  
  params <- tuning_grid[i, ]
  
  cf_tune <- cforest(
    model_formula,
    data = model_data,
    controls = cforest_unbiased(
      ntree     = ntree_tuning,
      mtry      = params$mtry,
      minbucket = params$minbucket,
      maxdepth  = params$maxdepth
    )
  )
  
  oob_preds_tune <- predict(cf_tune, OOB = TRUE)
  mse_tune <- mean((model_data$P_prop - oob_preds_tune)^2)
  
  tune_results[[i]] <- data.frame(
    mtry      = params$mtry,
    minbucket = params$minbucket,
    maxdepth  = params$maxdepth,
    oob_mse   = mse_tune
  )
  
  cat("Combination", i, "of", nrow(tuning_grid),
      "| mtry =", params$mtry,
      "| minbucket =", params$minbucket,
      "| maxdepth =", params$maxdepth,
      "| OOB MSE =", round(mse_tune, 5), "\n")
}

tuning_df <- bind_rows(tune_results) |> 
  arrange(oob_mse)
print(head(tuning_df, 10))

# Select best combination
best_params <- tuning_df[1, ]
cat("\nBest hyperparameters selected:\n")
print(best_params)

# however, given the small sample in pilot (n = 34), we need to drastically lower minbucket.

```

## Optimal Model

Once the loop finishes, look at grid to find the row with the lowest error score and use those exact parameters to train final model.

```{r}
#| label: identify-optimal-model

set.seed(123)

cf_model <- party::cforest(
  model_formula,
  data = model_data,
  controls = cforest_unbiased(
    ntree     = 1000,   # bump up ntree for the final model for stability
    mtry      = best_params$mtry,
    #minbucket = best_params$minbucket,
    minbucket = 3,
    maxdepth  = best_params$maxdepth
  )
)

oob_preds <- predict(cf_model, OOB = TRUE)

mse_oob <- mean((model_data$P_prop - oob_preds)^2)
r2_oob  <- 1 - (sum((model_data$P_prop - oob_preds)^2) /
                 sum((model_data$P_prop - mean(model_data$P_prop))^2))

cat("Final model - Out-of-Bag MSE:", round(mse_oob, 4), "\n")
cat("Final model - Out-of-Bag R2:", round(r2_oob, 4), "\n")

```

## Predictor importance

```{r}
set.seed(123)

cond_importance <- permimp(
  cf_model,
  conditional = TRUE,
  threshold = 0.95,
  asParty = TRUE,      # use the original party-style algorithm instead
  progressBar = TRUE
)

importance_df <- data.frame(
  predictor  = names(cond_importance$values),
  importance = as.numeric(cond_importance$values)
) %>%
  arrange(desc(importance))

print(importance_df)

# Plot
ggplot(importance_df, 
       aes(x = reorder(predictor, importance), y = importance,
       xend = reorder(predictor, importance))) +
  geom_hline(yintercept = 0, 
            lty = "dashed") +
  geom_segment(
    aes(xend = reorder(predictor, importance), y = 0, yend = importance),
    color = "grey70",
    linewidth = 0.8
  ) +
  geom_point(color = "#dd1c77", size = 2) +

  coord_flip() +  # Flips to horizontal for readable variable names
  labs(
    title = "Variable Importance (Fixed Effects Component)",
    subtitle = "Extracted from MERFranger Object",
    x = "",
    y = "Importance Score (Permutation)"
  ) +
  theme_minimal()

```

## Partial Dependence

```{r}

top_n <- 5
top_predictors <- importance_df$predictor[1:top_n]

pred_fun <- function(object, newdata) {
  predict(object, newdata = newdata, OOB = FALSE)
}

pdp_list <- lapply(top_predictors, function(var) {
  pd <- pdp::partial(
    cf_model,
    pred.var  = var,
    pred.fun  = pred_fun,
    train     = model_data,
    grid.resolution = 10,
    ice = FALSE
  )
  names(pd)[1] <- "x_value"
  pd$predictor_name <- var
  pd
})

combined_tbl <- map_dfr(pdp_list, function(pd) {
  pd |> 
    as_tibble() |> 
    mutate(x_value = as.character(x_value))})  |> 
  group_by(predictor_name, x_value)  |> 
  summarise(yhat = mean(yhat, na.rm = TRUE), .groups = "drop")  |> 
  mutate(x_numeric = suppressWarnings(round(as.numeric(x_value), 2)))

# separate factors from numeric to better plot
combined_tbl_numeric  <- 
  combined_tbl |> 
  filter(predictor_name %in% c("birthyear"))
combined_tbl_factors <- 
  combined_tbl |> 
  filter(predictor_name %in% c("sex", "non_L1_tonal_system",
                               "L1_tonal_system", "L1"))

# plot
combined_tbl_numeric |> 
  ggplot(aes(x =  x_numeric, 
             y = yhat*100)) + 
  geom_point() + 
  geom_line() +
  facet_wrap(~predictor_name,
             scales = "free_x") +
  labs(x = "",
      y = "recovered pitch proportion") +
  theme_minimal()

combined_tbl_factors |> 
  ggplot(aes(x =  x_value, 
             y = yhat*100,
             group = 1)) + 
  geom_point() + 
  geom_line() +
  facet_wrap(~predictor_name,
             scales = "free_x") +
    labs(x = "",
      y = "recovered pitch proportion") +
  theme_minimal()

```
