---
title: "Comparing three different models to model psychometric function"
author: Timo Roettger
date: 2026-06-06
format:
  html:
    embed-resources: true
editor: visual
---


```{r setup}
#| label: setup
#| include: FALSE

if (!require("pacman")) install.packages("pacman")
pacman::p_load(brms, ggplot2, patchwork, plotly, tidybayes, modelr, viridis, tidyverse, dplyr, truncnorm, lme4, faux, scam, 
               bridgesampling, 
               bayestestR)

# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# set the current working directory to the one where this file lives
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

```{r}
#| label: load-data

# load in data 2025/09/11
xdata <- read_csv("../data/derived/data_after_exclusion.csv") |> 
  filter(delta_t != 0)

```

# Modelling

We will attempt to identify properties of the underlying psychometric function that describes the relationship between perceived f0 responses, Δf0, and Δt. To that end, we systematically compare different models alongside their assumptions about this relationship. Hombert (1975) suggests a non-linear increase of perceived pitch as a function of Δt that approaches an asymptote towards higher Δt values, i.e. the benefit of a longer tonal slope decreases. We thus construct and compare three model types:
- a generalized linear mixed effects model that assumes a linear relationship between response and predictors (see Fig. #),
- a generalized linear mixed effects model that assumes a monotonic ordinal relationships (e.g. pitch response increases with Δt and/or Δf0 but does not do so linearly),
- a generalized additive mixed effects model that allows for more complex non-linear relationships.

## Linear model

```{r}
#| label: linear_model

# brms formula (full)
formula_linear = brms::bf(response_f_rel_st ~ delta_t * delta_f_raw
                          + (delta_t * delta_f_raw | participantID),
                          nu = 4)

# priors to specify
get_prior(formula_linear, family = student, data = xdata)

## priors
# the maximum change in f0 per unit of pitch slope should be 1/3 of the f0 perturbation, for delta-f = 1 Hz, this corresponds to 0.04 semitones which we chose as the width of for a weakly informative prior for both beta and the by-participants slopes of pitch slope.
# for delta-t, we don't have theoretical predictions for any effect magnitude. Hombert's data suggest a increase of pitch recovery with slope duration. Based on a linear  data, there is a 0.021 ST increase per ms in his strongest pitch perturbation (delta-f = 70 Hz). So our priors should scale for effects even stronger as the ones reported by Hombert. We choose 0.021 as the width of for a weakly informative prior for both beta and the by-participants slopes of slope duration.

# my priors 
mypriors <- c(
              # intercept allows for a lot of variation for baseline perception, but only allows values that are possible in the design 
              prior(cauchy(0, 3), class = "Intercept", 
                    lb = -7.01955, ub = 4.98045),
              prior(cauchy(0, 0.04), class = "b", coef = "delta_f_raw"),
              prior(cauchy(0, 0.021), class = "b", coef = "delta_t"),
              prior(cauchy(0, 0.04), class = "b", coef = "delta_t:delta_f_raw"),
              prior(cauchy(0, 0.04), class = "sd",
                    group = "participantID",
                    coef = "delta_f_raw"),
              prior(cauchy(0, 0.012), class = "sd",
                    group = "participantID",
                    coef = "delta_t"),
              prior(cauchy(0, 0.04), class = "sd",
                    group = "participantID",
                    coef = "delta_t:delta_f_raw"),
              prior(cauchy(0, 3), class = "sd"),
              prior(lkj(1), class = "cor")
              )

# Run model with
linear_model <- brm(
  formula_linear,
  data = xdata,
  file = "../models/Q1_linear_model",
  family = student,
  seed = 123,
  iter = 4000,
  warmup = 2000,
  save_pars = save_pars(all = TRUE),
  prior = mypriors, 
  control = list(adapt_delta = 0.95,
                 max_treedepth = 12)
  )

# FYI: it took my machine a while ()

# check caterpillar
plot(linear_model) # mixes well

# check samples against data
pp_check(linear_model, ndraws = 100, type = "dens_overlay")

# "issue" with discrete responses. however we do not expect this to bias estimates
# to keep models computationally feasible, we refrain from using discrete version of the student-t regression

print(linear_model, digits = 5)

```

## Monotonic predictors

```{r}
#| label: monotonic_model

# code predictors as ordered factors
xdata$delta_f_raw_o <- factor(xdata$delta_f_raw, 
                              levels = c(-40,-30,-20,-10,10,20,30,40),
                              ordered = TRUE)
xdata$delta_t_o <- factor(xdata$delta_t, 
                          levels = c(20,40,60,80,100,120,140),
                          ordered = TRUE)

# brms formula (full)
formula_mono = brms::bf(response_f_rel_st ~ mo(delta_f_raw_o) * mo(delta_t_o) + (mo(delta_f_raw_o) * mo(delta_t_o) | participantID), nu = 4)

get_prior(formula_mono, family = student, data = xdata)

## priors
# the maximum change in f0 per ordered unit of delta_f_raw should be 1/3 of the f0 perturbation, 
# for delta-f from -10 to 10, this corresponds to 0.75 semitones. We chose this as the width of for the prior for both beta and the by-participants slopes of pitch slope.
# for delta-t, we don't have theoretical predictions for effect magnitude. Hombert's data suggest a increase of pitch recovery with slope duration. Based on his data, there is a 0.021 ST increase per ms in his strongest pitch perturbation (delta-f = 70 Hz). So our priors should scale for effects even stronger as the ones reported by Hombert. In this case one step corresponds to 20 ms, so 0.42 STs.

# my priors 
mypriors <- c(
              # intercept allows for a lot of variation for baseline perception, but only allows values that are possible in the design 
              prior(cauchy(0, 3), class = "Intercept", 
                     lb = -7.01955, ub = 4.98045),
              # prior(cauchy(0, 1), class = "b"),
              # prior(cauchy(0, 1), class = "sd"),
              prior(cauchy(0, 0.75), class = "b", coef = "modelta_f_raw_o"),
              prior(cauchy(0, 0.42), class = "b", coef = "modelta_t_o"),
              prior(cauchy(0, 0.75), class = "b", coef = "modelta_f_raw_o:modelta_t_o"),
              prior(cauchy(0, 0.75), class = "sd",
                    group = "participantID",
                    coef = "modelta_f_raw_o"),
              prior(cauchy(0, 0, 0.42), class = "sd",
                    group = "participantID",
                    coef = "modelta_t_o"),
              prior(cauchy(0, 0.75), class = "sd",
                    group = "participantID",
                    coef = "modelta_f_raw_o:modelta_t_o"),
              prior(cauchy(0, 3), class = "sd"),
              prior(lkj(1), class = "cor")
              )

# Run model with gaussian error term
mono_model <- brm(
  formula_mono,
  data = xdata,
  file = "../models/Q1_mono_model",
  family = student,
  seed = 123,
  iter = 4000,
  warmup = 2000,
  save_pars = save_pars(all = TRUE),
  prior = mypriors, 
  control = list(adapt_delta = 0.99,
                 max_treedepth = 12)
  )

# check caterpillar
plot(mono_model) # mixes well

# check samples against data
pp_check(mono_model, ndraws = 100, type = "dens_overlay") 
# "issue" with discrete response

```

## SCAM

NOT DONE

```{r}
#| label: scam_model

# subset:
my_subset <- factor(unique(xdata$participantID))
chosen_levels <- sample(levels(my_subset), size = 10)
xdata_sub <- xdata[xdata$participantID %in% chosen_levels,]

# brms formula (full)
formula_scam = brms::bf(response_f_rel ~ delta_t * delta_f_raw +
                    s(delta_t, bs = "mpi") +
                    s(delta_f_raw, bs = "mpi") 
                    #+ random
                    )

# get priors
get_prior(formula_scam, xdata, family = "student")

# my priors 
mypriors <- c(
              prior(cauchy(0, 10), class = "b"),
              prior(cauchy(0, 10), class = "Intercept", lb = -45, ub = 45),
              prior(cauchy(0, 10), class = "sds", lb = 0)#,
              #prior(lkj(1), class = "cor")
              )

# Run model with gaussian error term
scam_model <- brm(
  formula_scam,
  data = xdata,
  #file = "../models/scam_model",
  family = gaussian,
  #family = student,
  seed = 123,
  iter = 4000,
  warmup = 2000,
  save_pars = save_pars(all = TRUE),
  prior = mypriors, 
  control = list(adapt_delta = 0.99,
                 max_treedepth = 12)
  )

# check caterpillar
plot(scam_model) # mixes well

# check samples against data
pp_check(scam_model, ndraws = 100, type = "dens_overlay") 
# "issue" with discrete responses as well as "spikyness", which could technically be fixed by student error term?

conditional_effects(scam_model)

```

## Exponential

```{r}
#| label: exponential_model

# brms formula (Marton)
formula_exp = bf(response_f_rel_st ~ a + b*(( (delta_t - x0) / (250 - x0))^inv_logit(c)),
     a ~ 1,
     b ~ delta_f_raw + (1|participant),
     c ~ delta_f_raw + (1|participant),
     x0 ~ 1,
     nu = 4,
     nl = T)

# get priors
get_prior(formula_exp, family = "student", xdata)

# my priors 
mypriors <- c(
              #prior(cauchy(0, 10), nlpar = "a"),
              prior(cauchy(0, 3), nlpar = "b"),
              prior(cauchy(0, 3), nlpar = "c")#,
              #prior(cauchy(0, 30), nlpar = "x0"),
              #prior(cauchy(0, 10), class = "sd", lb = 0)#,
              #prior(lkj(1), class = "cor")
              )

# Run model with student error term
exp_model <- brm(
  formula_exp,
  data = xdata,
  file = "../models/exp_model",
  family = student,
  seed = 123,
  iter = 4000,
  warmup = 2000,
  save_pars = save_pars(all = TRUE),
  prior = mypriors, 
  control = list(adapt_delta = 0.99,
                 max_treedepth = 12)
  )

# check caterpillar
plot(exp_model) # mixes well

# check samples against data
pp_check(exp_model, ndraws = 100, type = "dens_overlay") + xlim(-50,50) 
# "issue" with discrete responses as well as "spikyness", which could technically be fixed by student error term?

```

## GAMs

```{r}
#| label: gam

# brms formula?
formula_gam = brms::bf(response_f_rel_st ~ delta_t *
                         delta_f_raw +
                   #s(delta_t, bs = "bs", k = 6) +
                   #s(delta_f_raw, bs = "bs", k = 6) +
                 t2(delta_t, delta_f_raw),
                 nu = 4)


# get priors
get_prior(formula_gam, xdata, family = "student")

# my priors 
mypriors <- c(
               prior(cauchy(0, 3), class = "Intercept", 
                     lb = -7.01955, ub = 4.98045),
              # prior(cauchy(0, 1), class = "b"),
              # prior(cauchy(0, 1), class = "sd"),
              prior(cauchy(0, 3), class = "b", coef = "delta_f_raw"),
              prior(cauchy(0, 3), class = "b", coef = "delta_t"),
              prior(cauchy(0, 3), class = "b", coef = "delta_t:delta_f_raw"),
              prior(cauchy(0, 3), class = "b", coef = "t2delta_tdelta_f_raw_1"),
              prior(cauchy(0, 3), class = "b", coef = "t2delta_tdelta_f_raw_2"),
              prior(cauchy(0, 3), class = "b", coef = "t2delta_tdelta_f_raw_3"),
              # prior(cauchy(0, 0.75), class = "sd",
              #       group = "participantID",
              #       coef = "modelta_f_raw_o"),
              # prior(cauchy(0, 0, 0.42), class = "sd",
              #       group = "participantID",
              #       coef = "modelta_t_o"),
              # prior(cauchy(0, 0.75), class = "sd",
              #       group = "participantID",
              #       coef = "modelta_f_raw_o:modelta_t_o"),
              prior(cauchy(0, 3), class = "sds")#,
              #prior(lkj(1), class = "cor")
              )

# Run model 
gam_model <- brm(
  formula_gam,
  data = xdata,
  #file = "../models/Q1_gam_model",
  family = student,
  seed = 123,
  iter = 4000,
  warmup = 2000,
  save_pars = save_pars(all = TRUE),
  prior = mypriors, 
  control = list(adapt_delta = 0.999,
                 max_treedepth = 15)
  )

# check caterpillar
plot(gam_model) # does not mix well

# check samples against data
pp_check(gam_model, ndraws = 100, type = "dens_overlay") 
# assuming values way to spread

```

# Model comparison

```{r}

# leave one out model comparison
loo_comp_1 <- loo_compare(list(linear = loo(linear_model), monotonic = loo(mono_model))) # linear vs. monotonic
loo_comp_2 <- loo_compare(list(linear = loo(mono_model), monotonic = loo(exp_model))) # exp vs. monotonic

# Bayes Factor model comparison
## Need to rerun models with much more samples and saving all vars
linear_bridge <- bridge_sampler(linear_model, silent = T)
mono_bridge <- bridge_sampler(mono_model, silent = T)
bf_bridge <- bridgesampling::bf(mono_bridge, linear_bridge)
bf_bridge

```

# Plot models

Here is the descriptive plot:

```{r}
xdata_agg <- xdata |> 
  filter(delta_t != 0) |> 
  group_by(delta_f_raw, delta_t, participantID) |> 
  summarise(response_f_rel_st = mean(response_f_rel_st, na.rm = TRUE)) |> 
  ungroup()

xdata_agg_all <- xdata |> 
  filter(delta_t != 0) |> 
  group_by(delta_f_raw, delta_t) |> 
  summarise(response_f_rel_st = mean(response_f_rel_st, na.rm = TRUE)) |> 
  ungroup() |> 
  mutate(delta_f_raw_ord = as.ordered(delta_f_raw)) |> 
  arrange(delta_t, delta_f_raw_ord)

descriptive_plot <- 
ggplot(data = xdata_agg_all, 
       aes(x = delta_t,
           y = response_f_rel_st,
           color = delta_f_raw_ord,       
           group = delta_f_raw_ord)) +    
  geom_line(lwd = 2.5, color = "black") + # Background outline
  geom_line(lwd = 2) +                    # Colored line
  scale_y_continuous(limits = c(-1.5, 1.5), breaks = seq(-1.5, 1.5, 0.5)) +
  scale_x_continuous(breaks = unique(xdata$delta_t)) +
  labs(x = "\nslope duration in ms",
       y = "average observed pitch response in semitones\n",
       color = "f0 slope in Hz") +
  # Note: You might need to toggle reverse = TRUE/FALSE here 
  # now that the underlying data order is flipped!
  guides(color = guide_legend(reverse = TRUE)) + 
  theme_minimal()
  
descriptive_plot

```

Extract estimates from linear model

```{r}

newdata <- expand_grid(delta_t = unique(xdata$delta_t),
                       delta_f_raw = unique(xdata$delta_f_raw))
                     
# Visualize
#prediction_exp <- exp_model |> 
prediction <- linear_model |> 
  epred_draws(newdata = newdata,
              re_formula = NA) |> 
    #group_by(delta_t_o, delta_f_raw_o) |> 
    group_by(delta_t, delta_f_raw) |> 
    summarise(pred_m = mean(.epred, na.rm = TRUE),
              pred_0025 = quantile(.epred, prob = 0.025),
              pred_0975 = quantile(.epred, prob = 0.975)) |> 
  rename(delta_t = delta_t,
         delta_f_raw = delta_f_raw) |>
  mutate(delta_t = as.numeric(as.character(delta_t)),
         delta_f_raw = as.numeric(as.character(delta_f_raw)))

lm_model_plot <- 
prediction |> 
  ggplot(aes(x = delta_t,
             y = pred_m,
             color = as.ordered(delta_f_raw),
             group = as.ordered(delta_f_raw))) +
  geom_point() +
  geom_line(lwd = 2.5, 
             color = "black") +
  geom_line(lwd = 2) +
    scale_x_continuous(breaks = unique(xdata$delta_t)) +
    scale_y_continuous(limits = c(-1.5, 1.5), breaks = seq(-1.5, 1.5, 0.5)) +
  labs(x = "\nslope duration in ms",
       y = "average observed pitch response in semitones\n",
       color = "f0 slope in Hz") +
  guides(color = guide_legend(reverse = TRUE)) + 
  theme_minimal()

lm_model_plot

```

Extract estimates from mono model

```{r}
#xdata$preds <- posterior_summary(posterior_epred(mono_model_d_f0))[,1]

newdata <- expand_grid(delta_t_o = unique(xdata$delta_t_o),
                       delta_f_raw_o = unique(xdata$delta_f_raw_o))
                       
# Visualize
prediction <- mono_model |> 
  epred_draws(newdata = newdata,
              re_formula = NA) |> 
    group_by(delta_t_o, delta_f_raw_o) |> 
    summarise(pred_m = mean(.epred, na.rm = TRUE),
              pred_0025 = quantile(.epred, prob = 0.025),
              pred_0975 = quantile(.epred, prob = 0.975)) |> 
  rename(delta_t = delta_t_o,
         delta_f_raw = delta_f_raw_o) |>
  mutate(delta_t = as.numeric(as.character(delta_t)),
         delta_f_raw = as.numeric(as.character(delta_f_raw)))

mono_model_plot <- 
prediction |> 
  ggplot(aes(x = delta_t,
             y = pred_m,
             color = as.ordered(delta_f_raw),
             group = as.ordered(delta_f_raw))) +
  geom_point() +
  geom_line(lwd = 2.5, 
             color = "black") +
  geom_line(lwd = 2) +
    scale_x_continuous(breaks = unique(xdata$delta_t)) +
      scale_y_continuous(limits = c(-1.5, 1.5), breaks = seq(-1.5, 1.5, 0.5)) +
  labs(x = "\nslope duration in ms",
       y = "average observed pitch response in semitones\n",
       color = "f0 slope in Hz") +
  guides(color = guide_legend(reverse = TRUE)) + 
  theme_minimal()

mono_model_plot

```

Extract estimates from gam model

```{r}
#xdata$preds <- posterior_summary(posterior_epred(mono_model_d_f0))[,1]

newdata <- expand_grid(
  delta_t = seq(min(xdata$delta_t), max(xdata$delta_t), length.out = 50),
  delta_f_raw = unique(xdata$delta_f_raw))
                       
# Visualize
prediction <- newdata %>%
  add_epred_draws(gam_model) %>%
  # Summarize the posterior distribution for each combination
  median_qi(.epred)

gam_model_plot <- 
prediction |> 
  ggplot(aes(x = delta_t,
             y = .epred,
             color = as.ordered(delta_f_raw),
             group = as.ordered(delta_f_raw))) +
  geom_point() +
  geom_line(lwd = 2.5, 
             color = "black") +
  geom_line(lwd = 2) +
    scale_x_continuous(breaks = unique(xdata$delta_t)) +
      scale_y_continuous(limits = c(-6,6), breaks = seq(-5, 3, 0.5)) +
  labs(x = "\nslope duration in ms",
       y = "average observed pitch response in semitones\n",
       color = "f0 slope in Hz") +
  guides(color = guide_legend(reverse = TRUE)) + 
  theme_minimal()

gam_model_plot

```

```{r}
data_v_model <- (descriptive_plot | lm_model_plot) / (mono_model_plot | gam_model_plot) +
  plot_layout(axes = 'collect_y',
              guides = 'collect')

data_v_model

```
