---
title: "Comparing three different models to model psychometric function"
author: Timo Roettger
date: 2026-06-06
format:
  html:
    embed-resources: true
editor: visual
---


```{r setup}
#| label: setup
#| include: FALSE

if (!require("pacman")) install.packages("pacman")
pacman::p_load(brms, ggplot2, here, patchwork, plotly, tidybayes, modelr, viridis, tidyverse, dplyr, marginaleffects,
               bridgesampling, collapse, bayestestR)

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

```{r}
#| label: load-data

# load in data 2026/May
xdata <- read_csv(here::here("data", "derived", "data_after_exclusion.csv"))

# hombert for priors
hombert <- read_csv(here::here("data", "derived", "hombert.csv"))

```

# Modelling

We will attempt to identify properties of the underlying psychometric function that describes the relationship between perceived f0 responses, Δf0, and Δt. To that end, we systematically compare different models alongside their assumptions about this relationship. Hombert (1975) suggests a non-linear increase of perceived pitch as a function of Δt that approaches an asymptote towards higher Δt values, i.e. the benefit of a longer tonal slope decreases. We thus use as a starting point a linear mixed effects model that constraints the predictors to monotonicity (mo). This assumption is an efficient & computationally feasible way to relax the assumption of linearity, without having to use more complex models (GAMMs) with shape-constrained or unconstrained non-linear smooths.

## Priors
In the following we will justify prior widths for linear and monotonic effects of delta-f0 and delta-t. 

### Linear priors
The maximum change in f0 per unit of pitch slope should be 1/3 of the f0 perturbation (see Rossi 1971, 1978). For delta-f = 1 Hz, this corresponds to 0.33 Hz and ~0.04 STs (with reference value of 150 Hz) which we chose as the width of for a weakly informative priors for both beta and the by-participants slopes of delta-f. If delta-f is transformed to STs, a change in 1 ST is 0.33 STs, respectively.

```{r}
#| label: linear-prior-delta-f

semitones_to_hz <- function(st, ref = 150) {
  return(ref * 2^(st / 12))
}

# reference level
pitch_reference = 150

# ceiling of recovering 1 Hz
pitch_r_ceiling = 1 * 1/3

# st difference
prior_linear_delta_f <- 12 * log2((pitch_reference - pitch_r_ceiling) / pitch_reference)
# ~0.038

prior_linear_delta_f_st = 1/3

```

For delta-t, we don't have theoretical predictions for the effect magnitude. Hombert's data suggest a increase of pitch recovery with slope duration. Based on a linear interpolation of the most extreme pitch perturbation (70 Hz), there was a 0.017 STs increase per ms. So our priors should scale for effects reported by Hombert. We choose 0.021 as the width of for a weakly informative prior for both beta and the by-participants slopes of slope duration.

```{r}
#| label: linear-prior-delta-f

# linear interpolation for hombert
hombert_sub <- hombert |> 
  filter(target == 70)

lm_hombert <- lm(val_st ~ slope_dur, hombert_sub)
prior_linear_delta_t <- lm_hombert[1][[1]][[2]]

```

### Monotonic priors ('mo()')

mo() coefficients describe the total change from min level to max level for an ordered factor divided by the number of levels -1. That is, the coefficient for mo() should show the average effect for one change in ordered level of delta f0. But when an additional predictor is present, it also takes as the reference level the lowest level of the respective other predictors. We have to keep that in mind when defining priors.

For delta_f, one step corresponds to on average of 10 Hz. Making the assumption that the effect is constant at all delta-ts, we thus multiply the linear prior for Hz with this number, arriving at ~0.39 STs. Due to both the observation in Hombert and our own pilots, at the reference level of 20 ms slope duration, this effect is likely diminished. To keep our priors reasonably tight, we half this value, arriving at ~0.19 STs. If delta-f is transformed to STs, the largest change in STs form step to step is from -40 Hz (-5.37) to -30 Hz (3.86) ~ 1.5 STs. On third would be 0.5 STs and halving that ends up at 0.25 STs.

```{r}
#| label: mo-prior-delta-f

# translate 1 Hz prior into prior for one step mo()
mo_projection_delta_f <- prior_linear_delta_f * 10

# half because of delta-t = 20ms reference
prior_mo_delta_f <- mo_projection_delta_f / 2

prior_mo_delta_f_st = 0.25

```

For delta-t, one step corresponds to 20 ms. Making the assumption that the effect is constant at all delta-fs, we thus multiply the linear prior for 1 ms with 20, arriving at ~0.34 STs. 

```{r}
#| label: mo-prior-delta-t

# half because of delta-t = 20ms reference
prior_mo_delta_t <- prior_linear_delta_t * 20

```

## Baseline mo() model

```{r}
#| label: monotonic_model

# code predictors as ordered factors
# delta_f
xdata$delta_f_raw_o <- factor(xdata$delta_f_raw, 
                              levels = c(-40,-30,-20,-10,0,10,20,30,40),
                              ordered = TRUE)
# for st-transformed delta_f
xdata$delta_f_raw_st_o <- factor(xdata$delta_f_raw_st, 
                              levels = levels(as.ordered(xdata$delta_f_raw_st)),
                              ordered = TRUE)

# delta_t
xdata <- xdata |> 
  mutate(delta_t_o = factor(xdata$delta_t, 
                            levels = c(20,40,60,80,100,120,140),
                            ordered = TRUE))

# brms formula (full)
formula_mono = brms::bf(response_f_rel_st ~ mo(delta_f_raw_st_o) * mo(delta_t_o) + 
                          (mo(delta_f_raw_st_o) * mo(delta_t_o) | participantID)
                        # + (mo(delta_f_raw_o) * mo(delta_t_o) | language) # for later stage
                        , 
                        nu = 4)

get_prior(formula_mono, family = student, data = xdata)

prior_mo_delta_f  # 0.19
prior_mo_delta_f_st # 0.25
prior_mo_delta_t # 0.34

# my priors 
mypriors <- c(
              # intercept allows for a lot of variation for baseline perception, but only allows values that are possible within the experimental design 
              prior(cauchy(0, 2), class = "Intercept", 
                     lb = -7.02, ub = 4.98),
              prior(cauchy(0, 0.25), class = "b", coef = "modelta_f_raw_st_o"),
              prior(cauchy(0, 0.34), class = "b", coef = "modelta_t_o"),
              prior(cauchy(0, 0.25), class = "b", coef = "modelta_f_raw_st_o:modelta_t_o"),
              prior(cauchy(0, 0.25), class = "sd",
                    group = "participantID",
                    coef = "modelta_f_raw_st_o"),
              prior(cauchy(0, 0.34), class = "sd",
                    group = "participantID",
                    coef = "modelta_t_o"),
              prior(cauchy(0, 0.25), class = "sd",
                    group = "participantID",
                    coef = "modelta_f_raw_st_o:modelta_t_o"),
              # prior(cauchy(0, 0.19), class = "sd",
              #       group = "language",
              #       coef = "modelta_f_raw_o"),
              # prior(cauchy(0, 0.34), class = "sd",
              #       group = "language",
              #       coef = "modelta_t_o"),
              # prior(cauchy(0, 0.19), class = "sd",
              #       group = "language",
              #       coef = "modelta_f_raw_o:modelta_t_o"),
              prior(cauchy(0, 2), class = "sigma"),
              prior(lkj(1), class = "cor")
              )

# Run robust regression
mono_model <- brm(
  formula_mono,
  data = xdata,
  file = "../models/Q1_mono_model",
  family = student,
  seed = 123,
  iter = 4000,
  warmup = 2000,
  save_pars = save_pars(all = TRUE),
  prior = mypriors, 
  control = list(adapt_delta = 0.9,
                 max_treedepth = 12)
  )

# check caterpillar
plot(mono_model) # mixes well

# check samples against data
pp_check(mono_model, ndraws = 100, type = "dens_overlay") 
# discrete responses but distribution fine

```

### Marginalizing over effect

Extract contrasts between lowest and highest values for both predictors marginalized over the respective other predictor levels:

```{r}
#| label: marginalize

# marginalizing effects over each other 

## marginal contrast (ndraws = 800 due to memory constraints)
slope_draws_df0 <- avg_slopes(mono_model, 
                          variables = "delta_f_raw_st_o",
                          by = TRUE,
                          newdata = xdata,
                          ndraws = 800)

# We take the diff between extremes (40 vs -40) divided by steps - 1 (= 8) 
marginalized_df0 <- 
posterior_draws(slope_draws_df0) |> 
  filter(contrast == "4.0924430140208 - -5.36950772365466") |> 
  mutate(draw_step = draw / 8) 

marginalized_df0_summary <- 
  marginalized_df0 |> 
  summarise(mean = mean(draw_step),
            lower = quantile(draw_step, 0.05),
            upper = quantile(draw_step, 1),
            count_above_zero = sum(draw_step > 0),
            count_below_zero = sum(draw_step < 0),
            evidence_ratio = count_above_zero / count_below_zero,
            all_samples = n(),
            bf = ifelse(evidence_ratio == 0, 
                        paste("<", 1 / (all_samples + 1)),
                              ifelse(is.infinite(evidence_ratio),
                                     paste(">", all_samples),
                                     evidence_ratio
                        ))) 

# delta_t
slope_draws_dt <- avg_slopes(mono_model, 
                          variables = "delta_t_o",
                          by = TRUE,
                          newdata = xdata,
                          ndraws = 800)

# We take the diff between extremes (20 vs 140) divided by steps - 1 (= 5) 
marginalized_dt <- 
posterior_draws(slope_draws_dt) |> 
  filter(contrast == "140 - 20") |> 
  # flip because of direction
  mutate(draw_step = -draw / 5) 

marginalized_dt_summary <- 
  marginalized_dt |> 
  summarise(mean = mean(draw_step),
            lower = quantile(draw_step, 0.05),
            upper = quantile(draw_step, 1),
            count_above_zero = sum(draw_step > 0),
            count_below_zero = sum(draw_step < 0),
            evidence_ratio = count_above_zero / count_below_zero,
            all_samples = n(),
            bf = ifelse(evidence_ratio == 0, 
                        paste("<", 1 / (all_samples + 1)),
                              ifelse(is.infinite(evidence_ratio),
                                     paste(">", all_samples),
                                     evidence_ratio
                        ))) 

```

### Directional BF

```{r}
#| label: directionl-bf

# visualize them through density plots
marginalized_df0 |> 
  ggplot(aes(x = draw_step)) + 
  stat_halfeye() +
  scale_x_continuous(limits = c(-0.01, 0.1)) +
  geom_vline(xintercept = 0,
             lty = "dashed") +
  annotate("text",
           x = Inf, y = Inf, 
           hjust = 1.2, vjust = 1.5, size = 5,
           label = paste("bf:", marginalized_df0_summary$bf)) +
  theme_minimal()

marginalized_dt |> 
  ggplot(aes(x = draw_step)) + 
  stat_halfeye() +
  scale_x_continuous(limits = c(-0.025, 0.025)) +
  geom_vline(xintercept = 0,
             lty = "dashed") +
  annotate("text", 
           x = Inf, y = Inf, 
           label = paste("bf:", round(marginalized_dt_summary$bf,2)),
           hjust = 1.2, vjust = 1.5, size = 5) +
  theme_minimal()

```

## Testing linearity assumption

```{r}
#| label: linear_models

# only linear delta-t effect
formula_linear_dt = brms::bf(response_f_rel_st ~ delta_t * mo(delta_f_raw_st_o)
                          + (delta_t * mo(delta_f_raw_st_o) | participantID),
                          nu = 4)

# only linear delta-f0 effect
formula_linear_df0 = brms::bf(response_f_rel_st ~ mo(delta_t_o) * delta_f_raw_st
                          + (mo(delta_t_o) * delta_f_raw_st | participantID),
                          nu = 4)

# both linear effects
formula_linear_int = brms::bf(response_f_rel_st ~ delta_t * delta_f_raw_st
                          + (delta_t * delta_f_raw_st | participantID),
                          nu = 4)


# prior_linear_delta_f # 0.039
# prior_linear_delta_t # 0.017
# prior_mo_delta_f_st # 0.25
# prior_mo_delta_t # 0.34

# my priors for linear models
mypriors_linear <- c(
              prior(cauchy(0, 2), class = "Intercept", 
                    lb = -7.01955, ub = 4.98045),
              prior(cauchy(0, 2), class = "sigma"),
              prior(lkj(1), class = "cor")
              )

# specifically for linear delta-t effect
mypriors_linear_delta_t <- c(mypriors_linear,
                             prior(cauchy(0, 0.22), class = "b", coef = "modelta_f_raw_st_o"),
                             prior(cauchy(0, 0.017), class = "b", coef = "delta_t"),
                             prior(cauchy(0, 0.22), class = "b", coef = "modelta_f_raw_st_o:delta_t"),
                             prior(cauchy(0, 0.22), class = "sd",
                                   group = "participantID",
                                   coef = "modelta_f_raw_st_o"),
                             prior(cauchy(0, 0.017), class = "sd",
                                   group = "participantID",
                                   coef = "delta_t"),
                             prior(cauchy(0, 0.22), class = "sd",
                                   group = "participantID",
                                   coef = "modelta_f_raw_st_o:delta_t"))

# specifically for linear delta-f0 effect
mypriors_linear_delta_f0 <- c(mypriors_linear,
                             prior(cauchy(0, 0.038), class = "b", coef = "delta_f_raw_st"),
                             prior(cauchy(0, 0.34), class = "b", coef = "modelta_t_o"),
                             prior(cauchy(0, 0.34), class = "b", coef = "modelta_t_o:delta_f_raw_st"),
                             prior(cauchy(0, 0.038), class = "sd",
                                   group = "participantID",
                                   coef = "delta_f_raw_st"),
                             prior(cauchy(0, 0.34), class = "sd",
                                   group = "participantID",
                                   coef = "modelta_t_o"),
                             prior(cauchy(0, 0.34), class = "sd",
                                   group = "participantID",
                                   coef = "modelta_t_o:delta_f_raw_st"))

# specifically for both linear effects
mypriors_linear_int <- c(mypriors_linear,
                             prior(cauchy(0, 0.038), class = "b", coef = "delta_f_raw_st"),
                             prior(cauchy(0, 0.017), class = "b", coef = "delta_t"),
                             prior(cauchy(0, 0.017), class = "b", coef = "delta_t:delta_f_raw_st"),
                             prior(cauchy(0, 0.038), class = "sd",
                                   group = "participantID",
                                   coef = "delta_f_raw_st"),
                             prior(cauchy(0, 0.038), class = "sd",
                                   group = "participantID",
                                   coef = "delta_t"),
                             prior(cauchy(0, 0.038), class = "sd",
                                   group = "participantID",
                                   coef = "delta_t:delta_f_raw_st"))


# Run three linear models
linear_model_delta_t <- brm(
  formula_linear_dt,
  data = xdata,
  file = "../models/Q1_linear_model_linear_delta_t",
  family = student,
  seed = 123,
  iter = 4000,
  warmup = 2000,
  save_pars = save_pars(all = TRUE),
  prior = mypriors_linear_delta_t, 
  control = list(adapt_delta = 0.90,
                 max_treedepth = 12)
  )

linear_model_delta_f <- brm(
  formula_linear_df0,
  data = xdata,
  file = "../models/Q1_linear_model_linear_delta_f",
  family = student,
  seed = 123,
  iter = 4000,
  warmup = 2000,
  save_pars = save_pars(all = TRUE),
  prior = mypriors_linear_delta_f0, 
  control = list(adapt_delta = 0.90,
                 max_treedepth = 12)
  )

linear_model_inter <- brm(
  formula_linear_int,
  data = xdata,
  file = "../models/Q1_linear_model_linear_inter",
  family = student,
  seed = 123,
  iter = 4000,
  warmup = 2000,
  save_pars = save_pars(all = TRUE),
  prior = mypriors_linear_int, 
  control = list(adapt_delta = 0.90,
                 max_treedepth = 12)
  )


# check samples against data
pp_check(linear_model_inter, ndraws = 100, type = "dens_overlay")

```

### Model comparisons

```{r}
#| label: looic

# leave one out model comparison
loo_comp_1 <- loo_compare(list(linear1 = loo(linear_model_delta_f),
                               linear2 = loo(linear_model_delta_t), 
                               linear3 = loo(linear_model_inter), 
                               monotonic = loo(mono_model))) 
# linear2 wins in pilot

# we accept a model if elpd_diff / se_diff > 3

```

# Plot models 
## delta-t * delta-f0

Here is the descriptive plot:

```{r}
#| label: descriptive-1

xdata_agg <- xdata |> 
  group_by(delta_f_raw, delta_t, participantID) |> 
  summarise(response_f_rel_st = mean(response_f_rel_st, na.rm = TRUE)) |> 
  ungroup()

xdata_agg_all <- xdata |> 
  group_by(delta_f_raw, delta_t) |> 
  summarise(response_f_rel_st = mean(response_f_rel_st, na.rm = TRUE)) |> 
  ungroup() |> 
  mutate(delta_f_raw_ord = as.ordered(delta_f_raw)) |> 
  arrange(delta_t, delta_f_raw_ord)

descriptive_plot <- 
ggplot(data = xdata_agg_all, 
       aes(x = delta_t,
           y = response_f_rel_st,
           color = delta_f_raw_ord,       
           group = delta_f_raw_ord)) +    
  geom_line(lwd = 2.5, color = "black") + # Background outline
  geom_line(lwd = 2) +                    # Colored line
  scale_y_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.25)) +
  scale_x_continuous(breaks = unique(xdata$delta_t)) +
  labs(x = "\nslope duration in ms",
       y = "average observed pitch response in semitones\n",
       color = "f0 slope in Hz") +
  # Note: You might need to toggle reverse = TRUE/FALSE here 
  # now that the underlying data order is flipped!
  guides(color = guide_legend(reverse = TRUE)) + 
  theme_minimal()
  
descriptive_plot

```

Extract estimates from best fitting model

```{r}
#| label: model-pred-1

newdata <- expand_grid(delta_t = unique(xdata$delta_t),
                       delta_f_raw_st_o = unique(xdata$delta_f_raw_st_o))
                     
# Visualize
#prediction_exp <- exp_model |> 
prediction <- linear_model_delta_t |> 
  epred_draws(newdata = newdata,
              re_formula = NA) |> 
    #group_by(delta_t_o, delta_f_raw_o) |> 
    group_by(delta_t, delta_f_raw_st_o) |> 
    summarise(pred_m = mean(.epred, na.rm = TRUE),
              pred_0025 = quantile(.epred, prob = 0.025),
              pred_0975 = quantile(.epred, prob = 0.975)) |> 
  rename(delta_t = delta_t,
         delta_f_raw = delta_f_raw_st_o) |>
  mutate(delta_t = as.numeric(as.character(delta_t)),
         delta_f_raw = as.numeric(as.character(delta_f_raw)))

lm_model_plot <- 
prediction |> 
  ggplot(aes(x = delta_t,
             y = pred_m,
             color = as.ordered(delta_f_raw),
             group = as.ordered(delta_f_raw))) +
  geom_point() +
  geom_line(lwd = 2.5, 
             color = "black") +
  geom_line(lwd = 2) +
    scale_x_continuous(breaks = unique(xdata$delta_t)) +
    scale_y_continuous(limits = c(-1, 1), breaks = seq(-1, 1, 0.25)) +
  scale_color_discrete(labels = c(seq(-40, 40, by = 10)),
                       palette = "viridis"
                      ) +
  labs(x = "\nslope duration in ms",
       y = "average observed pitch response in semitones\n",
       color = "f0 slope in Hz") +
  guides(color = guide_legend(reverse = TRUE)) + 
  theme_minimal() +
  theme(legend.position = "none")

lm_model_plot

```

Put together:

```{r}
#| label: poatchwork-1

data_v_model <- (descriptive_plot | lm_model_plot)  +
  plot_layout(axes = 'collect_y',
              guides = 'collect')

data_v_model

```

## Prec%

```{r}
#| label: descriptive-2

xdata_agg_all_prop <- xdata |> 
  filter(delta_f_raw_st != 0) |> 
  group_by(delta_f_raw_st, delta_t) |> 
  summarise(response_f_rel_st = mean(response_f_rel_st, na.rm = TRUE)) |> 
  mutate(response_prop = response_f_rel_st / delta_f_raw_st) |> 
  ungroup() |> 
  mutate(delta_f_raw_ord = as.ordered(delta_f_raw_st)) |> 
  arrange(delta_t, delta_f_raw_ord)

descriptive_prop_plot <- 
ggplot(data = xdata_agg_all_prop, 
       aes(x = delta_t,
           y = response_prop,
           color = delta_f_raw_ord,       
           group = delta_f_raw_ord)) +    
  geom_line(lwd = 2.5, color = "black") + # Background outline
  geom_line(lwd = 2) +                    # Colored line
  scale_y_continuous(limits = c(-0.1, 0.35), 
                     breaks = c(-0.1, 0, 0.1, 0.2, 0.3)) +
  scale_x_continuous(breaks = unique(xdata$delta_t)) +
  scale_color_discrete(labels = c(-40,-30,-20,-10,10,20,30,40),
                       palette = "viridis") +
  labs(x = "\nslope duration in ms",
       y = "average proportion of recovered pitch in %\n",
       color = "f0 slope in Hz") +
  # Note: You might need to toggle reverse = TRUE/FALSE here 
  # now that the underlying data order is flipped!
  guides(color = guide_legend(reverse = TRUE)) + 
  theme_minimal()
  
descriptive_prop_plot

```

Extract estimates from best fitting model

```{r}
#| label: model-pred-2


# translate predictions in proportions
prediction <-  linear_model_delta_t |>
  epred_draws(newdata = newdata,
              re_formula = NA) |> 
  filter(delta_f_raw_st_o != 0) |> 
  mutate(delta_f_raw_st_o = as.numeric(as.character(delta_f_raw_st_o))) |> 
  mutate(prop = .epred / delta_f_raw_st_o) |> 
  group_by(delta_t, delta_f_raw_st_o) |> 
  summarise(pred_m = mean(prop, na.rm = TRUE),
            pred_0025 = quantile(prop, prob = 0.025),
            pred_0975 = quantile(prop, prob = 0.975)) |> 
  rename(delta_t = delta_t,
         delta_f_raw = delta_f_raw_st_o) |>
  mutate(delta_t = as.numeric(as.character(delta_t)),
         delta_f_raw = as.numeric(as.character(delta_f_raw)))

lm_model_prop_plot <- 
prediction |> 
  ggplot(aes(x = delta_t,
             y = pred_m,
             color = as.ordered(delta_f_raw),
             group = as.ordered(delta_f_raw))) +
  geom_point() +
  geom_line(lwd = 2.5, 
             color = "black") +
  geom_line(lwd = 2) +
  scale_y_continuous(limits = c(-0.1, 0.35), 
                     breaks = c(-0.1, 0, 0.1, 0.2, 0.3)) +
  scale_x_continuous(breaks = unique(xdata$delta_t)) +
  scale_color_discrete(labels = c(-40,-30,-20,-10,10,20,30,40),
                       palette = "viridis") +
  labs(x = "\nslope duration in ms",
       y = "average proportion of recovered pitch in %\n",
       color = "f0 slope in Hz") +
  guides(color = guide_legend(reverse = TRUE)) + 
  theme_minimal() +
  theme(legend.position = "none")

lm_model_prop_plot

```

```{r}
#| label: patchwork-2

data_v_model_prop <- (descriptive_prop_plot | lm_model_prop_plot)  +
  plot_layout(axes = 'collect_y',
              guides = 'collect')

data_v_model_prop

```

# Extract estimates for each participant (without shrinkage)

```{r}
#| label: blups

newdat <- expand.grid(
  delta_f_raw_st_o = unique(xdata$delta_f_raw_st_o),
  delta_t = unique(xdata$delta_t),
  participantID = unique(xdata$participantID)
)

preds <- fitted(linear_model_delta_t, newdata = newdat, re_formula = NULL, summary = TRUE)

# calculate P-pred%
predictions <- cbind(newdat, preds) |>
  select(1:4) |> 
  filter(delta_f_raw_st_o != 0) |> 
  as_tibble() |> 
  mutate(P_prop = Estimate / as.numeric(as.character(delta_f_raw_st_o))) 

# derive single estimate
predictions_agg <- predictions |> 
  group_by(participantID) |> 
  summarise(P_prop = mean(P_prop))
  
# plot
predictions_agg |> 
  ggplot(aes(x = P_prop)) +
  geom_point(y = 0,
             size = 3,
             pch = 21,
             color = "white",
             fill = "#dd1c77") +
  geom_density(fill = "#dd1c77",
               alpha = 0.5) +
  scale_x_continuous(limits = c(-0.1,0.4),
                     breaks = seq(-0.1,0.4,0.1),
                     labels = paste0(seq(-0.1,0.4,0.1)*100, "%")) +
  geom_vline(lty = c("dashed", "dotted"),
             xintercept = c(0,1/3)) +
  labs(title = "Individual differences for recovered f0 across all conditions",
       subtitle = "",
       y = "",
       x = "percentage of recovered f0") + 
  theme_minimal() +
  theme(legend.position = "none",
        strip.text = element_blank())

```

```{r}
#| label: store_part_estimates

write_csv(predictions_agg, "../data/derived/participant_estimated_forRF.csv")

```

