---
title: "Random Forests"
author: Timo Roettger
date: 2026-06-06
format: html
editor: visual
---

```{r}
#| include: FALSE

knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE)

```

```{r}
#| label: setup
#| include: FALSE

# install dev version of SAEforest
# devtools::install_github("krennpa/SAEforest")

# read in libraries
if (!require("pacman")) install.packages("pacman")
pacman::p_load(viridis, tidyverse, rstudioapi, knitr, kableExtra, broom, patchwork, plotly, ranger, SAEforest)

# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# set the current working directory to the one where this file lives
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

```

```{r}
#| label: data

# load in data 2026/06/01
xdata <- read_csv("../data/data_after_exclusion.csv") 

# generate average recovered_f0
xdata_agg <- xdata |> 
  filter(delta_f != 0) |> 
  group_by(participantID) |> 
  summarise(recovered_f0_mean = mean(recovered_f0, na.rm = FALSE))

# left_join with other infos
xdata <- xdata |> 
  select(1:16) |> 
  #filter(delta_f != 0) |> 
  left_join(xdata_agg, by = "participantID") |> 
  distinct()





```


```{r}
#| label: prepare-for-merf

# define DV and IVs
xdata$participantID <- as.factor(xdata$participantID)
y.cols <- xdata$recovered_f0
## need to define the relevant predictors 
x.cols <- xdata[,c(3:6,8:15)]

```

## Tuning the MERF
First, map out the parameter combinations you want to test.

```{r}
#| label: define-grid-for-tuning

# Define hyperparameters to test
tuning_grid <- expand.grid(
  mtry = seq(2, 10, 2),                     # Adjust based on your total number of features: Currently 12+ 
  min_node_size = c(1, 3, 5, 10, 20, 50, 100),       # Lower = deeper trees (more complex)
  num_trees = 500,                       # Fixed at 500 for speed
  stringsAsFactors = FALSE
)

# Create a placeholder column to store performance results
tuning_grid$OOB_MSE <- NA_real_

```

Loop through your grid, fit a MERFranger model for each combination, and extract the Out-of-Bag Mean Squared Error ($OOB\_MSE$) from the forest component to evaluate performance.

```{r}
#| label: run-grid-search

# Loop through each row of the grid
for (i in 1:nrow(tuning_grid)) {
  
  cat(sprintf("Tuning iteration %d/%d: mtry = %d, min.node.size = %d\n", 
              i, nrow(tuning_grid), tuning_grid$mtry[i], tuning_grid$min_node_size[i]))
  
  # Fit the model with current parameters
  temp_merf <- MERFranger(
    Y = y.cols,
    X = x.cols,
    random = "(1|participantID)",
    data = xdata,
    num.trees = tuning_grid$num_trees[i],
    mtry = tuning_grid$mtry[i],
    min.node.size = tuning_grid$min_node_size[i],
    verbose = FALSE # Turn off inner logs to keep console clean
  )
  
  # Extract the prediction error (MSE) from the ranger object
  tuning_grid$OOB_MSE[i] <- temp_merf$Forest$prediction.error
}

```

Once the loop finishes, look at your grid to find the row with the lowest error score and use those exact parameters to train your final model.

```{r}
#| label: identify-optimal-model

# Find the best parameters
best_params <- tuning_grid |> 
  arrange(OOB_MSE) |> 
  slice(1)

print(best_params)
# mtry = 2
# min_node_size = 10

# Train your final production model using the best parameters found
final_merf_model <- MERFranger(
  Y = y.cols,
  X = x.cols,
  random = "(1|participantID)",
  data = xdata,
  num.trees = best_params$num_trees,
  mtry = best_params$mtry,
  min.node.size = best_params$min_node_size,
  importance = "permutation" # Turn on importance for your final model
)

```

Now extract variable importance:

```{r}
#| label: var-importance
#| fig-cap: "Variable Importance (Fixed Effects Component)"

var_imp <- importance(final_merf_model$Forest)

print(var_imp)

imp_df <- tibble(
  Variable = names(var_imp),
  Importance = var_imp
) |> 
  arrange(desc(Importance))

# Plot
ggplot(imp_df, 
       aes(x = reorder(Variable, Importance), y = Importance,
       xend = reorder(Variable, Importance))) +
  geom_segment(
    aes(xend = reorder(Variable, Importance), y = 0, yend = Importance),
    color = "grey70",
    linewidth = 0.8
  ) +
  geom_point(color = "#dd1c77", size = 2) +
  coord_flip() +  # Flips to horizontal for readable variable names
  labs(
    title = "Variable Importance (Fixed Effects Component)",
    subtitle = "Extracted from MERFranger Object",
    x = "",
    y = "Importance Score (Permutation)"
  ) +
  theme_minimal()

```

