---
title: "ManyTones analysis"
author: "Timo Roettger"
date: "2025-07-21"
output: html_document
---

```{r setup, include=FALSE}

if (!require("pacman")) install.packages("pacman")
pacman::p_load(brms, ggplot2, patchwork, tidybayes, modelr, viridis, tidyverse, dplyr, truncnorm, lme4, faux, scam)

# Set the current working directory to the one where this file is
current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path)
setwd(current_working_dir)

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

## Data

```{r}

# load in data 2025/09/11
xdata <- read_csv("../data/merged_250911.csv")

# derive some relevant measurements
xdata <- xdata |> 
  # include trials when slider was not moved?
  #filter(stimuli_direction != 0) |>
  filter(delta_t != 0) |> 
  # make delta_f positive / negative
  mutate(delta_f_raw = ifelse(stimuli_direction == "p", delta_f, delta_f * -1),
         # make response_f relative to baseline 150hz      
         response_f_rel = response_f - 150,
         # make response_f absolute without sign differences
         response_f_rel_abs = abs(response_f_rel),
         # relative to target
         response_f_prop = response_f_rel / delta_f,
         # calculate actual f0 of presented tonal onset
         presented_f = 150 + delta_f_raw,
         # make is.correct numeric
         correct = as.numeric(is_correct)
         )

```

# Continuous dependent variable

## Verbal predictions

(1) The duration of the f0 slope affects perceived pitch such that longer duration lead to perceived pitch that is closer to the presented f0 onset.

(2) The difference between the f0 onset and the level tone affects perceived pitch such that larger differences lead to larger perceived pitch differences.

Visually assess relationship between slope duration (delta_t), f0 difference between onset and level tone (delta_f_raw) and response.

```{r}

xdata_agg <- xdata |> 
  group_by(delta_f_raw, delta_t, participant) |> 
  summarise(response_f_rel = mean(response_f_rel, na.rm = TRUE))
  
xdata_agg_all <- xdata |> 
  group_by(delta_f_raw, delta_t) |> 
  summarise(response_f_rel = mean(response_f_rel, na.rm = TRUE))
  
descriptive_plot_01 <- 
ggplot(data = xdata_agg, 
       aes(x = delta_t,
           y = response_f_rel,
           color = delta_f_raw)) +
  geom_line(data = xdata_agg, 
            aes(group = interaction(participant, delta_f_raw)),
          alpha = 0.2) +
  geom_point(data = xdata_agg_all) +
  geom_line(data = xdata_agg_all,
             aes(group = delta_f_raw),
             lwd = 2.5,
            color = "black") +
  geom_line(data = xdata_agg_all,
             aes(group = delta_f_raw),
             lwd = 2) +
  scale_y_continuous(limits = c(-40,30),
                     breaks = seq(-40,30,10)) +
  scale_x_continuous(breaks = c(unique(xdata$delta_t))) +
  scale_color_viridis(option="viridis") +
  theme_minimal()
  
descriptive_plot_01

```

Slight non-linear trajectory, but predictions are descriptively confirmed.

## Statistical model

Hypotheses are tested, and relevant parameters are estimated through the following model (following Marton's parametric suggestion)

`response_f_rel \~ a + b \* (( (delta_t - x0) / (x_max - x0) )\^inv_logit(c))`

a = assesses intercept, i.e. perceived pitch differences at `delta_t = 0` and `delta_f_raw = 0`

b = assesses linear relationship of `delta_t` and `response_f_rel` and how this relationship is modulated by `delta_f_raw`

c = delta_f_raw \# assess non-linear relationship of `delta_t` and `response_f_rel` and how this relationship is modulated by `delta_f_raw`

x0 \# ?

## Statistical hypotheses

(1) `delta_t` linearly affects `response_f_rel`, i.e. there is evidence of `b_intercept > 0`, over `b_intercept = 0`.

(2) `delta_f_raw` linearly affects the relationship in (1), i.e. there is evidence of `b_delta_f_raw > 0`, over `b_delta_f_raw = 0`.

## Further exploratory parameter estimation

(3) ...

```{r}

# add ceiling
xdata <- xdata |> 
  mutate(
  #x_min=min(delta_t),
  x_max = max(delta_t)
)

# TO DO: Explore other functions
## gumbel function?
## Naka-rushton?

# Model of the following basic structure:
# y ~ a + b*x^

# TO DO: Justify priors
# TO DO: random slopes for all these parameters? Computationally feasible?
# TO DO: inferential procedure? Savage Dickey-based BF against point zero?

# run model
parametric_model <- brm(
  bf(response_f_rel ~ a + b * (( (delta_t - x0) / (x_max - x0) )^inv_logit(c)),
     a ~ 1,
     b ~ delta_f_raw,
     c ~ delta_f_raw,
     x0 ~ 1,
     nl = T),
  data = xdata,
  file = "../models/model_nl_TR",
  prior = prior(normal(0,100), nlpar = "a")+
          prior(normal(0,100), nlpar = "b")+
          prior(normal(0,100), nlpar = "x0")+
          prior(normal(0,1), nlpar = "c", lb = 0, ub = 1),
  control = list(adapt_delta=0.99)
  )

# comment TR: takes 90 seconds on my machine, so computationally expensive
parametric_model
# model predicts mostly linear relationship

# posterior predictive check
pp_check(parametric_model)
# 1. not truly continuous measure, problem?
# 2. regardless, distribution looks spikier than gaussian

# check chains
plot(parametric_model, pars = "^b_") # okay

xdata$preds <- posterior_summary(posterior_epred(parametric_model))[,1]

# Visualize
model_plot_01 <- 
ggplot(data = xdata, 
       aes(x = delta_t,
           y = preds,
           color = delta_f_raw,
           group = delta_f_raw)) +
  geom_line(lwd = 2.5, color = "black") +
  geom_line(lwd = 2) +
  scale_y_continuous(limits = c(-40,30),
                     breaks = seq(-40,30,10)) +
  scale_x_continuous(breaks = c(unique(xdata$delta_t))) +
  scale_color_viridis(option="viridis") +
  theme_minimal()
  
model_plot_01

data_v_model <- descriptive_plot_01 + model_plot_01
data_v_model

# pretty decent fit, but asymmetries between positive and negative delta_f are not captured

```

# Accuracy of pitch direction

## Verbal predictions

(1) Directional perception can be reliably perceived for some combination of slope duration and pitch difference between the f0 onset and the level tone.

Question: What to do with those trials that are neither higher or lower but chosen to be level? Options: (1) categorical GLM with three response options (2) remove "m"s, and run logistic regression or (3) consider "m"s incorrect, and run logistic regression (currently implemented)

```{r}

xdata_agg_all <- xdata |> 
  group_by(delta_f, delta_t, stimuli_direction) |> 
  summarise(correct = mean(correct, na.rm = TRUE))
  
# Plot empirical accuracy
#bare_plot <- 
  ggplot(data = xdata_agg_all, 
       aes(x = delta_f,
           y = correct,
           color = stimuli_direction)) +
  geom_hline(yintercept = 0.5,
             lty = "dashed") +
  geom_path(lwd = 2) +
  #scale_color_viridis(option="viridis") +
  scale_y_continuous(limits = c(0,1),
                     breaks = seq(0,1,.25)) +
  scale_x_continuous(breaks = seq(-40,40,20)) +
  facet_wrap(delta_t ~ ., ncol = 3) +
  theme_minimal()
  
# asymmetric effects of positive vs. negative delta_f:
  # negative: stable >75% accuracy regardless of delta_f and delta_t
  # positive: increases with delta_f and delta_t

```

## Statistical model

Hypotheses are tested, and relevant parameters are estimated through the following model:

`correct ~ delta_t * delta_f * direction`

## Statistical hypotheses

(1) 

(2) 

## Further exploratory parameter estimation

(3) ...

```{r}

# contrast code
xdata <- xdata |> 
  mutate(delta_f_c = delta_f - mean(delta_f),
         delta_t_c = delta_t - mean(delta_t),
         stimuli_direction_c = ifelse(stimuli_direction == "p", 0.5, -0.5))

# formula
formula = bf(correct ~ delta_f_c * delta_t_c * stimuli_direction_c)

# run model
logistic_model <- brm(
  formula,
  data = xdata,
  file = "../models/model_glm_TR",
  family = "bernoulli",
  prior = prior(normal(0,1), class = "b"), 
  control = list(adapt_delta=0.99)
  )

# comment TR: takes 50 seconds on my machine, so computationally expensive and 500 convergence
logistic_model
# model predicts mostly linear relationship

# posterior predictive check
pp_check(logistic_model)

# check chains
plot(logistic_model, variable = "^b_") # okay

xdata$preds_glm <- posterior_summary(posterior_epred(logistic_model))[,1]

# plot
ggplot(data = xdata, 
       aes(x = delta_f,
           y = preds_glm,
           color = stimuli_direction)) +
  geom_hline(yintercept = 0.5,
             lty = "dashed") +
  # geom_ribbon(aes(ymin = lci, 
  #                 ymax = hci),
  #             #color = "grey",
  #             fill = "lightgrey") +
  geom_path(lwd = 2) +
  #scale_color_viridis(option="viridis") +
  scale_y_continuous(limits = c(0,1),
                     breaks = seq(0,1,.25)) +
  scale_x_continuous(breaks = seq(0,40,20)) +
  facet_wrap(delta_t ~ ., ncol = 3) +
  theme_minimal()
 
```

Now we can relate these functions to a reliable perceptual difference (above 50%).

Goal would be to find an inferential criterion that assesses how much evidence we have for above JND performance for each parameter combination.

```{r}

JND = c(-3,3)

```
